---
title: "Mapping stats and TET3 genotyping by reads"
subtitle: "_Tomato TET3 BS-seq screening test_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

```{r include=FALSE}
# Libraries
library(tidyverse)
library(cowplot); theme_set(theme_cowplot())
library(scales)
```

Read the bismark summary report to use it as a sample table

```{r fig.height=10, fig.width=15}
bismark_report <- read_tsv("../qa/3_mapped/bismark_summary_report.txt", col_types = cols()) %>% 
  transmute(sample = str_remove(File, "_bismark_bt2.bam"),
            raw_reads = `Total Reads`,
            mapped_reads = `Aligned Reads`,
            mapping_rate = mapped_reads / raw_reads,
            unmapped_reads = `Unaligned Reads`,
            ambiguous_reads = `Ambiguously Aligned Reads`,
            dedup_reads = `Unique Reads (remaining)`,
            dup_rate = `Duplicate Reads (removed)` / dedup_reads,
            cytosines = `Total Cs`,
            meth_CG = `Methylated CpGs`,
            unmeth_CG = `Unmethylated CpGs`,
            mCG = 100 * meth_CG / meth_CG + unmeth_CG,
            meth_CHG = `Methylated chgs`,
            unmeth_CHG = `Unmethylated chgs`,
            mCHG = 100 * meth_CHG / meth_CHG + unmeth_CHG,
            meth_CHH = `Methylated CHHs`,
            unmeth_CHH = `Unmethylated CHHs`,
            mCHH = 100 * meth_CHH / meth_CHH + unmeth_CHH) %>% 
  # filter out the controls and samples with almost no reads
  filter( ! sample %in% c("A1_2", "B2_13", "neg","PBS", "Undetermined")) %>% 
  # set order for the sampes
  mutate(sample = factor(sample, levels = c(
    "WT_1", "WT_2", "WT_3", "WT_4", "WT_6",
    "A_1", "A_4", "A_12", "A_14", "A_15", "A_17", "A_21", "A_22",  
    "A1_4", "A1_6", "A1_7", "A1_8", 
    "A11_3", "A11_5", "A11_8", "A11_10", "A11_11", 
    "B_2", "B_3", "B_4", "B_5", "B_6", "B_8", "B_10", "B_11", 
    "B1_2", "B1_3", "B1_6", "B1_7", "B1_8", "B1_9",
    "B2_1", "B2_5", "B2_9", "B2_10", "B2_14", "B2_16", "B2_18", "B2_20"
  ))) %>% 
  arrange(sample)
```

```{r}
(sample_table <- bismark_report %>% 
  select(sample, raw_reads, dedup_reads, cytosines) %>%
  mutate(condition = str_remove(sample, "_.*") %>% factor(levels = c("WT", "A", "A1", "A11", "B", "B1", "B2")),
         line = str_remove_all(condition, "[:digit:]*") %>% factor(levels = c("WT", "A", "B")),
         replica = str_remove(sample, ".*_") %>% as_factor(),
         description = case_when(condition == "WT" ~ "Control",
                                 condition == "A" ~ "A T1",
                                 condition == "A1" ~ "A T2",
                                 condition == "A11" ~ "A T3",
                                 condition == "B" ~ "B T1",
                                 condition == "B1" ~ "B T2-",
                                 condition == "B2" ~ "B T2+") %>% 
           fct_inorder()) %>%
  arrange(condition, replica))
```

```{r aesthetic-variables}
# Aesthetics to plot the PCA and other plots
subline_colors <- c(A   = RColorBrewer::brewer.pal(8, "Reds")[4], 
                    A1  = RColorBrewer::brewer.pal(8, "Reds")[6],
                    A11 = RColorBrewer::brewer.pal(8, "Reds")[8], 
                    B   = RColorBrewer::brewer.pal(8, "Blues")[4], 
                    B1  = RColorBrewer::brewer.pal(8, "Blues")[6], 
                    B2  = RColorBrewer::brewer.pal(8, "Blues")[8], 
                    WT  = RColorBrewer::brewer.pal(8, "Greys")[8])

line_shapes <- c(WT = 21, A = 23, B = 24)
genotype_shape <- c("TET3" = 24, "NT" = 21) 
```

Previously I had used the threshold of 15 Million Cs covered to discard some samples and make the PCAs of 100 kbp bins less noisy. After having another go at the PCAs I think it is better to use **14 Million Cs**. This way I keep more 2 more samples for the analysis and they don't seem to be outliers.

```{r fig.width=8, fig.height=3}
bismark_report %>% 
  left_join(sample_table, by = c("sample", "raw_reads", "dedup_reads", "cytosines")) %>% 
  ggplot(aes(sample, cytosines, fill = condition)) +
  facet_grid(cols = vars(description), scales = "free_x", space = "free_x") +
  geom_col(alpha = 0.8, width = 0.5, color = "black") + 
  geom_point(aes(color = cytosines > 14e6), shape = 18, size = 4) +
  geom_hline(yintercept = 14e6, linetype = "dashed") +
  scale_fill_manual(values = subline_colors) +
  scale_color_manual(values = (c("red2", "grey20"))) +
  scale_y_continuous(expand = expansion(c(0, 0.1)), labels = label_number(scale_cut = cut_short_scale())) +
  labs(title = "Cytosines covered by BS-seq", 
       subtitle = "We discarded samples with less than 14 Million Cytosines covered",
       y = "Covered Cs") +
  theme(legend.position = "none", 
          strip.background = element_rect(colour = NA, fill = "lightpink"),
          panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5), axis.title.x = element_blank())

ggsave("../qa/covered-cytosines.pdf", width=8, height=3)
```

## Determine TET3 genotype from mapped reads {.tabset}

I will use the number of mapped reads as an indication if a plant is carying the TET3 transgene or not. This information is summarised in the samtools idxstats output

```{r}
idxstats <- read_tsv("../qa/3_mapped/03_mapped_samtools_multiqc_data/multiqc_samtools_idxstats.txt", col_types = cols()) %>% 
  select(-`*`) %>% 
  dplyr::rename(sample = Sample) %>% 
  pivot_longer(SL4.0ch00:last_col(), names_to = "chr", values_to = "reads") %>% 
  separate(reads, c(NA, "reads", "chr_length", NA), convert = TRUE) %>% 
  group_by(sample) %>% 
  mutate(reads = replace_na(reads, 0),
         total_reads = sum(reads, na.rm = TRUE),
         reads_per_million = 1e6 * reads / total_reads,
         reads_per_million_per_megabase = 1e6 * reads_per_million / chr_length) %>%
  # filter controls and samples with almost no reads
  filter( ! sample %in% c("A1_2", "B2_13", "neg","PBS", "Undetermined")) %>% 
  mutate(sample = factor(sample, levels = c(
    "WT_1", "WT_2", "WT_3", "WT_4", "WT_6",
    "A_1", "A_4", "A_12", "A_14", "A_15", "A_17", "A_21", "A_22",  
    "A1_4", "A1_6", "A1_7", "A1_8", 
    "A11_3", "A11_5", "A11_8", "A11_10", "A11_11", 
    "B_2", "B_3", "B_4", "B_5", "B_6", "B_8", "B_10", "B_11", 
    "B1_2", "B1_3", "B1_6", "B1_7", "B1_8", "B1_9",
    "B2_1", "B2_5", "B2_9", "B2_10", "B2_14", "B2_16", "B2_18", "B2_20"
  ))) %>%
  left_join(sample_table, by = "sample")

# filter samples with less than 14 Million mapped reads
idxstats_filtered <- idxstats %>% 
  filter(cytosines > 14e6) # c("B2_18", "B2_20", "B_6", "B_11", "WT_6", "B_5", "A_1", "A_12", "B1_2", "B2_14")) #, "B1_7", "B2_16"))
```

### Chr9 anomaly

```{r fig.height=4, fig.width=6}
ggplot(idxstats_filtered, aes(chr, reads_per_million_per_megabase, color = condition, group = sample)) +
  geom_hline(yintercept = 1350, linetype = "dashed") +
  geom_jitter() +
  scale_color_manual(values = subline_colors) +
  labs(title = "normalised reads mapped to each chromosome", 
       subtitle = "normalised both for library size and chromosome size",
       y = "reads per million per Mbp",
       caption = "Note the drop in all samples for chr 9") +
  theme(strip.background = element_rect(colour = NA, fill = "lightpink"),
          panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
          axis.text.x = element_text(angle = -90, hjust = 0), axis.title.x = element_blank())
ggsave("../qa/chr9-anomaly.pdf", width = 6, height = 4)
```

### All plants {}

```{r fig.width=9, fig.height=3}
# TDNA_TET3_v2.1
ggplot(idxstats %>% filter(chr == "TDNA_TET3_v2.1"), aes(sample, reads, fill = condition)) +
  facet_grid(cols = vars(description), scales = "free_x", space = "free_x") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_col(alpha = 0.8, width = 0.5, color = "black") +
  geom_point(aes(shape = reads > 10), color = "black", size = 3) +
  scale_shape_manual(values = c(21, 24 )) +
  scale_fill_manual(values = subline_colors) +
  labs(title = "Genotyping TET3 plants", 
       subtitle = "using reads mapped uniquely to TET3 T-DNA",
       y = "reads per million",
       caption = "samples with more than 10 reads per million are considered TET3+") +
  theme(legend.position = "none", 
          strip.background = element_rect(colour = NA, fill = "lightpink"),
          panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
          axis.text.x = element_text(angle = -90, hjust = 0), axis.title.x = element_blank())

ggsave("../figures/genotyping-TET3-all.pdf", width = 9, height = 3)
```

### samples with > 14M Cs {.active}

```{r fig.width=8, fig.height=3}
# TDNA_TET3_v2.1
ggplot(idxstats_filtered %>% filter(chr == "TDNA_TET3_v2.1"), aes(sample, reads, fill = condition)) +
  facet_grid(cols = vars(description), scales = "free_x", space = "free_x") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_col(alpha = 0.8, width = 0.5, color = "black") +
  geom_point(aes(shape = reads > 10), color = "black", size = 3) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = subline_colors) +
  labs(title = "Genotyping TET3 plants", 
       subtitle = "using reads mapped uniquely to TET3 T-DNA",
       y = "reads per million",
       caption = "samples with more than 10 reads per million are considered TET3+") +
  theme(legend.position = "none", 
          strip.background = element_rect(colour = NA, fill = "lightpink"),
          panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
          axis.text.x = element_text(angle = -90, hjust = 0), axis.title.x = element_blank())

ggsave("../figures/genotyping-TET3.pdf", width = 8, height = 3)
```

## Write sample table for future use

Define transgenic plants and write to table

```{r}
(genotype_table <- idxstats %>% 
  arrange(sample) %>% 
  filter(chr == "TDNA_TET3_v2.1") %>% 
  select(sample, TET3_reads = reads, total_reads_idxstats = total_reads, TET3_reads_per_million = reads_per_million) %>% 
  mutate(TET3_reads_per_million = round(TET3_reads_per_million, 2),
         genotype = case_when(TET3_reads_per_million > 10 ~ "TET3",
                              TET3_reads_per_million < 10 ~ "NT")))

sample_table_genotype <- sample_table %>% 
  full_join(genotype_table, by = "sample")

write_csv(sample_table_genotype, "../data/tet3-bs-screen-sample-table.csv")
```

