---
title: "Genome wide plots of mC in 100 kbp bins, with moving average"
subtitle: "_Tomato TET3 BS-seq selected plants_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

------------------------------------------------------------------------

Analysis of DNA methylation in 100 kbp bins created with `CX_report2summary_v0.1.pl` from Bismark CX_reports.

```{r libraries-and-functions, include=FALSE}
# Libraries
library(tidyverse)
library(cowplot); theme_set(theme_cowplot())
library(patchwork)
library(slider)
library(ggforce)
library(scales)

facet_chr <- function(chr = chr, rows = NULL, scales = "free_x", space = "free_x", breaks = 40e6) {
  list(facet_grid(rows = vars({{ rows }}), cols = vars({{ chr }}), scales = scales, space = space),
       scale_x_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(breaks), 
                          labels = label_number(scale_cut = cut_short_scale())),
       xlab("Genomic coordinates (bp)"))
}

heterochromatin_H3K27ac = read_tsv("/path/to/genome/SL4.0_heterochromatin_H3K27ac_chXX.bed", col_names = c("chr","start","end"), col_types = cols()) %>% 
  mutate(chr = str_c("SL4.0", chr))

SL4_chrs <- str_c("SL4.0ch", str_pad(c(0:8, 10:12), 2, pad = "0"))

geom_heterochromatin <- function(hchrom = heterochromatin_H3K27ac, chr_col = chr, chrs = SL4_chrs) {
  hchrom <- dplyr::rename(hchrom, {{ chr_col }} := chr)
  list(geom_rect(data = hchrom %>% filter(chr %in% chrs),
                 aes(x = NULL, y = NULL, xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
                 color = NA, fill = "black", alpha = 0.1))
}
```

# Data

```{r load-bins}
bins_long_table <- read_tsv("../data/4_methylation/tomato_tet3_screening.mC_bins.100kbp.tsv", col_types = cols())  %>% 
  # remove some ANSI carriage movement character
  mutate(chr = str_remove_all(chr, "\\\033\\[K")) %>% 
  # filter controls and samples with almost no reads
  filter( ! sample %in% c("A1_2", "B2_13", "neg","PBS", "Undetermined")) %>% 
  # order the samples as a factor
  mutate(sample = factor(sample, levels = c(
    "WT_1", "WT_2", "WT_3", "WT_4", "WT_6",
    "A_1", "A_4", "A_12", "A_14", "A_15", "A_17", "A_21", "A_22",  
    "A1_4", "A1_6", "A1_7", "A1_8", 
    "A11_3", "A11_5", "A11_8", "A11_10", "A11_11", 
    "B_2", "B_3", "B_4", "B_5", "B_6", "B_8", "B_10", "B_11", 
    "B1_2", "B1_3", "B1_6", "B1_7", "B1_8", "B1_9",
    "B2_1", "B2_5", "B2_9", "B2_10", "B2_14", "B2_16", "B2_18", "B2_20"
  ))) %>% 
  # remove ch09 with the anomaly and organelle genomes
  filter(!(chr %in% c("SL4.0ch09", "Mt", "Pt")))
```

```{r pivot-bins}
bins_wide_table <- bins_long_table %>% 
  pivot_wider(chr:context, names_from = sample, values_from = mR, names_sort = TRUE)
```

I filter out the samples that contain less than 14 Million Cytosines covered (WT_6, A_1, A_12, B_6, B_11, B_5, B1_2, B2_14, B2_18 & B2_20) as outliers as explained in the bins PCA report. 

```{r}
bins_wide_table_filtered <- bins_wide_table %>%
  select(-c(B2_18, B2_20, B_6, B_11, WT_6, B_5, A_1, A_12, B1_2, B2_14))
```

```{r read-sample-table}
sample_table <- read_csv("../data/tet3-bs-screen-sample-table.csv", col_types = cols()) %>% 
  mutate(sample = fct_inorder(sample),
         line = fct_inorder(line),
         condition = fct_inorder(condition),
         description = fct_inorder(description))
```

```{r aesthetic-variables}
# Aesthetics to plot the PCA and other plots
subline_colors <- c(A   = RColorBrewer::brewer.pal(8, "Reds")[4], 
                    A1  = RColorBrewer::brewer.pal(8, "Reds")[6],
                    A11 = RColorBrewer::brewer.pal(8, "Reds")[8], 
                    B   = RColorBrewer::brewer.pal(8, "Blues")[4], 
                    B1  = RColorBrewer::brewer.pal(8, "Blues")[6], 
                    B2  = RColorBrewer::brewer.pal(8, "Blues")[8], 
                    WT  = RColorBrewer::brewer.pal(8, "Greys")[8])

line_shapes <- c(WT = 21, A = 23, B = 24)
genotype_shapes <- c("TET3" = 24, "NT" = 21) 
```

## Scatter matrices {.tabset .tabset-pills}

```{r}
scatter_matrix <- function(bins, condition, context = "C", title = str_c("Scatter plot of %m", context, " from ", condition)) {
  ctx <- context
  bins %>%
    select(chr:context, starts_with(condition)) %>%
    filter(str_detect(context, ctx)) %>%
    add_column(slope = 1) %>% 
    add_column(intercept = 0) %>% 
    na.omit() %>% 
    ggplot(aes(x = .panel_x, y = .panel_y)) +
    geom_point(shape = 16, size = 0.3, alpha = 0.5) +
    geom_point(shape = 16, size = 0, alpha = 0) +
    geom_density_2d_filled(contour_var = "ndensity") +
    geom_abline(aes(intercept = intercept, slope = slope), color = "red", alpha = 0.5) +
    geom_autohistogram(color = 1, fill = 3) +
    facet_matrix(vars(-c(chr:context, slope, intercept)), layer.upper = c(1,4), layer.lower = c(2,3,4), layer.diag = 5) +
    theme(axis.line = element_blank(), panel.border = element_rect(colour = "black", size = 1, fill = NA), 
        strip.background = element_blank(), plot.title = element_text(hjust = 0.5), legend.position = "none") +
    ggtitle(title)
}
```

### WT CG

```{r fig.height=6, fig.width=6}
scatter_matrix(bins_wide_table_filtered, "WT",  context = "CG",  title = "WT %mCG")
```

### WT CHG

```{r fig.height=6, fig.width=6}
scatter_matrix(bins_wide_table_filtered, "WT",  context = "CHG",  title = "WT %mCHG")
```

### WT CHH

```{r fig.height=6, fig.width=6}
scatter_matrix(bins_wide_table_filtered, "WT",  context = "CHH",  title = "WT %mCHH")
```

### A T1 CG

```{r fig.height=7, fig.width=7}
scatter_matrix(bins_wide_table_filtered, "A_",  context = "CG",  title = "A T1 %mCG")
```

### A T2 CG

```{r fig.height=7, fig.width=7}
scatter_matrix(bins_wide_table_filtered, "A1_",  context = "CG",  title = "A T2 %mCG")
```

### A T3 CG

```{r fig.height=7, fig.width=7}
scatter_matrix(bins_wide_table_filtered, "A11_",  context = "CG",  title = "A T3 %mCG")
```

### B T1 CG

```{r fig.height=7, fig.width=7}
scatter_matrix(bins_wide_table_filtered, "B_",  context = "CG",  title = "B T1 %mCG")
```

### B T2- CG

```{r fig.height=7, fig.width=7}
scatter_matrix(bins_wide_table_filtered, "B1_",  context = "CG",  title = "B T2- %mCG")
```

### B T2+ CG

```{r fig.height=7, fig.width=7}
scatter_matrix(bins_wide_table_filtered, "B2_",  context = "CG",  title = "B T2+ %mCG")
```

# Moving average 

Function to generate sliding mean and median of methylation and methylation difference from a reference sample (renamed). It takes a wide table and produces a long table.

```{r}
pivot_longer_diff_to_reference <- function(bins_wide, reference = WT_mean, values = mR) {
  bins_wide %>%
    dplyr::rename(ref = {{ reference }}) %>%
    pivot_longer(-c(chr:context,ref), names_to = "sample", values_to = as.character(ensym(values)),
                 names_transform = list(sample = forcats::fct_inorder)) %>%
    mutate(diff = {{ values }} - ref)
}

compute_sliding_stats <- function(bins_long_with_diff, values = mR, flanking = 100) {
  bins_long_with_diff %>% 
    arrange(sample, chr, context, start) %>%
    group_by(sample, chr, context) %>%
    mutate(mean        = slide_dbl({{ values }}, ~mean(.x,   na.rm = T), .before = flanking, .after = flanking),
           median      = slide_dbl({{ values }}, ~median(.x, na.rm = T), .before = flanking, .after = flanking),
           mean_diff   = slide_dbl(diff,         ~mean(.x,   na.rm = T), .before = flanking, .after = flanking),
           median_diff = slide_dbl(diff,         ~median(.x, na.rm = T), .before = flanking, .after = flanking)) %>% 
  ungroup()
}
```

Compute the sliding binned data for 21 bins (10 bins (1 Mb) to each side)
```{r}
bins_long_table_sliding <- bins_wide_table_filtered %>%
  mutate(WT_mean = (WT_1 + WT_2 + WT_3 + WT_4) / 4) %>%
  pivot_longer_diff_to_reference() %>% 
  compute_sliding_stats(flanking = 20) %>%
  left_join(sample_table, by = "sample")
```

Compute the sliding binned data for 101 bins (50 bins (5 Mb) to each side)
```{r}
bins_long_table_sliding_10Mbp <- bins_wide_table_filtered %>%
  mutate(WT_mean = (WT_1 + WT_2 + WT_3 + WT_4) / 4) %>%
  pivot_longer_diff_to_reference() %>% 
  compute_sliding_stats(flanking = 100) %>%
  left_join(sample_table, by = "sample")
```

## some dispersion plot

```{r fig.width=10, fig.height=10}
bins_long_table_sliding %>% 
  left_join(hchrom_H3K27ac_100kbp_bins, by = c("chr", "start", "end")) %>% 
  filter(sample %in% c("WT_1", "A_4", "A_17"),
         !is.na(chromatin)) %>% 
  ggplot(aes(ref, diff, color = chromatin)) +
  facet_grid(rows = vars(sample), cols = vars(context), scales = "free") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(shape = 16, size = 1, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  theme(axis.line=element_blank(), strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour="black", size=1, fill=NA), legend.position = "bottom",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
```

## Trellis plots

```{r plot function meth}
plot_trellis <- function(data, y, chr_col = chr, scales = "free", chrs = SL4_chrs, breaks = 40e6) {
  data %>%
    filter({{ chr_col }} %in% chrs) %>%
    ggplot(aes(x = start, y = {{ y }}, color = condition)) +
    facet_chr(chr = {{ chr_col }}, rows = context, scales = scales, space = "free_x", breaks = breaks) +
    geom_heterochromatin(chrs = chrs) +
    scale_y_continuous(expand = expansion(mult = 0.05), limits = c(NA, NA)) +
    geom_line(aes(group = sample)) +
    scale_color_manual(values = subline_colors)  +
    theme_cowplot() +
    theme(panel.spacing.x=unit(0.8, "lines"),
          strip.background = element_rect(colour = NA, fill = "lightpink"),
          panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

}
```

```{r fig.height=6, fig.width=12}
bins_long_table_sliding %>% 
  #filter(sample %in% c("WT_1", "A_4", "A_17")) %>% 
  plot_trellis(y = mean)+#, chrs = "SL4.0ch02", breaks = 10e6) +
  #geom_hline(yintercept = 0, linetype = "dashed", size = 3) +
  theme(legend.position = "bottom")
```

```{r fig.height=8, fig.width=16}
bins_long_table_sliding %>% 
  #filter(sample %in% c("WT_1", "A_4", "A_17")) %>% 
  plot_trellis(y = mean_diff)+#, chrs = "SL4.0ch02", breaks = 10e6) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 2, color = "lightgreen") +
  theme(legend.position = "bottom")

ggsave2("../figures/trellis-plots/trellis10Mbp-avg-all-ch-diff.pdf", width = 16, height = 8)
```


```{r fig.height=6, fig.width=12}
bins_long_table_sliding %>% 
  plot_trellis(y = mean, chrs = "SL4.0ch01", breaks = 10e6) +
  #geom_hline(yintercept = 0, linetype = "dashed", size = 3) +
  theme(legend.position = "bottom")
```

```{r fig.height=6, fig.width=12}
bins_long_table_sliding %>% 
  plot_trellis(y = mean, chrs = "SL4.0ch02", breaks = 10e6) +
  #geom_hline(yintercept = 0, linetype = "dashed", size = 3) +
  theme(legend.position = "bottom")
```

```{r fig.height=6, fig.width=6}
bins_long_table_sliding_10Mbp %>% 
  plot_trellis(y = mean, chrs = "SL4.0ch02", breaks = 10e6) +
  #geom_hline(yintercept = 0, linetype = "dashed", size = 2, color = "lightgreen") +
  theme(legend.position = "bottom")

ggsave2("../figures/trellis-plots/trellis-ch02-10Mbp-avg.pdf", width = 6, height = 6)
```

```{r fig.height=6, fig.width=6}
bins_long_table_sliding_10Mbp %>% 
  plot_trellis(y = mean_diff, chrs = "SL4.0ch02", breaks = 10e6) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 2, color = "lightgreen") +
  theme(legend.position = "bottom")

ggsave2("../figures/trellis-plots/trellis-ch02-10Mbp-avg-diff.pdf", width = 6, height = 6)
```


## Graphs {.tabset}

Moving average of the methylation rate (or difference in methylation rate) over 200 bins (2 Mb)

Big increase in CHH methylation, almost twice the level of WT. Bigger increase in heterochromatin than in chromosome arms, where there is less CHH methylation in the first place.

Small loss in CG, but very small proportionally. Bigger decrease in chromosome arms than in heterochromatin

No real differences in CHG methylation

### Methylation {.tabset}
#### All chromosomes
```{r fig.height=6, fig.width=12}
plot_summary(meth_summary_201, y = mean) 
```






## Heatmap plots

```{r plot-function-meth-tile}
plot_tile <- function(data, fill, chr_col = chr, scales = "free", chrs = SL4_chrs, breaks = 40e6) {
  data %>%
    filter({{ chr_col }} %in% chrs) %>%
    ggplot(aes(y = start, x = sample, fill = {{ fill }})) +
    #facet_chr(chr = {{ chr_col }}, rows = context, scales = scales, space = "free_x", breaks = breaks) +
    facet_grid(rows = vars(context), cols = vars(description), scales = scales, space = "free") +
    scale_y_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(breaks), 
                       labels = label_number(scale_cut = cut_short_scale())) +
    ylab("Genomic coordinates (bp)") +
    #geom_heterochromatin(chrs = chrs) +
    #scale_y_continuous(expand = expansion(mult = 0.05), limits = c(NA, NA)) +
    geom_tile() +
    scale_fill_distiller(palette = "RdBu", limits = c(-5, 5), oob = scales::squish)  +
    #scale_fill_distiller(palette = "YlGnBu", direction = 1)  +
    #scale_y_discrete(limits = rev) +
    theme_cowplot() +
    theme(panel.spacing.x=unit(0.3, "lines"),
          strip.background = element_rect(colour = NA, fill = "lightpink"),
          panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

}
```


```{r fig.height=8, fig.width=8}
bins_long_table_sliding %>% 
  plot_tile(fill = mean_diff, chrs = "SL4.0ch02", breaks = 10e6)
```


```{r}
genes <- read_bed("/path/to/genome/ITAG4.1/ITAG4.1_gene_positions.bed")

k27ac <- read_bigwig("../../tracks/v4/marks/wt_K27ac.bw")

bins_100kb <- tileGenome(seqlengths = seqinfo(k27ac), tilewidth = 1e5, cut.last.tile.in.chrom = TRUE)

n_genes_threshold = 8
flanking = 10
scale_genes <- scale_fill_viridis_c("Genes per 100 kb", option = "viridis", limits = c(0, 15), oob=scales::squish)

hacky_genes <- binned_genes %>% 
  filter(seqnames == "ch02") %>% 
  mutate(context = "CG") %>% 
  bind_rows(binned_genes %>% filter(seqnames == "ch02") %>% mutate(context = "CHG")) %>% 
  bind_rows(binned_genes %>% filter(seqnames == "ch02") %>% mutate(context = "CHH"))

gene_tile_continous <- ggplot(hacky_genes, aes(y = start, x = "", fill = mean)) +
  facet_grid(rows = vars(context)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(10e6), 
                     labels = label_number(scale_cut = cut_short_scale())) +
  ylab("Genomic coordinates (bp)") +
  geom_raster() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_genes +
  theme(panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank(), axis.text = element_blank(),
        strip.text = element_blank(), plot.margin = margin(t = 0, b = 0))

gene_tile_discrete <- ggplot(hacky_genes, aes(y = start, x = "", fill = chromatin)) +
  facet_grid(rows = vars(context)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(10e6), 
                     labels = label_number(scale_cut = cut_short_scale())) +
  ylab("Genomic coordinates (bp)") +
  geom_raster() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = c("royalblue1", "grey80")) +
  theme(strip.text = element_blank(), axis.ticks.x = element_blank(), 
        panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank(),
        axis.title.x = element_blank(), plot.margin = margin(t = 0))

gene_tile_discrete + 
  gene_tile_continous +
  plot_layout(widths = c(1, 1), guides = "collect") &
  theme(legend.position='bottom')
```

```{r fig.height=8, fig.width=8}
gene_tile_discrete + 
  gene_tile_continous +
  bins_long_table_sliding %>% 
  plot_tile(fill = mean_diff, chrs = "SL4.0ch02", breaks = 10e6) +
  geom_hline(yintercept = 33e6, linetype = "dashed") +
  theme(axis.ticks.y = element_blank(), axis.title = element_blank(), axis.text.y = element_blank(),
        plot.margin = margin(t = 0, b = 0)) +
  plot_layout(widths = c(1, 1, 18), guides = "collect") +
  plot_annotation(
    title = "%mC difference to control across Chromosome 2",
    subtitle = "Dotted line represent Euchromatin/Heterochromatin boundary")

ggsave2("../figures/trellis-plots/heatmap-ch02-diff-with-chromatin.pdf", width = 8, height = 8)
```


```{r fig.height=8, fig.width=8}
gene_tile_discrete + 
  gene_tile_continous +
  bins_long_table_sliding %>% 
  plot_tile(fill = mean, chrs = "SL4.0ch02", breaks = 10e6) +
  geom_hline(yintercept = 33e6, linetype = "dashed") +
  scale_fill_distiller(palette = "YlGnBu", direction = 1)  +
  theme(axis.ticks.y = element_blank(), axis.title = element_blank(), axis.text.y = element_blank(),
        plot.margin = margin(t = 0, b = 0)) +
  plot_layout(widths = c(1, 1, 18), guides = "collect") +
  plot_annotation(
    title = "%mC level across Chromosome 2",
    subtitle = "Dotted line represent Euchromatin/Heterochromatin boundary")

ggsave2("../figures/trellis-plots/heatmap-ch02-with-chromatin.pdf", width = 8, height = 8)
```

