---
title: "PCAs of mC in 100 kbp bins"
subtitle: "_Tomato TET3 BS-seq screening test_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

------------------------------------------------------------------------

Analysis of DNA methylation in 100 kbp bins created with `CX_report2mC_bins.v1.0.pl` from Bismark CX_reports.

```{r libraries-and-functions, include=FALSE}
# Libraries
library(tidyverse)
library(magrittr)
library(cowplot); theme_set(theme_cowplot())
library(ggforce)
library(patchwork)
```

# Load data

I filter out the samples that contain less than 14 Million Cytosines covered (WT_6, A_1, A_12, B_6, B_11, B_5, B1_2, B2_14, B2_18 & B2_20) as outliers as explained in the bins PCA report. 

```{r}
# Read matrix
bins_wide_table_filtered <- read_tsv("../data/4_methylation/tomato_tet3_screening.mC_bins.100kbp.wide.tsv", show_col_types = FALSE) %>%
  select(-c(B2_18, B2_20, B_6, B_11, WT_6, B_5, A_1, A_12, B1_2, B2_14)) %>% 
  select(chr:context, WT_1, WT_2, WT_3, WT_4, B_2, B_3, B_4, B_8, B_10, B1_3, B1_6, B1_7, B1_8, B1_9, B2_1, B2_5, B2_9, B2_10, B2_16)
```

# Compute inheritance

Calculate the difference in methylation between B T1 and B T2- for CG. Threshold defined as 5% loss.

```{r}
ctx <- "CG"         # "CG", "CHG", "CHH"
delta_thr <- -5     # percentage points

wt_cols  <- c("WT_1","WT_2","WT_3","WT_4")
bt1_cols <- c("B_2","B_3","B_4","B_8","B_10")
bt2m_cols <- c("B1_3","B1_6","B1_7","B1_8","B1_9")  # B_T2- non-transgenic sibs is actualy B1 in this nomenclature 

bins_ctx <- bins_wide_table_filtered %>%
  filter(context == ctx) %>%
  mutate(
    mWT  = rowMeans(across(all_of(wt_cols)),  na.rm = TRUE),
    mT1  = rowMeans(across(all_of(bt1_cols)), na.rm = TRUE),
    mT2m = rowMeans(across(all_of(bt2m_cols)), na.rm = TRUE),
    dT1  = mT1  - mWT,   # Δ(T1 - WT), in % units
    dT2m = mT2m - mWT    # Δ(T2- - WT)
  )

# Define T1-affected bins and "propagated" bins (directional stability)
t1_affected <- bins_ctx %>%
  filter(dT1 <= delta_thr)

prop_stats <- t1_affected %>%
  summarise(
    n_bins = n(),
    propagated_n = sum(dT2m <= delta_thr, na.rm = TRUE),
    propagated_frac = propagated_n / n_bins,
    spearman_rho = cor(dT1, dT2m, method = "spearman", use = "pairwise.complete.obs"),
    pearson_r = cor(dT1, dT2m, method = "pearson", use = "pairwise.complete.obs"),
    median_retention = median(dT2m/dT1, na.rm = TRUE) # effect retention ratio; interpret with care
  )

prop_stats
```

Calculate how many bins retain at least 50% of the demethylation magnitude.

```{r retention-metric}
alpha <- 0.5  # retains at least 50% of the demethylation magnitude

t1_affected %>%
  summarise(
    retains50_n = sum(dT2m <= alpha * dT1, na.rm = TRUE),
    retains50_frac = retains50_n / n()
  )
```

Make a plot to illustrate this.

```{r inheritance-scatter, fig.height=5, fig.width=5}
ggplot(t1_affected, aes(x = dT1, y = dT2m)) +
  geom_point(alpha = 0.25, size = 0.7, shape = 16) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = delta_thr, linetype = "dashed", color = "red3") +
  geom_hline(yintercept = delta_thr, linetype = "dashed", color = "red3") +
  geom_hline(yintercept = delta_thr/2, linetype = "dashed", color = "red3", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = paste0("diff m", ctx, " (B_T1 - WT) in 100 kb bins (%)"),
       y = paste0("diff m", ctx, " (B_T2- - WT) in 100 kb bins (%)"),
       title = paste0("Stability of ", ctx, " loss from B_T1 to B_T2-"),
       subtitle = paste0("Bins defined as T1-affected: diff <= ", delta_thr, " %")) +
  coord_cartesian(xlim = c(-21, NA))

#ggsave2("../figures/mC_bins/inheritance-CG.pdf", height=5, width=5)
```
