---
title: "Scatter plots of mC and coverage in 100 kbp bins"
subtitle: "_Tomato TET3 BS-seq selected plants_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook:
    theme: flatly
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

------------------------------------------------------------------------

Analysis of DNA methylation in 100 kbp bins created with `CX_report2mC_bins.v1.0.pl` from Bismark CX_reports.

```{r libraries-and-functions, include=FALSE}
# Libraries
library(tidyverse)
library(magrittr)
library(cowplot); theme_set(theme_cowplot())
library(ggforce)
library(patchwork)
library(scales)
```

Create sample table

```{r create-sample-table}
(sample_table <- list(sample = c("WT", "A4", "A17"), replica = c(1, 2)) %>%
  expand_grid(!!!.) %>%  
  pmap_chr(paste, sep = "_") %>%
  tibble(sample = .) %>%
   mutate(condition = str_remove(sample, "_[:digit:]") %>% factor(levels = c("WT", "A4", "A17")),
          replica = str_remove(sample, ".*_") %>% as_factor()) %>%
   arrange(condition, replica) %>% 
   mutate(description = c("WT", "WT", "A.T1 TET3", "A.T1 TET3", "A.T1 NT", "A.T1 NT")))

write_tsv(sample_table, "../data/sample_table.tsv")
```

```{r read-sample-table}
sample_table <- read_tsv("../data/sample_table.tsv", col_types = cols()) %>% 
  mutate(sample = fct_inorder(sample),
         condition = fct_inorder(condition),
         replica = as_factor(replica),
         description = fct_inorder(description))
```

# Load data

When loading data we filter chromosome 9 as there is an anomaly in the mapping (and extranuclear genomes).

```{r load-bins}
bins_long_table <- read_tsv("../data/4_methylation/tomato_tet3_selected.mC_bins.100kbp.tsv", col_types = cols()) %>% 
  # remove ch09 with the anomaly and organelle genomes
  filter(!(chr %in% c("SL4.0ch09", "Mt", "Pt"))) %>% 
  # join with sample table
  mutate(sample = toupper(sample)) %>% 
  left_join(sample_table, by = "sample") %>% 
  # order the samples as a factor
  mutate(sample = factor(sample, levels = c("WT_1", "WT_2", "A4_1", "A4_2", "A17_1", "A17_2")))
```

```{r pivot-bins}
bins_wide_table <- bins_long_table %>% 
  pivot_wider(id_cols = chr:context, names_from = sample, values_from = mR, names_sort = TRUE)
```

```{r eval=FALSE}
# Write matrix for future loading
write_tsv(bins_wide_table, "../data/4_methylation/tomato_tet3_selected.mC_bins.100kbp.wide.tsv")
```

```{r eval=FALSE}
# Read matrix
bins_wide_table <- read_tsv("../data/4_methylation/tomato_tet3_selected.mC_bins.100kbp.wide.tsv", show_col_types = FALSE)
```

```{r aesthetic-variables}
# Aesthetics to plot the PCA and other plots
sample_colors <- c(WT  = "grey40", A4  = "tomato2", A17 = "slateblue2")

replica_colors <- c(WT_1  = "grey60", WT_2  = "grey30", 
                    A4_1  = "tomato1", A4_2  = "tomato3", 
                    A17_1 = "slateblue1", A17_2 = "slateblue3")

replica_shapes <- c("1" = 21, "2" = 24)
```

# Coverage check

Some quality check for the coverage levels of these 100 kbp bins

## Raincloud plot of coverage

```{r fig.width=11, fig.height=5}
bins_long_table %>% 
  mutate(coverage = reads_m + reads_n) %>% 
  ggplot(aes(y = coverage, x = sample, fill = condition)) + 
  facet_grid(cols = vars(context), scales = "free", space = "free_x") +
  ggdist::stat_halfeye(adjust = .5, width = .6, .width = 0, justification = -.2, point_colour = NA) + 
  geom_boxplot(width = .15, outlier.shape = NA, color = 1) +
  gghalves::geom_half_point(side = "l", shape = 16, size = 0.2, alpha = 0.2, range_scale = .4) +
  scale_fill_manual(values = sample_colors) + 
  scale_y_continuous(trans = "pseudo_log", breaks = c(2000, 5000, 10000, 20000, 50000, 100000, 200000),
                     labels = label_number(scale_cut = cut_short_scale())) +
  coord_cartesian(ylim = c(2000, 200000)) +
  labs(title = "Cytosine coverage for 100 kbp bins", 
       subtitle = "Sample A4_1 has a much wider distribution than the rest") +
  theme(axis.line=element_blank(), strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5), axis.title.x = element_blank())

ggsave(filename = "../figures/mC_bins/coverage-100kbp-raincloud.pdf", height = 5, width = 11)
```

## scatter plot coverage vs %mC

```{r fig.width=9, fig.height=7}
bins_long_table %>% 
  mutate(coverage = reads_m + reads_n) %>% 
  ggplot(aes(coverage, mR, color = context)) +
  facet_grid(cols = vars(condition), rows = vars(replica), scales = "free") +
  geom_point(shape = 16, size = 0.5, alpha = 0.3) +
  scale_x_continuous(trans = "pseudo_log", breaks = c(2000, 5000, 10000, 20000, 50000, 100000, 200000),
                     labels = label_number(scale_cut = cut_short_scale())) +
  coord_cartesian(xlim = c(2000, 200000)) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.7))) +
  labs(title = "Cytosine coverage for 100 kbp bins vs %mC", y = "%mC", 
       subtitle = "No clear correlation, which is a good sign") +
  theme(axis.line=element_blank(), strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), legend.position = "bottom",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
```

# Scatter plots {.tabset}

```{r}
scatter_matrix <- function(bins, condition, context = "C", title = str_c("Scatter plot of %m", context, " from ", condition)) {
  ctx <- context
  bins %>%
    select(chr:context, starts_with(condition)) %>%
    filter(str_detect(context, ctx)) %>%
    add_column(slope = 1) %>% 
    add_column(intercept = 0) %>% 
    na.omit() %>% 
    ggplot(aes(x = .panel_x, y = .panel_y)) +
    geom_point(shape = 16, size = 0.3, alpha = 0.5) +
    geom_point(shape = 16, size = 0, alpha = 0) +
    geom_density_2d_filled(contour_var = "ndensity") +
    geom_abline(aes(intercept = intercept, slope = slope), color = "red", alpha = 0.5) +
    geom_autodensity(color = 1, fill = 3) +
    facet_matrix(vars(-c(chr:context, slope, intercept)), layer.upper = c(1,4), layer.lower = c(2,3,4), layer.diag = 5) +
    theme(axis.line = element_blank(), panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), 
        strip.background = element_blank(), plot.title = element_text(hjust = 0.5), legend.position = "none") +
    ggtitle(title)
}
```

```{r fig.height=14, fig.width=12}
(scatter_matrix_bins <- scatter_matrix(bins_wide_table, "WT",  context = "CG",  title = "WT %mCG") +
   scatter_matrix(bins_wide_table, "WT",  context = "CHG", title = "WT %mCHG") +
   scatter_matrix(bins_wide_table, "WT",  context = "CHH", title = "WT %mCHH") +
   scatter_matrix(bins_wide_table, "A4",  context = "CG",  title = "A4 %mCG") +
   scatter_matrix(bins_wide_table, "A4",  context = "CHG", title = "A4 %mCHG") +
   scatter_matrix(bins_wide_table, "A4",  context = "CHH", title = "A4 %mCHH") +
   scatter_matrix(bins_wide_table, "A17", context = "CG",  title = "A17 %mCG") +
   scatter_matrix(bins_wide_table, "A17", context = "CHG", title = "A17 %mCHG") +
   scatter_matrix(bins_wide_table, "A17", context = "CHH", title = "A17 %mCHH") +
   plot_annotation(title = "Scatter plots of %mC for 100 kbp bins by context",
                   subtitle = "diagonal of the matrix represents distribution of methylation values \nbottom half show density plot of methylation values",
                   caption = "Disclaimer: y axis on diagonal matrix (histograms) doesn't match the axis labels")) 

ggsave(filename = "../figures/mC_bins/scatter-100kbp-matrices.pdf", plot = scatter_matrix_bins, height = 14, width = 12)
```

From all the Scatter plots it seems clear that something is wrong with sample A4_1. It has higher methylation than all the other samples for those bins with intermediate levels of %mC. As this is across all the genome and it is not seen in the other replica of A4, it is probably a technical problem rather than biological.

## CG

```{r fig.height=10, fig.width=9}
scatter_matrix(bins_wide_table, c("W", "A"), "CG", title = "") +
   labs(title = "Scatter plots of %mCG for 100 kbp bins",
        subtitle = "diagonal of the matrix represents distribution of methylation values \nbottom half show density plot of methylation values",
        caption = "Disclaimer: y axis on diagonal matrix (histograms) doesn't match the axis labels") 

ggsave(filename = "../figures/mC_bins/scatter-100kbp-matrix-all-CG.pdf", height = 13, width = 12)
```

## CHG

```{r fig.height=10, fig.width=9}
scatter_matrix(bins_wide_table, c("W", "A"), "CHG", title = "") +
   labs(title = "Scatter plots of %mCHG for 100 kbp bins",
        subtitle = "diagonal of the matrix represents distribution of methylation values \nbottom half show density plot of methylation values",
        caption = "Disclaimer: y axis on diagonal matrix doesn't match the axis labels") 

ggsave(filename = "../figures/mC_bins/scatter-100kbp-matrix-all-CHG.pdf", height = 13, width = 12)
```

## CHH

```{r fig.height=10, fig.width=9}
scatter_matrix(bins_wide_table, c("W", "A"), "CHH", title = "") +
   labs(title = "Scatter plots of %mCHH for 100 kbp bins",
        subtitle = "diagonal of the matrix represents distribution of methylation values \nbottom half show density plot of methylation values",
        caption = "Disclaimer: y axis on diagonal matrix doesn't match the axis labels") 

ggsave(filename = "../figures/mC_bins/scatter-100kbp-matrix-all-CHH.pdf", height = 13, width = 12)
```
