---
title: "PCAs of mC in 100 kbp bins"
subtitle: "_Tomato TET3 BS-seq selected plants_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

------------------------------------------------------------------------

Analysis of DNA methylation in 100 kbp bins created with `CX_report2mC_bins.v1.0.pl` from Bismark CX_reports.

```{r libraries-and-functions, include=FALSE}
# Libraries
library(tidyverse)
library(magrittr)
library(cowplot); theme_set(theme_cowplot())
library(ggforce)
library(patchwork)
```

# Load data

```{r read-sample-table}
sample_table <- read_tsv("../data/sample_table.tsv", col_types = cols()) %>% 
  mutate(sample = fct_inorder(sample),
         condition = fct_inorder(condition),
         replica = as_factor(replica),
         description = fct_inorder(description))
```

When loading data we filter chromosome 9 as there is an anomaly in the mapping (and extranuclear genomes). We also remove those bins with low coverage (< 1000 reads / 100 kbp)

```{r load-bins}
bins_long_table <- read_tsv("../data/4_methylation/tomato_tet3_selected.mC_bins.100kbp.tsv", col_types = cols()) %>% 
  # remove ch09 with the anomaly and organelle genomes
  filter(!(chr %in% c("SL4.0ch09", "Mt", "Pt"))) %>% 
  # join with sample table
  mutate(sample = toupper(sample)) %>% 
  left_join(sample_table, by = "sample") %>% 
  # order the samples as a factor
  mutate(sample = factor(sample, levels = c("WT_1", "WT_2", "A4_1", "A4_2", "A17_1", "A17_2"))) %>% 
  # remove bins with low coverage (> 1000 reads)
  filter((reads_m + reads_n) > 1000)
```

```{r pivot-bins}
bins_wide_table <- bins_long_table %>% 
  pivot_wider(id_cols = chr:context, names_from = sample, values_from = mR, names_sort = TRUE)
```

```{r aesthetic-variables}
# Aesthetics to plot the PCA and other plots
sample_colors <- c(WT  = "grey40", A4  = "tomato2", A17 = "slateblue2")

replica_colors <- c(WT_1  = "grey60", WT_2  = "grey30", 
                    A4_1  = "tomato1", A4_2  = "tomato3", 
                    A17_1 = "slateblue1", A17_2 = "slateblue3")

replica_shapes <- c("1" = 21, "2" = 24)
```

# PCAs 

```{r}
# General function for all DNA methylation data that takes table 
#  (with format `chr`, `start`, `end`, `context`, [samples]) 
#  and returns a table of Principal Components 
pca_tidy_mC <- function(mC_table, context, ntop = 5000, sampletable = NULL) {
  ctx <- context
  # pick the top n observations by variance
  top_obs <- mC_table %>%
    filter(str_detect(context, ctx)) %>% # filter by context (including C for all)
    rowwise(chr:context) %>% # calculate the variance for each observation
    mutate(var = var(c_across(everything()))) %>%
    ungroup() %>%
    slice_max(var, n = ntop) %>% # pick only the top n
    select(-var)
  
  # perform a PCA on the data
  pca <- top_obs %>%
    select( -(chr:context) ) %>% 
    na.omit() %>% 
    t() %>% 
    prcomp()
  
  # convert it to a table
  pca$x %>%
    as_tibble(rownames = "sample") %>%
    rename_with(~ pca$sdev^2 %>%        # square the sd to get the variance
                  divide_by(sum(.)) %>% # divide it by the sum of variances
                  multiply_by(100) %>%  # *100 to get a percentage of variance
                  round() %>%           # round it
                  str_c(.x, " [", ., "%]"), # paste it to the PCx 
                .cols = -sample) %>% # exclude the sample column 
    {
      if (is.null(sampletable)) .
      else left_join(., sampletable, by = "sample")
    }
}
```

```{r}
# Function to make a matrix of PCA plots
pca_matrix = function(pctable, pcs = 1:3, title = NULL) {
  ggplot(pctable, aes(x = .panel_x, y = .panel_y)) + 
    theme_cowplot() +
    theme(axis.line=element_blank(), panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), strip.background = element_blank(),
          plot.title = element_text(hjust = 0.5)) +
    scale_shape_manual(values = replica_shapes) +
    scale_fill_manual(values = sample_colors) +
    guides(fill = guide_legend(override.aes = list(shape = 22)), 
           shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
    geom_mark_hull(aes(fill = condition), radius = 0.15, concavity = 1, expand = 0.1) +
    #geom_mark_ellipse(aes(fill = condition)) +
    geom_autodensity(aes(fill = condition), color = "black", alpha = 0.5, show.legend = FALSE) +
    #geom_autohistogram(aes(fill = condition), color="black", size = 0.6, alpha=0.7, show.legend = FALSE) +
    geom_point(size=2, color="black", aes(fill = condition, shape = replica)) +
    facet_matrix(vars(pcs +1), layer.lower = c(1, 3), layer.upper = 3, layer.diag = 2) +
    ggtitle(title)
}
```

### CG PCA table {.tabset .tabset-pills}

#### .

#### show table

```{r}
pca_tidy_mC(bins_wide_table, "CG", sampletable = sample_table)
```

# Using the PCAs to remove ouliers {.tabset}

We can identify outlier samples by looking at the PCAs, those samples that constitute one Principal Component on their own will probably be outliers.

## All samples {.tabset .tabset-pills}

```{r, include=FALSE, fig.height=6, fig.width=16}
pca_plot <- pca_tidy_mC(bins_wide_table, "CG", sampletable = sample_table) %>% pca_matrix(title = "CG") +
   pca_tidy_mC(bins_wide_table, "CHG", sampletable = sample_table) %>% pca_matrix(title = "CHG") +
   pca_tidy_mC(bins_wide_table, "CHH", sampletable = sample_table) %>% pca_matrix(title = "CHH") +
   pca_tidy_mC(bins_wide_table, "C", sampletable = sample_table) %>% pca_matrix(title = "all Cs") +
   plot_annotation(title = "Matrix of PCs of %mC for 100 kbp bins",
                   subtitle = "Percentage of variance explained by each PC in square brackets",
                   caption = "Let's see what to write here") +
  plot_layout(guides = 'collect')

dir.create("../figures/mC_bins", showWarnings = FALSE)
ggsave(filename = "../figures/mC_bins/PCA-100kb-matrices.pdf", plot = pca_plot, width = 12, height = 12)
```

Most of the dispersion is due to A4_1, which we already saw in the scatter plots that is a strange one

### mC

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table, "C", sampletable = sample_table) %>% 
  pca_matrix(pcs = 1:3, title = "%mC")
```

### mCG {.active}

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table, "CG", sampletable = sample_table) %>% 
  pca_matrix(1:3, title = "%mCG")
```

### mCHG

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table, "CHG", sampletable = sample_table) %>% 
  pca_matrix(1:3, title = "%mCHG")
```

### mCHH

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table, "CHH", sampletable = sample_table) %>% 
  pca_matrix(1:3, title = "%mCHH")
```


## Removing A4_1 {.tabset .tabset-pills .active}

After removing sample A4_1 

```{r}
bins_wide_table_no_A4_1 <- bins_wide_table %>%
  select(-A4_1)
```

```{r, include=FALSE, fig.height=6, fig.width=16}
pca_plot <- pca_tidy_mC(bins_wide_table_no_A4_1, "CG", sampletable = sample_table) %>% pca_matrix(title = "CG") +
   pca_tidy_mC(bins_wide_table_no_A4_1, "CHG", sampletable = sample_table) %>% pca_matrix(title = "CHG") +
   pca_tidy_mC(bins_wide_table_no_A4_1, "CHH", sampletable = sample_table) %>% pca_matrix(title = "CHH") +
   pca_tidy_mC(bins_wide_table_no_A4_1, "C", sampletable = sample_table) %>% pca_matrix(title = "all Cs") +
   plot_annotation(title = "Matrix of PCs of %mC for 100 kbp bins",
                   subtitle = "Percentage of variance explained by each PC in square brackets",
                   caption = "After removing A4_1") +
  plot_layout(guides = 'collect')

ggsave(filename = "../figures/mC_bins/PCA-100kb-matrices-no-A4_1.pdf", plot = pca_plot, width = 12, height = 12)
```

### mC

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table_no_A4_1, "C", sampletable = sample_table) %>% 
  pca_matrix(pcs = 1:3, title = "%mC")
```

### mCG {.active}

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table_no_A4_1, "CG", sampletable = sample_table) %>% 
  pca_matrix(1:3, title = "%mCG")
```

### mCHG

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table_no_A4_1, "CHG", sampletable = sample_table) %>% 
  pca_matrix(1:3, title = "%mCHG")
```

### mCHH

```{r fig.height=5, fig.width=6}
pca_tidy_mC(bins_wide_table_no_A4_1, "CHH", sampletable = sample_table) %>% 
  pca_matrix(1:3, title = "%mCHH")
```

# Correlation between PC1 and %mC {.tabset .tabset-pills}
  
From the PCAs there seems to be a correlation between CG methylation loss and position across the PC1. To check if it is true I compared the average methylation across the genome (mean of all 100 kbp bins) to the position across PC1 and calculate the correlation using a linear model. The Pearson correlation coefficient is 0.98, so PC1 is reflecting global CG methylation levels. I tried this for the other context and the correlation is also perfect.

## CG

```{r fig.height=5, fig.width=7}
pca_tidy_mC(bins_wide_table, "CG", sampletable = sample_table) %>% 
  select(sample, PC1 = `PC1 [80%]`, condition, replica) %>% 
  left_join(bins_long_table %>% 
              filter(context == "CG") %>% 
              group_by(sample) %>% 
              summarise(mean_mR = mean(mR)),
            by = "sample") %>% 
  ggplot(aes(x = PC1, y = mean_mR)) + 
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_mark_hull(aes(fill = condition), radius = unit(5, "mm"), 
                concavity = 9, con.cap = 0, label.fill = "grey95") +
  geom_smooth(method = "lm", se=FALSE, formula = y ~ x, linewidth = 3, alpha = 0.1, color = "darkorchid1") +
  geom_point(size=3, color="black", aes(fill = condition, shape = replica)) +
  ggpubr::stat_regline_equation(label.y = 79.5, label.x = -280, color = "darkorchid3") +
  ggpubr::stat_cor(label.y = 79, label.x = -280, color = "darkorchid3") +
  scale_shape_manual(values = replica_shapes) +
  scale_fill_manual("Line", values = sample_colors) +
  labs(title = "Relation between global %mCG and PC1 for 100 kbp bins", y = "global %mCG",
       subtitle = "There is a strong correlation between PC1 and %mCG",) +
  guides(fill = guide_legend(override.aes = list(shape = 22, color = NA)), 
         shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
  theme(axis.line=element_blank(), panel.border = element_rect(colour="black", linewidth=1, fill=NA))
ggsave2(filename = "../figures/mC_bins/PCA-100kb-CG-PC1-mCG-correlation.pdf", width = 7, height = 5)
```

## CHG

```{r fig.height=5, fig.width=7}
pca_tidy_mC(bins_wide_table, "CHG", sampletable = sample_table) %>% 
  select(sample, PC1 = `PC1 [85%]`, condition, replica) %>% 
  left_join(bins_long_table %>% 
              filter(context == "CHG") %>% 
              group_by(sample) %>% 
              summarise(mean_mR = mean(mR)),
            by = "sample") %>% 
  ggplot(aes(x = PC1, y = mean_mR)) + 
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_mark_hull(aes(fill = condition), radius = unit(5, "mm"), 
                concavity = 9, con.cap = 0, label.fill = "grey95") +
  geom_smooth(method = "lm", se=FALSE, formula = y ~ x, linewidth = 3, alpha = 0.1, color = "darkorchid1") +
  geom_point(size=3, color="black", aes(fill = condition, shape = replica)) +
  ggpubr::stat_regline_equation(label.y = 55, label.x = -500, color = "darkorchid3") +
  ggpubr::stat_cor(label.y = 54, label.x = -500, color = "darkorchid3") +
  scale_shape_manual(values = replica_shapes) +
  scale_fill_manual("Line", values = sample_colors) +
  labs(title = "Relation between global %mCHG and PC1 for 100 kbp bins", y = "global %mCHG",
       subtitle = "There is a strong correlation between PC1 and %mCG",) +
  guides(fill = guide_legend(override.aes = list(shape = 22, color = NA)), 
         shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
  theme(axis.line=element_blank(), panel.border = element_rect(colour="black", linewidth=1, fill=NA))
```

## CHH

```{r fig.height=5, fig.width=7}
pca_tidy_mC(bins_wide_table, "CHH", sampletable = sample_table) %>% 
  select(sample, PC1 = `PC1 [97%]`, condition, replica) %>% 
  left_join(bins_long_table %>% 
              filter(context == "CHH") %>% 
              group_by(sample) %>% 
              summarise(mean_mR = mean(mR)),
            by = "sample") %>% 
  ggplot(aes(x = PC1, y = mean_mR)) + 
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_mark_hull(aes(fill = condition), radius = unit(5, "mm"), 
                concavity = 9, con.cap = 0, label.fill = "grey95") +
  geom_smooth(method = "lm", se=FALSE, formula = y ~ x, linewidth = 3, alpha = 0.1, color = "darkorchid1") +
  geom_point(size=3, color="black", aes(fill = condition, shape = replica)) +
  ggpubr::stat_regline_equation(label.y = 12.5, label.x = -250, color = "darkorchid3") +
  ggpubr::stat_cor(label.y = 12, label.x = -250, color = "darkorchid3") +
  scale_shape_manual(values = replica_shapes) +
  scale_fill_manual("Line", values = sample_colors) +
  labs(title = "Relation between global %mCHH and PC1 for 100 kbp bins", y = "global %mCHH",
       subtitle = "There is a strong correlation between PC1 and %mCG",) +
  guides(fill = guide_legend(override.aes = list(shape = 22, color = NA)), 
         shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
  theme(axis.line=element_blank(), panel.border = element_rect(colour="black", linewidth=1, fill=NA))
```
