---
title: "Boxplots of mC in 100 kbp bins"
subtitle: "_Tomato TET3 BS-seq selected plants_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

------------------------------------------------------------------------

Analysis of DNA methylation in 100 kbp bins created with `CX_report2mC_bins.v1.0.pl` from Bismark CX_reports.

```{r libraries-and-functions, include=FALSE}
# Libraries
library(tidyverse)
library(cowplot); theme_set(theme_cowplot())
library(patchwork)
library(ggbeeswarm)
library(scales)
```

# Load and filter data

```{r read-sample-table}
sample_table <- read_tsv("../data/sample_table.tsv", col_types = cols()) %>% 
  mutate(sample = fct_inorder(sample),
         condition = fct_inorder(condition),
         replica = as_factor(replica),
         description = fct_inorder(description))
```

When loading data we filter chromosome 9 as there is an anomaly in the mapping (and extranuclear genomes). We also remove those bins with low coverage (\< 1000 reads / 100 kbp)

```{r load-bins}
bins_long_table <- read_tsv("../data/4_methylation/tomato_tet3_selected.mC_bins.100kbp.tsv", col_types = cols()) %>% 
  # remove ch09 with the anomaly and organelle genomes
  filter(!(chr %in% c("SL4.0ch09", "Mt", "Pt"))) %>% 
  # join with sample table
  mutate(sample = toupper(sample)) %>% 
  left_join(sample_table, by = "sample") %>% 
  # order the samples as a factor
  mutate(sample = factor(sample, levels = c("WT_1", "WT_2", "A4_1", "A4_2", "A17_1", "A17_2"))) %>% 
  # remove bins with low coverage (> 1000 reads)
  filter((reads_m + reads_n) > 1000)
```

```{r aesthetic-variables}
# Aesthetics to plot the PCA and other plots
sample_colors <- c(WT  = "grey40", A4  = "tomato2", A17 = "slateblue2")

replica_colors <- c(WT_1  = "grey60", WT_2  = "grey30", 
                    A4_1  = "tomato1", A4_2  = "tomato3", 
                    A17_1 = "slateblue1", A17_2 = "slateblue3")

replica_shapes <- c("1" = 21, "2" = 24)
```

For each bin I computed the average across the WT replicas and obtain the difference to that from each sample.

```{r}
pivot_longer_diff_to_reference <- function(bins_wide, reference = WT_mean, values = mR, remove_reference = TRUE) {
  bins_wide %>%
    #rename(ref = {{ reference }}) %>%
    separate_wider_delim(WT_mean, delim = "fake_dlim", names = c("ref"), cols_remove = remove_reference) %>% 
    pivot_longer(-c(chr:context,ref), names_to = "sample", values_to = as.character(ensym(values)),
                 names_transform = list(sample = forcats::fct_inorder)) #%>%
#    mutate(diff = {{ values }} - ref)
}

# this version works for the merged table
pivot_longer_diff_to_reference <- function(bins_wide, reference = WT_mean, values = mR, remove_reference = TRUE) {
  bins_wide %>%
    rename(ref = {{ reference }}) %>%
    separate_wider_delim(ref, delim = "fake_dlim", names = c("ref"), cols_remove = remove_reference) %>% 
    pivot_longer(-c(chr:context,ref), names_to = "sample", values_to = as.character(ensym(values)),
                 names_transform = list(sample = forcats::fct_inorder)) %>%
    mutate(diff = {{ values }} - ref)
}
```

```{r}
bins_long_table_diff <- bins_wide <- bins_long_table %>% 
  pivot_wider(id_cols = chr:context, names_from = sample, values_from = mR, names_sort = TRUE) %>% 
  mutate(WT_mean = (WT_1 + WT_2) / 2) %>%
  pivot_longer_diff_to_reference() %>% 
  left_join(sample_table, by = "sample")
```

```{r}
bins_long_table %>% 
  pivot_wider(id_cols = chr:context, names_from = sample, values_from = mR, names_sort = TRUE) %>% 
  mutate(WT_mean = (WT_1 + WT_2) / 2) %>%
  pivot_longer_diff_to_reference()
```

# All samples

## Comparing replicas {.tabset .tabset-pills}

Each boxplot represent one condition, and each dot the average for each replica

### %mC

```{r fig.width=5, fig.height=4}
bins_long_table_diff %>% 
  group_by(condition, replica, sample, context) %>% 
  summarise(mR_avg = mean(mR, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(condition, mR_avg, fill = condition)) +
  facet_wrap(vars(context), scales = "free") +
  geom_boxplot(outlier.colour = NA, color = "black", alpha = 0.3, width = 0.9) +
  geom_quasirandom(aes(shape = replica), size = 3, dodge.width = 0.75) +
  scale_shape_manual(values = replica_shapes) +
  scale_fill_manual(values = sample_colors) + 
  guides(fill = "none", shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
  ylab("% mC") +
  theme(legend.position = "bottom", 
        strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.title.x = element_blank()) + 
  labs(title = "Methylation levels of TET3 plants",
       subtitle = "using average for each plant", 
       caption = "each dot represents average methylation rate for the whole genome")

ggsave2("../figures/mC_bins/boxplot-100kb-summarised.pdf", width = 5, height = 4)
```

### %mC diff {.active}

```{r fig.width=5, fig.height=4}
bins_long_table_diff %>% 
  group_by(condition, replica, sample, context) %>% 
  summarise(mR_avg = mean(diff, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(condition, mR_avg, fill = condition)) +
  facet_wrap(vars(context)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot(outlier.colour = NA, color = "black", alpha = 0.3, width = 0.9) +
  geom_quasirandom(aes(shape = replica), size = 3, dodge.width = 0.75) +
  scale_shape_manual(values = replica_shapes) +
  scale_fill_manual(values = sample_colors) + 
  guides(fill = "none", shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
  ylab("% mC difference to WT average") +
  theme(legend.position = "bottom", 
        strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.title.x = element_blank()) + 
  labs(title = "Methylation levels of TET3 plants",
       subtitle = "using average for each plant", 
       caption = "each dot represents average methylation rate for the whole genome")

ggsave2("../figures/mC_bins/boxplot-100kb-summarised-diff.pdf", width = 5, height = 4)
```

## Distribution of mC levels {.tabset .tabset-pills}

Each boxplot represent one replica, with the violin showing the whole set of 100 kb bins

### %mC

```{r fig.width=6, fig.height=7}
ggplot(bins_long_table_diff, aes(sample, mR, fill = condition)) +
  facet_grid(cols = vars(context)) +
  geom_violin(scale = "count", alpha = 0.5, color = NA) + 
  geom_boxplot(outlier.shape = NA, coef = 0, notch = TRUE, color = "black", alpha = 0.3, width = 0.3) +
  stat_summary(geom = "point", fun = mean, shape = 23, color = "black", size = 3) +
  scale_fill_manual(values = sample_colors) + 
  labs(title = "Methylation levels of TET3 plants", y = "% mC",
       subtitle = "using averages over 100 kbp bins", 
       caption = "Dots represent average of methylation rate") +
  theme(legend.position = "none", strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5), axis.title.x = element_blank())

ggsave2("../figures/mC_bins/boxplot-100kb.pdf", width = 6, height = 7)
```

### Difference

```{r fig.width=6, fig.height=7}
ggplot(bins_long_table_diff, aes(sample, diff, fill = condition)) +
  facet_grid(cols = vars(context)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_violin(scale = "count", alpha = 0.5, color = NA) + 
  geom_boxplot(outlier.shape = NA, coef = 0, notch = TRUE, color = "black", alpha = 0.3, width = 0.3) +
  stat_summary(geom = "point", fun = mean, shape = 23, color = "black", size = 3) +
  scale_fill_manual(values = sample_colors) + 
  coord_cartesian(ylim = c(-12, 12)) +
  labs(title = "Methylation levels of TET3 plants", y = "% mC difference to WT average",
       subtitle = "using averages over 100 kbp bins", 
       caption = "Dots represent average of methylation rate") +
  theme(legend.position = "none", strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5), axis.title.x = element_blank())

ggsave2("../figures/mC_bins/boxplot-100kb-diff.pdf", width = 6, height = 7)
```

# Merging replicas and removing A4_1

From these boxplots and the previous graphs showing coverage, correlation of %mC and the PCAs we can conclude that sample A4_1 is a bad replica and should be excluded from the further analysis.

```{r merge-replicas}
bins_long_table_merged_only_sample <- bins_long_table %>% 
  # pivot keeping reads_n and reads_m
  pivot_wider(id_cols = chr:context, names_from = sample, values_from = mR:reads_n) %>% 
  replace(is.na(.), 0) %>% 
  # remove mR columns
  select(-starts_with("mR_")) %>% 
  rowwise() %>% 
  # sum the reads excluding A4_1
  mutate(reads_m_WT = sum(reads_m_WT_1, reads_m_WT_2, na.rm = TRUE),
         reads_m_A4  = reads_m_A4_2,
         reads_m_A17 = sum(reads_m_A17_1, reads_m_A17_2, na.rm = TRUE),
         
         reads_n_WT = sum(reads_n_WT_1, reads_n_WT_2, na.rm = TRUE),
         reads_n_A4 = reads_n_A4_2,
         reads_n_A17 = sum(reads_n_A17_1, reads_n_A17_2, na.rm = TRUE)) %>% 
  #ungroup() %>% 
  # remove unmerged columns
  select(-(reads_m_A17_1:reads_n_WT_2)) %>% 
  # pivot to separate into 2 columns for m and n
  pivot_longer(
    cols = reads_m_WT:reads_n_A17,
    #cols_vary = "slowest", 
    names_to = c(".value", "sample"), 
    names_pattern = "(reads_.)_(.*)") %>% 
  # compute methylation rate
  mutate(mR = 100 * reads_m / (reads_m + reads_n)) %>% 
  relocate(mR, .before = reads_m) %>% 
  relocate(sample, .after = last_col()) %>% 
  # order the samples as a factor and 
  mutate(sample = factor(sample, levels = c("WT", "A4", "A17")))
```


```{r}
write_tsv(bins_long_table_merged_only_sample, "../data/4_methylation/tomato_tet3_selected.mC_bins.100kbp.merged.tsv")
```

```{r add-description}
bins_long_table_merged <- bins_long_table_merged_only_sample %>% 
  # add description
  mutate(description = case_when(sample == "WT" ~ "WT", 
                                 sample == "A4" ~ "A.T1\nTET3", 
                                 sample == "A17" ~ "A.T1\nNT") %>% 
           fct_inorder())

```

For each bin I computed the difference from each sample to WT.

```{r}
bins_long_table_merged_diff <- bins_long_table_merged %>% 
  pivot_wider(id_cols = chr:context, names_from = sample, values_from = mR, names_sort = TRUE) %>% 
  pivot_longer_diff_to_reference(reference = WT) %>% 
  # add description
  mutate(description = case_when(sample == "A4" ~ "A.T1\nTET3", 
                                 sample == "A17" ~ "A.T1\nNT") %>% 
           fct_inorder())
```

## Boxplots whole genome

Both A4 (A.T1 TET3) and A17 (A.T1 NT) have lower methylation than WT.

For CG and CHG a big majority of the 100 kbp bins have less methylation in A.T1 TET3 and A.T1 NT than in WT. A.T1 TET3 loses around 2.5% on average while A.T1 NT loses around 5%.

This is interesting because we see the biggest drop in methylation in the genetically WT plant.

For CHG in A.T1 NT we see a bimodal distribution of differences, with one of the peaks being the WT levels. This suggest than some regions lose methylation while others stay as in WT.

in CHH the changes are smaller, and it is not clear if A.T1 TET3 has less methylation than WT, while A.T1 NT has a drop around 2%.

```{r fig.width = 6, fig.height = 7}
((ggplot(bins_long_table_merged, aes(description, mR, fill = sample)) +
   labs(y = "% mC")) /
  (ggplot(bins_long_table_merged_diff, aes(description, diff, fill = sample)) + 
     geom_hline(yintercept = 0, linetype = "dashed") +
     stat_summary(geom = "col", fun = mean, color = NA, alpha = 0.9, width = 0.15) +
     coord_cartesian(ylim = c(-12, 5)) +
     labs(y = "% mC difference to WT")) &
  facet_grid(cols = vars(context)) &
  geom_violin(scale = "count", alpha = 0.5, color = NA) & 
  geom_boxplot(outlier.shape = NA, coef = 0, notch = TRUE, color = "black", alpha = 0.3, width = 0.3) &
  stat_summary(geom = "point", fun = mean, shape = 23, color = "black", size = 3) &
  scale_fill_manual(values = sample_colors) &
  theme(legend.position = "none", strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.title.x = element_blank())) +
  plot_annotation(title = "Methylation levels of TET3 plants",
                   subtitle = "using averages over 100 kbp bins",
                   caption = "Dots represent average of methylation rate")
  
ggsave2("../figures/mC_bins/boxplot-merged-100kb.pdf", width = 6, height = 7)
```

## Split by chromatin state

Using information about location of boundaries between Euchromatin and Heterochromatin inferred from H3K27mac ChIP-seq data, we can compare the changes in DNA methylation between the two compartments of the genome.

```{r fig.width=6, fig.height=11}
hchrom_H3K27ac_100kbp_bins <- read_tsv("/path/to/genome/SL4.0_heterochromatin_100kbp_bins_H3K27ac_chXX.tsv", show_col_types = FALSE) %>% 
  mutate(seqnames = str_c("SL4.0", seqnames),
         start = start - 1) %>% 
  dplyr::rename(chr = seqnames)
```

We can see the big differences in % mC between the two compartments, which are most striking for CHG context.

For A.T1 TET3 the patterns are similar to what we saw in the low input WGBS: CG demethylation happening in both compartments while CHG mainly in Heterochromatin.

for A.T1 NT we see an even stronger demethylation of the Heterochromatin that extends to CG context, and that explains the bigger global loss compared to A.T1 TET3 and the bimodal distribution for CHG.

For CHH the patterns look similar but much smaller

```{r fig.width=9, fig.height=7}
((bins_long_table_merged %>% 
   left_join(hchrom_H3K27ac_100kbp_bins, by = c("chr", "start", "end")) %>% 
   filter(!is.na(chromatin)) %>% 
   ggplot(aes(description, mR, fill = sample)) +
   ylab("% mC")) /
  (bins_long_table_merged_diff %>% 
     left_join(hchrom_H3K27ac_100kbp_bins, by = c("chr", "start", "end")) %>% 
     filter(!is.na(chromatin)) %>% 
     ggplot(aes(description, diff, fill = sample)) +
     geom_hline(yintercept = 0, linetype = "dashed") +
     stat_summary(geom = "col", fun = mean, color = NA, alpha = 0.9, width = 0.15) +
     coord_cartesian(ylim = c(-10, 1)) + 
     ylab("% mC difference to WT")) &
  facet_grid(cols = vars(context, chromatin)) &
  geom_violin(scale = "count", alpha = 0.5, color = NA) &
  geom_boxplot(outlier.shape = NA, coef = 0, notch = TRUE, color = "black", alpha = 0.3, width = 0.3) &
  stat_summary(geom = "point", fun = mean, shape = 23, color = "black", size = 3) &
  scale_fill_manual(values = sample_colors) &
  theme(legend.position = "none", strip.background = element_rect(colour = NA, fill = "lightpink"),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.title.x = element_blank())) +
  plot_annotation(title = "Methylation levels of TET3 plants by Chromatin compartment",
                   subtitle = "using averages over 100 kbp bins",
                   caption = "Dots represent average of methylation rate")

ggsave2("../figures/mC_bins/boxplot-merged-100kb-chromatin.pdf", width = 7, height = 7)
```
