---
title: "DMRs with CHG using bins method"
subtitle: "_Tomato TET3 BS-seq reanalysis_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: pygments
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
    code_folding: hide
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = TRUE)
```

# Filtering and data writing in Mac

```{r include=FALSE}
## Mac R 
# libraries ----
library(plyranges)
library(tidyverse)
```

```{r dmrs-make-tibble}
dmrs_whole_genome <- read_rds("../data/6_dmrs-2025/dmrs_whole_genome_A4_A17_CHG_relaxed.rds")

# combine into a table with all DMRs together
tb_dmrs_unfiltered <- filter(dmrs_whole_genome[[1]][[1]], seqnames == "empty") %>% 
  as_tibble()
for (sample in names(dmrs_whole_genome)) {
  for (context in c("CHG")) {
    tb_dmrs_unfiltered <- bind_rows(
      tb_dmrs_unfiltered,
      dmrs_whole_genome[[sample]][[context]] %>%
        as_tibble() %>% 
        mutate(chr = str_remove(seqnames, "SL4.0"),
               comparison = case_when(sample == "a4"  ~ "TET3",
                                      sample == "a17" ~ "NTS"))
    )
  }
}

# Filter p-val and C count as a list of GRanges (list of lists)
dmrs <- list(
  a17 = map(dmrs_whole_genome$a17, ~ filter(.x, pValue < 0.01, cytosinesCount >= 4)),
  a4  = map(dmrs_whole_genome$a4,  ~ filter(.x, pValue < 0.01, cytosinesCount >= 4))
)

# Filter p-val and C count as a table with all DMRs combined
tb_dmrs <- tb_dmrs_unfiltered %>% 
  filter(pValue < 0.01, cytosinesCount >= 4)

# Remove DMRs on ch09, as it has some abnormalities: I DECIDED NOT TO FO THIS
# dmrs_no_ch09 <- list(
#  a17 = map(dmrs$a17, ~ filter(.x, !(seqnames %in% c("SL4.0ch09", "Mt", "Pt")))),  
#  a4  = map(dmrs$a4,  ~ filter(.x, !(seqnames %in% c("SL4.0ch09", "Mt", "Pt"))))
#)
#save(dmrs_no_ch09, file = "../data/6_dmrs/dmrs-whole-genome-a17-a4-filtered-no-ch09.Rdata")

# Save them for future use
write_rds(dmrs, file = "../data/6_dmrs-2025/dmrs_whole_genome_A4_A17_CHG_filtered.rds")

write_tsv(tb_dmrs_unfiltered, "../data/6_dmrs-2025/table-dmrs-whole-genome_A4_A17_CHG_not-filtered.tsv")
write_tsv(tb_dmrs, "../data/6_dmrs-2025/table-dmrs-whole-genome_A4_A17_CHG_filtered.tsv")
```

Save DMRs in a more modern format

```{r}
tb_dmrs <- read_tsv("../data/6_dmrs-2025/table-dmrs-whole-genome_A4_A17_CHG-bins-combined_filtered.tsv")

dmrs_list_df <- tb_dmrs %>% 
  mutate(comparison = factor(comparison, levels = c("TET3", "NTS"))) %>% 
  split(f = .$comparison) %>% 
  map(~ split(.x, f = .x$context)) %>% 
  map_depth(2, ~ split(.x, f = .x$regionType))

dmrs_list_gr <- dmrs_list_df %>% 
  map_depth(3, as_granges)

dmrs_df <- dmrs_list_df %>%
  map_depth(2, list_rbind) %>%
  map_depth(1, list_rbind) %>%
  map_depth(0, list_rbind) 
  
write_rds(dmrs_list_df, "../data/6_dmrs-2025/DMR-list_TET3_NTS_CHG-combined_dataframe.rds")
write_rds(dmrs_list_gr, "../data/6_dmrs-2025/DMR-list_TET3_NTS_CHG-combined_GRanges.rds")
write_tsv(dmrs_df, "../data/6_dmrs-2025/DMRs-TET3_NTS_CHG-combined_tibble.tsv")
```

## Merge data with previous DMRs

```{r}
tb_dmrs_filtered_CHG <- read_tsv("../data/6_dmrs-2025/table-dmrs-whole-genome_A4_A17_CHG_filtered.tsv", show_col_types = FALSE) 
tb_dmrs_filtered_CG_CHH <- read_tsv("../data/6_dmrs/table-dmrs-whole-genome-a17-a4-filtered.tsv", show_col_types = FALSE) %>% 
  filter(context %in% c("CG", "CHH")) %>% 
  mutate(comparison = case_when(comparison == "A.T1-TET3"  ~ "TET3",
                                comparison == "A.T1-NT" ~ "NTS") %>% 
           factor(levels = c("TET3", "NTS")))

tb_dmrs_filtered <- bind_rows(tb_dmrs_filtered_CG_CHH, tb_dmrs_filtered_CHG)

write_tsv(tb_dmrs_filtered, "../data/6_dmrs-2025/table-dmrs-whole-genome_A4_A17_CHG-bins-combined_filtered.tsv")
```

Save DMRs as GFF and BED files for compatibility

```{r write-DMRs-as-gff-and-bed}
for (compared_sample in c("NTS", "TET3")) {
  contexts <- c(CN = "C", CG = "CG", CHG = "CHG", CHH = "CHH")
  for (context_name in names(contexts)) {
    regiontypes <- c(all = ".*", gain = "gain", loss = "loss")
    for (region_name in names(regiontypes)) {
      region_value <- regiontypes[region_name]
      context_value <- contexts[context_name]
      
      file_name <- str_c("../data/6_dmrs-2025/gffs/DMRs_", compared_sample, "_", context_name, "_", region_name)
      
      # Filter the data and convert to GRanges object
      granges_data <- tb_dmrs_filtered %>%
        filter(comparison == compared_sample, 
               str_detect(context, context_value),  
               str_detect(regionType, region_value)) %>%
        mutate(type = "DMR", source = "Javier") %>%
        as_granges()
      
      # Check if the GRanges object is non-empty before proceeding
      if (length(granges_data) > 0) {
        # Write to .bed and .gff3 independently
        write_bed(granges_data, str_c(file_name, ".bed"))
        write_gff(granges_data, str_c(file_name, ".gff3"))
      } else {
        message("No data for ", file_name, " - skipping write.")
      }
    }
  }
}
```

# Plots

```{r mac-libraries-functions-variables, include=FALSE}
## Mac R 
# libraries ----
library(plyranges)
library(tidyverse)
library(cowplot); theme_set(theme_cowplot())
library(scales)
library(patchwork)
library(ggforce)
library(ggplotify)
library(ComplexHeatmap)
ht_opt$message = FALSE
library(shadowtext)
library(GeneOverlap)
library(RColorBrewer)
library(kableExtra)
library(slider)
library(tidyHeatmap)
library(ggblend)
library(ggrastr)
```

```{r aesthetic-variables}
# they have to be in alphabetical order for tidyheatmap
comparison_colors  <- c("NTS"  = "slateblue2", "TET3"  = "tomato2", WT  = "grey40")
```

## Column plots {.tabset}

```{r}
tb_dmrs_filtered <- read_tsv("../data/6_dmrs-2025/table-dmrs-whole-genome_A4_A17_CHG-bins-combined_filtered.tsv", show_col_types = FALSE) %>% 
  mutate(comparison = factor(comparison, levels = c("TET3", "NTS")))
```

```{r}
DMR_numbers <- tb_dmrs_filtered %>%
  mutate(comparison = fct_inorder(comparison)) %>% 
  group_by(comparison, context,  regionType) %>%
  summarise(DMRs = n(), length = sum(width), .groups = "drop") %>% 
  dplyr::rename(kind = regionType) %>%
  mutate(kind = factor(kind, levels = c("gain","loss")),
         comparison = factor(comparison, levels = c("TET3", "NTS")))
```

### Numbers

```{r col-plot-numbers, fig.height=5, fig.width=8}
(ggplot(DMR_numbers %>% filter(kind == "loss"), aes(x = DMRs, y = comparison, fill = comparison, label = DMRs)) +
  facet_grid(rows = vars(context), cols = vars(kind), switch = "y") +
  scale_y_discrete(position = "right", limits = rev) +
  scale_x_reverse(labels = scales::label_number(scale_cut = cut_si("")), expand = expansion(mult = c(0.05, 0))) +
  theme(axis.text.y = element_blank()) +
  
  ggplot(DMR_numbers %>% filter(kind == "gain"), aes(x = DMRs, y = comparison, fill = comparison, label = DMRs)) +
  facet_grid(rows = vars(context), cols = vars(kind)) +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(labels = scales::label_number(scale_cut = cut_si("")), expand = expansion(mult = c(0, 0.05))) +
  theme(axis.text.y = element_text(hjust = 0.5)) &
  
  geom_col() &
  geom_text(position = position_stack(vjust = 0.5)) &
  coord_fixed(ratio = 4000) &
  scale_fill_manual(values = comparison_colors) &
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),legend.position= "none",
        axis.ticks.y = element_blank(), axis.title.y = element_blank(), axis.line = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA))) +
  
  plot_annotation(title = "Differentially Methylated Regions (DMRs). NUMBER",
                  subtitle = "CG: Noise filter method, mpd = 0.5
CHG: Bins method, mpd = 0.5
CHH: Bins method, mpd = 0.3") +
  plot_layout(widths = c(1, 1.3))

ggsave2("../figures/dmrs-2026-CHG-bins/colplot-DMRs-numbers.pdf", width = 8, height = 5)
```

### Length

```{r col-plot-lengths, fig.height=5, fig.width=8}
(ggplot(DMR_numbers %>% filter(kind == "loss"), aes(x = length, y = comparison, fill = comparison, label = str_c(round(length/1000, 0), " kb"))) +
  facet_grid(rows = vars(context), cols = vars(kind), switch = "y") +
  scale_y_discrete(position = "right", limits = rev) +
  scale_x_reverse(labels = scales::label_number(scale_cut = cut_si("bp")), expand = expansion(mult = c(0.05, 0))) +
  theme(axis.text.y = element_blank()) +
  
  ggplot(DMR_numbers %>% filter(kind == "gain"), aes(x = length, y = comparison, fill = comparison, label = str_c(round(length/1000, 0), " kb"))) +
  facet_grid(rows = vars(context), cols = vars(kind)) +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(labels = scales::label_number(scale_cut = cut_si("bp")), expand = expansion(mult = c(0, 0.05))) +
  theme(axis.text.y = element_text(hjust = 0.5)) &
  
  geom_col() &
  geom_text(position = position_stack(vjust = 0.5)) &
  coord_fixed(ratio = 2.5e5) &
  scale_fill_manual(values = comparison_colors) &
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),legend.position= "none",
        axis.ticks.y = element_blank(), axis.title.y = element_blank(), axis.line = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA))) +
  
  plot_annotation(title = "Differentially Methylated Regions (DMRs). LENGTH",
                  subtitle = "CG: Noise filter method, mpd = 0.5
CHG: Bins method, mpd = 0.5
CHH: Bins method, mpd = 0.3") +
  plot_layout(widths = c(1.1, 1))

ggsave2("../figures/dmrs-2026-CHG-bins/colplot-DMRs-length-Figure-2B.pdf", width = 8, height = 5)
```

## Volcano plots {.tabset}

```{r}
# function to display y-axis values
reverselog_trans <- function(base = exp(1)) {
  trans <- function(x) -log(x, base)
  inv <- function(x) base^(-x)
  trans_new(paste0("reverselog-", format(base)), trans, inv, 
            log_breaks(base = base), 
            domain = c(1e-1000, Inf))
}

# mpd thresholds
vlines <- tibble(context = c("CG", "CHG", "CHH"),
                 threshold = c(0.5, 0.5, 0.3))

# function to plot volcano plots of DMRs
volcano_dmrs <- function(tb_dmrs, v_lines = vlines, pval_limit = 1e-30) {
  ggplot(tb_dmrs %>% filter(cytosinesCount >= 4), aes(x = (proportion2 - proportion1), y = pValue)) +
  facet_grid(cols = vars(comparison), rows = vars(context)) +
  geom_vline(data = v_lines, aes(xintercept = threshold),  color = "dodgerblue3") +
  geom_vline(data = v_lines, aes(xintercept = -threshold), color = "dodgerblue3") +
  geom_hline(yintercept = 0.01, color = "dodgerblue3") +
  geom_point(shape = 16, size = 0.3, alpha = 0.3) %>% rasterise(dpi = 600) +
  scale_x_continuous(expand = c(0.01, 0.01), breaks = breaks_width(0.4), labels = label_percent(), name = "%mC - WT %mC") +
  scale_y_continuous(trans = reverselog_trans(10), limits = c(1, pval_limit),
                     breaks = breaks_log(n = 6), expand = c(0.01, 0.01), 
                     labels = trans_format("log10", math_format(10^.x))) +
  labs(title = "Volcano plot of DMRs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),
        panel.border = element_rect(colour = "black", size = 1), axis.line = element_blank())
}
```

### No zoom
 
```{r fig.width=6, fig.height=7, warning=FALSE}
volcano_dmrs(tb_dmrs_filtered, pval_limit = NA)
```

### Zoom (p-Value > 10^-100)
 
```{r fig.width=6, fig.height=7, warning=FALSE}
volcano_dmrs(tb_dmrs_filtered, pval_limit = 1e-100)
```

### Zoom (p-Value > 10^-30)
 
```{r fig.width=6, fig.height=7, warning=FALSE}
volcano_dmrs(tb_dmrs_filtered, pval_limit = 1e-30)

ggsave(filename = "../figures/dmrs-2026-CHG-bins/volcano-DMRs-zoom-30-Figure-2C-.pdf", width = 6, height = 7)
```

## Heatmaps {.tabset}

Prepare data

```{r load-data-from-dmr-status}
load("../data/6_dmrs-2025/dmrs_whole_genome_A4_A17_CHG_relaxed-all-samples-data.Rdata")
```

```{r dmrs-make-tibble-with-methylation}
tb_dmrs_meth_all_samples <- dmr_status$a17$CHG %>% 
  as_tibble() %>%
  rename_with(~ str_remove(.x, str_c("_", "CHG"))) %>% 
  select(-starts_with(str_c("cytosinesCount", "CHG"))) %>% 
  filter(seqnames == "empty")

for (compar in c("a17", "a4")) {
  for (context in c("CHG")) {
    tb_dmrs_meth_all_samples <- bind_rows(tb_dmrs_meth_all_samples,
                                          dmr_status[[compar]][[context]] %>% 
                                            as_tibble() %>%
                                            rename_with(~ str_remove(.x, str_c("_", context))) %>% 
                                            select(-starts_with(str_c("cytosinesCount", context))) %>% 
                                            mutate(comparison = compar,
                                                   seqnames = str_remove(seqnames, "SL4.0"),
                                                   diff = proportion2 - proportion1))
  }
}
```

```{r filter-dmrs-with-methylation}
tb_dmrs_meth_filtered <- tb_dmrs_meth_all_samples %>%
  filter(pValue < 0.01,
         cytosinesCount >= 4) 
write_tsv(tb_dmrs_meth_filtered, "../data/6_dmrs-2025/table-dmrs-whole-genome-a17-a4-CHG-bins-all-samples-filtered.tsv")

tb_dmrs_meth_filtered_CG_CHH <- read_tsv("../data/6_dmrs/table-dmrs-whole-genome-a17-a4-all-samples-filtered.tsv", show_col_types = FALSE)
tb_dmrs_meth_filtered_CHG <- read_tsv("../data/6_dmrs-2025/table-dmrs-whole-genome-a17-a4-CHG-bins-all-samples-filtered.tsv", show_col_types = FALSE)

tb_dmrs_meth_filtered_combined <- bind_rows(tb_dmrs_meth_filtered_CG_CHH, tb_dmrs_meth_filtered_CHG)
write_tsv(tb_dmrs_meth_filtered_combined, "../data/6_dmrs-2025/table-dmrs-whole-genome-a17-a4-CHG-bins-combined-all-samples-filtered.tsv")
```

Load data

```{r}
tb_dmrs_meth_filtered_combined <- read_tsv("../data/6_dmrs-2025/table-dmrs-whole-genome-a17-a4-CHG-bins-combined-all-samples-filtered.tsv", show_col_types = FALSE)
```

```{r heatmap-functions}
# color ramp from a Brewer palette
brewer.ramp <- function(start, end, pal) {
  palette = RColorBrewer::brewer.pal(n = RColorBrewer::brewer.pal.info[pal, "maxcolors"],  name = pal)
  sequence = seq(start, end, (end - start)/(RColorBrewer::brewer.pal.info[pal, "maxcolors"] - 1))
  ramp = circlize::colorRamp2(sequence, palette)
  return(ramp)
}

# Function that takes a tibble of combined DMR comparisons and returns a long table
filter_dmrs_and_pivot <- function(tb_dmrs, comp, ctx) {
  tb_dmrs %>% 
    filter(comparison == comp, 
           context == ctx)  %>% 
    pivot_longer(cols = starts_with("proportion_"), values_to = "methylation", names_to = "sample", 
                 names_prefix = "proportion_", names_transform = toupper) %>% 
    mutate(index = str_c(seqnames, ":", start, ",", end),
           # add description
           description = case_when(sample == "WT" ~ "WT", 
                                   sample == "A4" ~ "TET3", 
                                   sample == "A17" ~ "NTS") %>% 
             fct_inorder()) %>% 
    # scale methylation
    group_by(index) %>% 
    mutate(scaled = scale(methylation)) %>% 
    ungroup()
}

#Function that takes a ComplexHeatmap list and makes it work with Patchwork syntax
wrap_heatmap_list <- function(heatmap_list, draw = FALSE) {
  heatmap_list %>% 
    {if (draw) ComplexHeatmap::draw(.) else .} %>%
    grid::grid.grabExpr() %>% 
    patchwork::wrap_elements()
}

# Function that takes a long tibble of dmr methylation and generates a list of two heatmaps, unscaled and scaled
heatmap_dmrs <- function(tb_dmrs_meth_long, show_legends = TRUE, title = character(0), max_n = 4000) {
  n_dmrs = unique(tb_dmrs_meth_long$index) %>% length()
  
  tb_dmrs_meth_long_sample <- tb_dmrs_meth_long %>% 
    filter(index %in% sample(unique(tb_dmrs_meth_long$index), size = max_n))
  
  (heatmap(tb_dmrs_meth_long_sample, .row = index, .column = description, .value = methylation, cluster_columns = TRUE, show_row_names = FALSE,
             palette_value = color_ramp_unscaled, name = "% mC",
             row_title = NULL, column_title = NULL) %>% 
     annotation_tile(description, palette = description_colors, show_annotation_name = FALSE) %>% 
     annotation_tile(diff, palette = color_ramp_diff, annotation_name_side = "top") + 
     heatmap(tb_dmrs_meth_long_sample, .row = index, .column = description, .value = scaled, cluster_columns = TRUE, show_row_names = FALSE, 
           palette_value = color_ramp_scaled, row_km = 2, name = "scaled",
           row_title = NULL, column_title = NULL) %>% 
     annotation_tile(description, palette = description_colors, show_annotation_name = FALSE)) %>% 
  draw(column_title = str_c(title, " (" , n_dmrs, " DMRs)"), merge_legend = TRUE, 
       show_heatmap_legend = show_legends, show_annotation_legend = show_legends) %>% 
  wrap_heatmap_list(draw = FALSE)
}

heatmap_dmrs_split <- function(tb_dmrs_meth_long, show_legends = TRUE, title = character(0), max_n = 4000, 
                         top_tile_variable = description, top_tile_colors = description_colors, subset_trick = FALSE,
                         use_raster = TRUE) {
  
  n_dmrs = unique(tb_dmrs_meth_long$index) %>% length()
  
  if (subset_trick == TRUE) {
    # remove some DMRs, as if not, the diff tile doesn't match the heatmap
    tb_dmrs_meth_long <- tb_dmrs_meth_long %>% 
      filter(seqnames %in% str_c("chr", 1:9))
  }
  
  final_n_dmrs = unique(tb_dmrs_meth_long$index) %>% length()
  if (final_n_dmrs < max_n) max_n <- final_n_dmrs
  
  tb_dmrs_meth_long_sample <- tb_dmrs_meth_long %>% 
    filter(index %in% sample(unique(tb_dmrs_meth_long$index), size = max_n)) %>% 
    group_by(regionType) %>% 
    mutate(genotype = {{ top_tile_variable }})
  
  (heatmap(tb_dmrs_meth_long_sample, .row = index, .column = description, .value = methylation, 
           cluster_columns = TRUE, show_row_names = FALSE, use_raster = use_raster, 
           palette_value = color_ramp_unscaled, name = "mC", palette_grouping = list(c("firebrick3", "dodgerblue3")), 
           row_title = NULL, column_title = NULL) %>%
      annotation_tile(genotype, palette = top_tile_colors, show_annotation_name = FALSE) %>% 
      annotation_tile(diff, palette = color_ramp_diff, annotation_name_side = "top") + 
      heatmap(tb_dmrs_meth_long_sample, .row = index, .column = description, .value = scaled, 
              cluster_columns = TRUE, show_row_names = FALSE, use_raster = use_raster, 
              palette_value = color_ramp_scaled, row_km = 2, name = "scaled", palette_grouping = list(c("firebrick3", "dodgerblue3"), use_raster = TRUE), 
              row_title = NULL, column_title = NULL) %>% 
      annotation_tile(genotype, palette = top_tile_colors, show_annotation_name = FALSE)) %>% 
    draw(column_title = str_c(title, " (" , n_dmrs, " DMRs)"), merge_legend = TRUE, 
       show_heatmap_legend = show_legends, show_annotation_legend = show_legends) %>% 
    wrap_heatmap_list(draw = FALSE)
}

heatmap_dmrs_split_no_scaled <- function(tb_dmrs_meth_long, show_legends = TRUE, title = character(0), max_n = 4000, 
                         top_tile_variable = description, top_tile_colors = description_colors, subset_trick = FALSE,
                         use_raster = FALSE) {
  
  n_dmrs = unique(tb_dmrs_meth_long$index) %>% length()
  
  if (subset_trick == TRUE) {
    # remove some DMRs, as if not, the diff tile doesn't match the heatmap
    tb_dmrs_meth_long <- tb_dmrs_meth_long %>% 
      filter(seqnames %in% str_c("chr", 1:9))
  }
  
  final_n_dmrs = unique(tb_dmrs_meth_long$index) %>% length()
  if (final_n_dmrs < max_n) max_n <- final_n_dmrs
  
  tb_dmrs_meth_long_sample <- tb_dmrs_meth_long %>% 
    filter(index %in% sample(unique(tb_dmrs_meth_long$index), size = max_n)) %>% 
    group_by(regionType) %>% 
    mutate(genotype = {{ top_tile_variable }})
  
  (heatmap(tb_dmrs_meth_long_sample, .row = index, .column = description, .value = methylation, 
           cluster_columns = TRUE, show_row_names = FALSE, use_raster = use_raster, 
           palette_value = color_ramp_unscaled, name = "mC", palette_grouping = list(c("firebrick3", "dodgerblue3")), 
           row_title = NULL, column_title = NULL, width = unit(3, "cm")) %>%
      annotation_tile(genotype, palette = top_tile_colors, show_annotation_name = FALSE) %>% 
      annotation_tile(diff, palette = color_ramp_diff, annotation_name_side = "top", size = unit(1, "cm"))) %>% # + 
      #heatmap(tb_dmrs_meth_long_sample, .row = index, .column = description, .value = scaled, 
      #        cluster_columns = TRUE, show_row_names = FALSE, use_raster = use_raster, 
      #        palette_value = color_ramp_scaled, row_km = 2, name = "scaled", palette_grouping = list(c("firebrick3", "dodgerblue3"), use_raster = TRUE), 
      #        row_title = NULL, column_title = NULL) %>% 
      #annotation_tile(genotype, palette = top_tile_colors, show_annotation_name = FALSE)) %>% 
    as_ComplexHeatmap() %>% 
    draw(column_title = str_c(title, " (" , n_dmrs, " DMRs)"), merge_legend = TRUE, 
       show_heatmap_legend = show_legends, show_annotation_legend = show_legends) %>% 
    wrap_heatmap_list(draw = FALSE)
}
```

```{r colors-for-heatmaps}
color_ramp_diff  <- brewer.ramp(1, -1, "Spectral")
color_ramp_scaled   <- brewer.ramp(2, -2, "RdBu")
color_ramp_unscaled <- brewer.ramp(0, 1, "YlOrRd")
description_colors <- c(WT  = "grey40", "NTS" = "slateblue2", "TET3" = "tomato2") %>% fct_inorder() # they have to be in this order, for some reason
```

### TET3

```{r fig.width=12, fig.height=7}
tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a4", ctx = "CG") %>% 
  heatmap_dmrs_split(title = "TET3 CG", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a4", ctx = "CHG") %>% 
  heatmap_dmrs_split(title = "TET3 CHG", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a4", ctx = "CHH") %>% 
  heatmap_dmrs_split(title = "TET3 CHH") +
  
  plot_layout(nrow = 1, widths = c(4, 4, 5)) +
  plot_annotation(title = "TET3 positive plant (TET3 vs WT)")

ggsave2("../figures/dmrs-2026-CHG-bins/heatmaps-DMRs-TET3.pdf", width = 12, height = 7)
```

### NTS

```{r fig.width=12, fig.height=7}
tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a17", ctx = "CG") %>% 
  heatmap_dmrs_split(title = "NTS CG", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a17", ctx = "CHG") %>% 
  heatmap_dmrs_split(title = "NTS CHG", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a17", ctx = "CHH") %>% 
  heatmap_dmrs_split(title = "NTS CHH") +
  
  plot_layout(nrow = 1, widths = c(4, 4, 5)) +
  plot_annotation(title = "TET3 negative plant (NTS vs WT)")

ggsave2("../figures/dmrs-2026-CHG-bins/heatmaps-DMRs-NTS.pdf", width = 12, height = 7)
```

### CG

```{r fig.width=6, fig.height=5}
tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a4", ctx = "CG") %>% 
  heatmap_dmrs_split_no_scaled(title = "TET3 CG\n", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a17", ctx = "CG") %>% 
  heatmap_dmrs_split_no_scaled(title = "NTS CG\n") +

  plot_layout(nrow = 1, widths = c(8.5, 11))

ggsave2("../figures/dmrs-2026-CHG-bins/heatmaps-DMRs-CG.pdf", width = 6, height = 5)
```

### CHG

```{r fig.width=6, fig.height=5}
tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a4", ctx = "CHG") %>% 
  heatmap_dmrs_split_no_scaled(title = "TET3 CHG\n", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a17", ctx = "CHG") %>% 
  heatmap_dmrs_split_no_scaled(title = "NTS CHG\n") +

  plot_layout(nrow = 1, widths = c(8.5, 11))

ggsave2("../figures/dmrs-2026-CHG-bins/heatmaps-DMRs-CHG_Figure-2S3.pdf", width = 6, height = 5)
```


### CHH

```{r fig.width=6, fig.height=5}
tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a4", ctx = "CHH") %>% 
  heatmap_dmrs_split_no_scaled(title = "TET3 CHH\n", show_legends = FALSE) +
  
  tb_dmrs_meth_filtered_combined %>% 
  filter_dmrs_and_pivot(comp = "a17", ctx = "CHH") %>% 
  heatmap_dmrs_split_no_scaled(title = "NTS CHH\n") +

  plot_layout(nrow = 1, widths = c(8.5, 11))

ggsave2("../figures/dmrs-2026-CHG-bins/heatmaps-DMRs-CHH.pdf", width = 6, height = 5)
```

## Genome plots DMR {.tabset}

```{r genome-plots-functions}
# Function to plot chr as facets in ggplot
facet_chr <- function(chr = chr, rows = NULL, scales = "free_x", space = "free_x", breaks = 40e6, switch = NULL) {
  list(facet_grid(rows = vars({{ rows }}), cols = vars({{ chr }}), scales = scales, space = space, switch = switch),
       scale_x_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(breaks), 
                          labels = label_number(scale_cut = cut_short_scale())),
       xlab("Genomic coordinates (bp)"))
}

## Data and Function to create heterochromatin ribbon
heterochromatin_H3K27ac = read_tsv("/path/to/genome/SL4.0_heterochromatin_H3K27ac_chXX.bed", col_names = c("chr","start","end"), col_types = cols())

SL4_chrs <- str_c("ch", str_pad(c(1:8, 10:12), 2, pad = "0"))

geom_heterochromatin <- function(hchrom = heterochromatin_H3K27ac, chr_col = chr, chrs = SL4_chrs) {
  hchrom <- dplyr::rename(hchrom, {{ chr_col }} := chr)
  list(geom_rect(data = hchrom %>% filter(chr %in% chrs),
                 aes(x = NULL, y = NULL, xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
                 inherit.aes = FALSE, color = NA, fill = "black", alpha = 0.1))
}
```

### Distribution across genome 

```{r fig.width=12, fig.height=10}
# DMR density across the genome
tb_dmrs_filtered %>% 
  filter(chr %in% SL4_chrs) %>% 
  ggplot(aes(start, color = comparison, fill = comparison)) +
  facet_chr(chr = chr, rows = str_c(context, "\n", regionType, "\n", comparison), , scales = "free", space = "free_x") +
  geom_heterochromatin() +
  geom_area(aes(y = after_stat(count)), stat = "bin", alpha = 0.2, binwidth = 3e6) +
  scale_color_manual("comparison", values = comparison_colors) +
  scale_fill_manual("comparison", values = comparison_colors) +
  labs(title = "DMR distibution across the genome", 
       subtitle = "Note that different chromosomes are affected for TET3 and NTS (mainly for CG loss)",
       y = "DMRs per 300 Mbp") +
  theme(panel.spacing.x=unit(0.5, "lines"),
        strip.background.y = element_rect(colour = NA, fill = "papayawhip"), strip.background.x = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
        legend.position = "bottom", axis.title.x = element_blank())

ggsave("../figures/dmrs-2026-CHG-bins/genome-plot-DMRs-distribution.pdf", width = 12, height = 10)
```

### Difference across genome 

```{r fig.width=12, fig.height=8, warning=FALSE}
tb_dmrs_filtered %>% 
  filter(chr %in% SL4_chrs) %>% 
  mutate(mC_diff = proportion2 - proportion1) %>% 
  ggplot(aes(x = start, y = mC_diff, color = comparison, partition = comparison)) +
  #facet_chr(chr = seqnames, rows = (context, regionType), scales = "free", space = "free_x", breaks = 20e6) +
  facet_grid(rows = vars(context, regionType), cols = vars(chr), scales = "free", space = "free") +
  geom_heterochromatin() +
  #geom_point(shape = 16, size = 0.5, alpha = 0.6) |> partition(vars(description)) |> blend("multiply") +
  #geom_point(shape = 16, size = 0.6, alpha = 0.4) * (blend("darken") + blend("multiply", alpha = 0.5)) +
  geom_point(shape = 16, size = 0.6, alpha = 0.4) %>% rasterise(dpi = 300) * 
     (blend("darken") + blend("multiply", alpha = 0.5)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(20e6), 
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(expand = c(0, 0), breaks = breaks_width(0.2), labels = label_percent(), name = "%mC - WT %mC") +
  scale_color_manual("comparison", values = comparison_colors)  +
  labs(title = "DMR methylation difference across the genome") +
  xlab("Genomic coordinates (bp)") +
  theme(panel.spacing.x=unit(0.5, "lines"),
        strip.background.y = element_rect(colour = NA, fill = "papayawhip"), strip.background.x = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(),
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
        legend.position = "bottom", axis.title.x = element_blank())

ggsave("../figures/dmrs-2026-CHG-bins/genome-plot-DMRs-diff-blended-Figure-3A.pdf", width = 12, height = 8, dpi = 600, bg = "white", device = cairo_pdf)
```

## Correlation DMRs  Genes {.tabset}

```{r read-genes-RDS}
genes = readRDS("/path/to/genome/v4_Mt_Pt_TET3/ITAG4.1_gene_models_Mt_Pt_TET3-named.RDS")
```

### Scatter plot

```{r fig.width=10, fig.height=10}
# scatter plot of density
si_number <- function(x, digits = 2) {
  compress <- function(x, n) {
    signif(x * 10^(-n), digits)
  }
  case_when(x >= 1e6   ~ paste0(compress(x, 6), "M"),
            x >= 1000  ~ paste0(compress(x, 3), "k"),
            x >= 1     ~ as.character(compress(x, 0)),
            x >= 0.001 ~ paste0(compress(x, -3), "m"),
            x >= 1e-6  ~ paste0(compress(x, -6), "u"))
}

scatter_density <- function(bin_width = 1e6) {
  binned_genes <- genes %>%
    as_tibble() %>%
    group_by(seqnames, bins = cut_width(start, width = bin_width, boundary = 0)) %>%
    summarise(n_genes = n(), .groups = "drop") %>%
    mutate(#seqnames = str_remove(seqnames, "SL4.0"),
           n_genes = 100 * jitter(n_genes, amount=.5) / length(genes))
  
  binned_dmrs <- tb_dmrs_filtered %>%
    #filter(context != "CHG", regionType == "gain") %>% 
    group_by(context, regionType, comparison) %>%
    mutate(total = n()) %>% 
    group_by(context, seqnames, regionType, comparison, bins = cut_width(start, width = bin_width, boundary = 0), total) %>%
    summarise(dmrs = n(), .groups = "drop") %>%
    mutate(dmrs = 100 * jitter(dmrs, amount=.5) / total)
  
  binned_joined <- full_join(binned_genes, binned_dmrs,  by = c("seqnames", "bins")) %>%
    na.omit() 
    
  ggplot(binned_joined, aes(x = n_genes, y = dmrs)) +
    facet_grid(cols = vars(context), rows = vars(regionType, comparison)) +
    geom_smooth(method = "lm", se=FALSE, formula = y ~ x, linewidth = 0.5) +
    geom_point(shape = 16, size = 1, alpha = 0.4) +
    #geom_jitter(shape = 16, size = 1, alpha = 0.4, width = 0.00005, height = 0.0005) +
    scale_x_continuous(name = str_c("% of genes per ", si_number(bin_width), "bp"), expand = c(0, 0), limits = c(0, NA)) +
    scale_y_continuous(name = str_c("% of DMRs per ", si_number(bin_width), "bp"),  expand = c(0, 0), limits = c(0, 1.5)) +
    ggpubr::stat_regline_equation(label.y = 1.35,#0.9 * max(binned_joined$dmrs), 
                                  label.x = 0.1 * max(binned_joined$n_genes)) +
    ggpubr::stat_cor(label.y = 1.2,#0.8 * max(binned_joined$dmrs), 
                     label.x = 0.1 * max(binned_joined$n_genes)) +
    theme(axis.line = element_blank(), panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          strip.background = element_rect(fill = "papayawhip"))
}

scatter_density(2e6) +
  ggtitle("Correlation of density of DMRs and Genes")
ggsave(filename = "../figures/dmrs-2026-CHG-bins/correlation-DMRs-genes-scatter-2-Mbp_Figure-3S2.pdf", width = 10, height = 10)
```

### Heatmap

```{r}
# color ramp from a Brewer palette
brewer.ramp <- function(start, end, pal) {
  palette = RColorBrewer::brewer.pal(n = RColorBrewer::brewer.pal.info[pal, "maxcolors"],  name = pal)
  sequence = seq(start, end, (end - start)/(RColorBrewer::brewer.pal.info[pal, "maxcolors"] - 1))
  ramp = circlize::colorRamp2(sequence, palette)
  return(ramp)
}
```

```{r}
bin_width = 2e6

binned_genes <- genes %>%
  as_tibble() %>%
  group_by(seqnames, bins = cut_width(start, width = bin_width, boundary = 0)) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
  mutate(n_genes = 100 * n_genes / length(genes))

binned_dmrs <- tb_dmrs_filtered %>%
  #filter(context != "CHG", regionType == "gain") %>% 
  group_by(context, regionType, comparison) %>%
  mutate(total = n()) %>% 
  group_by(context, seqnames, regionType, comparison, bins = cut_width(start, width = bin_width, boundary = 0), total) %>%
  summarise(dmrs = n(), .groups = "drop") %>%
  mutate(dmrs = 100 * dmrs / total)
  
binned_joined <- full_join(binned_genes, binned_dmrs,  by = c("seqnames", "bins")) %>%
  na.omit() 
  
# Calculate correlation coefficients for each combination
correlation_data <- binned_joined %>%
  group_by(context, regionType, comparison) %>%
  #filter(n() > 10) %>% 
  summarize(cor = cor(n_genes, dmrs, use = "complete.obs"), .groups = "drop")

# Spread the data for heatmap plotting
heatmap_data <- correlation_data %>%
  pivot_wider(names_from = c(regionType, comparison), values_from = cor) %>% 
  column_to_rownames("context")
```

```{r, fig.width=3.5, fig.height=3}
library(shadowtext)

mat = as.matrix(heatmap_data)

# make my own heatmap function
hms = Heatmap(mat, col =  brewer.ramp(1, -1, "RdYlBu"), name = "R ", rect_gp = gpar(col = "black", lwd = 2),
              column_title = "R coefficient", cluster_columns = FALSE, cluster_rows = FALSE,
              cell_fun = function(j, i, x, y, width, height, fill) {
                #grid.shadowtext(sprintf(sprintf, mat2[i, j]), x, y, gp = gpar(fontsize = 13))
                grid.shadowtext(sprintf("%.2f", mat[i, j]), x, y, gp = gpar(fontsize = 12 , col = "white"), bg.colour = "black")
              }
)


# draw the heatmaps and save them
draw(hms, column_title = gt_render("**Correlations DMRs vs genes**"))
dev.copy2pdf(file = "../figures/dmrs-2026-CHG-bins/correlation-DMRs-genes-heatmap-2-Mbp_Figure-3B--.pdf", width = 3.5, height = 3) %>% invisible()
```

## Distance to genes distribution  {.tabset}

### count & ECDF

```{r fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
p0 <- tb_dmrs_filtered %>% 
  as_granges() %>% 
  add_nearest_distance(genes) %>% 
  as_tibble() %>% 
  ggplot(aes(distance, fill = context, color = context)) + 
  facet_grid(rows = vars(regionType), cols = vars(comparison), scales = "free", space = "free") +
  geom_vline(xintercept = 3000, color = "red3") +  
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), 
        panel.border = element_rect(colour = "black", linewidth = 1), axis.line = element_blank(), )

p_log <- p0 %+% 
  transform(tb_dmrs_filtered %>% as_granges() %>% add_nearest_distance(genes) %>% as_tibble(), distance = distance + 1) + 
  scale_x_continuous("distance to closest gene", trans = log10_trans(), breaks = breaks_log(n = 6), labels = label_number(scale_cut = cut_short_scale()))

p_lin <- p0 + 
  scale_x_continuous("distance to closest gene") + 
  coord_cartesian(xlim = c(0, 1e4))

(p_log + 
    geom_area(aes(y = after_stat(count)), stat = "bin", alpha = 0.2, binwidth = 0.05, position = "identity") + 
    scale_x_continuous(trans = log10_trans(), breaks = breaks_log(n = 6), labels = label_number(scale_cut = cut_short_scale()), limits = c(2, NA))) +
  
  (p_lin + 
     geom_area(aes(y = after_stat(count)), stat = "bin", alpha = 0.2, binwidth = 50, position = "identity") + 
     scale_x_continuous(limits = c(1, 1e4))) +
  
  (p_log + 
     stat_ecdf(pad = FALSE) + 
     scale_y_continuous("cumulative DMRs", labels = label_percent())) +
  
  (p_lin + 
     stat_ecdf(pad = FALSE) + 
     scale_y_continuous("cumulative DMRs", labels = label_percent())) +
  
  plot_layout(guides = "collect") &
  theme(legend.position='bottom')

ggsave("../figures/dmrs-2026-CHG-bins/distribution-gene-distance-to-dmrs.pdf", width = 10, height = 8)
```

### Freqpoly for figure

```{r fig.width=5.5, fig.height=5, warning=FALSE}
tb_dmrs_filtered %>% 
  #mutate(seqnames = str_c("SL4.0", seqnames)) %>% 
  group_by(context, regionType, comparison) %>%
  filter(n() > 10) %>% 
  ungroup() %>% 
  as_granges() %>% 
  add_nearest_distance(genes) %>% 
  mutate(distance = distance + 1) %>% 
  as_tibble() %>%
  ggplot(aes(distance, fill = context, color = context)) +
  facet_grid(rows = vars(regionType), cols = vars(comparison), scales = "free", space = "free") + # cols = vars(context), 
  geom_area(aes(y = after_stat(density)), stat = "bin", alpha = 0.2, binwidth = 0.05, position = "identity") +
  #geom_density(adjust = 0.5, alpha = 0.3) +
  #geom_freqpoly(binwidth = 0.05) +
  scale_x_continuous("distance to closest gene", trans = log10_trans(), breaks = breaks_log(n = 6), labels = label_number(scale_cut = cut_short_scale()),
                     limits = c(2, NA)) +
  theme_cowplot() +
  labs(title = "Distance to nearest gene from DMRs", 
       subtitle = "Loss CG & CHG DMRs are farther from genes") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), panel.border = element_rect(colour = "black", linewidth = 1), 
        axis.line = element_blank(), legend.position = "bottom")

ggsave("../figures/dmrs-2026-CHG-bins/distribution-gene-distance-to-dmrs-density_Figure-3S2.pdf", width = 5.5, height = 5)
```


## Metagene plot of DMR location

```{r}
source("/path/to/project/bin/R/enriched_heatmaps_functions.R")
```

```{r}
test <- tb_dmrs_filtered %>% 
  filter(comparison == "NTS", 
         context == "CHG", 
         regionType == "loss") %>% 
  as_granges() %>% 
  filter_by_overlaps(genes, maxgap = 2000)
```

```{r}
meta_tb <- tibble()
for (comp in c("TET3", "NTS")) {
  for (ctx in c("CG", "CHG", "CHH")) {
    for (kind in c("gain", "loss")) {
      dmrs <- read_gff3(str_c("../data/6_dmrs-2025/gffs/DMRs_", comp, "_", ctx, "_", kind, ".gff3"))
      
      mat <- normalise_to_matrix(signal = dmrs, target = filter_by_overlaps(genes, dmrs, maxgap = 2000), 
                                 bed = TRUE, target_ratio = 1/3) 
      meta_tb <- mat[] %>% 
        as_tibble() %>% 
        replace(is.na(.), 0) %>%
        pivot_longer(everything(), names_to = "position", values_to = "value") %>%
        mutate(position = factor(position, levels = unique(position))) %>%
        group_by(position) %>%
        summarise(mean   =   mean(value, na.rm = T), .groups = "drop") %>%
        mutate(r_mean   = slide_dbl(mean,   ~mean(.x,   na.rm = TRUE), .before = 2, .after = 2), 
               comparison = comp, 
               context = ctx, 
               kind = kind) %>% 
        bind_rows(meta_tb)
      
      
    }
  }
}
```

```{r fig.width=7, fig.height=5}
sample_colors <- c(WT  = "grey40", TET3 = "tomato2", NTS = "slateblue2")

ggplot(meta_tb, aes(x = position, y = mean, color = comparison, linetype = kind, group = paste(comparison, kind))) +
  facet_grid(rows = vars(context), scales = "free_y") +
  #geom_point(pch=16, size=1, alpha = 0.6) +
  geom_line(mapping = aes(y = r_mean), linewidth = 0.5, alpha=1) +
  scale_x_discrete(breaks = c("u1","t1","d1","d50"), labels = c(paste0("-",2," Kb"), "Start", "End", paste0("+",2," Kb"))) +
  scale_y_continuous("genes with DMRs", labels = scales::label_percent(), limits = c(0, NA)) +
  scale_color_manual(values = sample_colors) +
  annotate("segment", x=-Inf, xend=-Inf, y=-Inf, yend=Inf, linewidth = 1) + # y axis on all facets
  annotate("segment", x=-Inf, xend=Inf, y=-Inf, yend=-Inf, linewidth = 1) + # y axis on all facets
  annotate("rect", xmin="t1", xmax="d1", ymin=-Inf, ymax=Inf, alpha=0.1, fill="black") +
  theme(axis.line=element_blank(), axis.title.x = element_blank(),
        strip.background.y = element_rect(colour = NA, fill = "papayawhip"), strip.background.x = element_blank(),
        panel.spacing=unit(2, "lines"), strip.background=element_blank()) +
  labs(title = "Distribution of DMRs across genes")

ggsave2("../figures/dmrs-2026-CHG-bins/metaplot-DMRs-genes_Figure-3C.pdf", width = 150, height = 150, units = "mm")
```


## Annotation DMRs to genes and TEs

```{r annotation functions}
read_chrom.sizes <- function(file) {
  read_tsv(file, col_names = c("seqnames", "end"), col_types = cols()) %>%
    mutate(start = 1) %>%
    as_granges()
}

extract_intersect <- function(query, subject, name = deparse(substitute(subject))) {
  # query needs to be a list of granges with an element called remainder
  result <- query
  
  result[[name]] <- query$remainder %>%
    intersect_ranges(subject) %>%
    { if (length(.) > 0) mutate(., feature = name) else . }
  
  result$remainder <- query$remainder %>%
    setdiff_ranges(subject)
  
  return(result)
}

annotate_features_list <- function(regions, list_of_features = tomato_v4, name = deparse(substitute(regions))) {
  result <- list(remainder = regions)
  # intersect hierarchically
  for (feature in names(list_of_features)) {
    result <- result %>%
      extract_intersect(list_of_features[[feature]],  name = feature)
  }
  # summarise to table
  result %>%
    bind_ranges() %>%
    group_by(feature) %>%
    summarise(bases = sum(width)) %>%
    as_tibble() %>%
    mutate(fraction = 1*bases/sum(bases),
           feature = factor(feature, levels = names(list_of_features)),
           region = name) %>%
    arrange(feature)
}

feature_distribution_plot <- function(annotations, x = fraction, type = deparse(substitute(x)), 
                                      genome_name = "Tomato Genome", palette = "YlGnBu") {
  if (type == "fraction") {
    ggplot(annotations, aes(y = as_factor(region), x = {{ x }}, fill = as_factor(feature))) +
      geom_col(color = "black", linewidth = 0.5) +
      scale_fill_brewer(palette = palette, name = "Feature", direction = -1) +
      scale_x_continuous(expand = c(0,0), labels = scales::label_percent()) +
      scale_y_discrete(position = "right", limits = rev) +
      theme_cowplot() +
      theme(axis.title = element_blank(), axis.line.y  = element_blank(), axis.ticks.y = element_blank()) 
  } else if (type == "bases") {
    annotations %>%
      filter(region != genome_name) %>% add_row(feature = "genes", bases = 0, fraction = 0, region = genome_name, .before = 1) %>%
      ggplot(aes(y = as_factor(region), x = bases, fill = as_factor(feature))) +
      geom_col(color = "black", linewidth = 0.5) + 
      scale_fill_brewer(palette = palette, name = "Feature", direction = -1) +
      scale_x_continuous(expand = c(0,0), labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
      scale_y_discrete(limits = rev) +
      theme_cowplot() +
      theme(axis.line.y  = element_blank(), axis.ticks.y = element_blank(), axis.title.y = element_blank())
  }
}
```

```{r read-genomic-data}
tomato_chromosomes <- str_c("SL4.0ch", str_pad(0:12, width = 2, pad = 0))

genes <- read_rds("/path/to/genome/v4_Mt_Pt_TET3/ITAG4.1_gene_models_Mt_Pt_TET3-named.RDS")

repeats_light <- read_gff("/path/to/genome/ITAG4.0/TEtranscripts/solLyc_ITAG4.0_RepeatModeler_repeats_light_TE.gtf")  

tomato_chrs <- read_chrom.sizes("/path/to/genome/v4_Mt_Pt_TET3/S_lycopersicum_chromosomes.4.00_Mt_Pt_TET3.chrom.sizes") %>% 
  filter(seqnames %in% tomato_chromosomes)
```

```{r distribution-without-TEs}
tomato_v4 <- list(
  genes = genes,
  up    = genes %>% flank_upstream(width = 2000),
  down  = genes %>% flank_downstream(width = 2000),
  intergenic = tomato_chrs 
)

annotate_features_list(regions = tomato_chrs, list_of_features = tomato_v4, name = "Tomato Genome")
```

```{r distribution-with-TEs}
tomato_v4_repeats <- list(
  DNA_TEs = repeats_light %>% filter(class_id == "DNA"),
  #Helitrons_TEs = repeats_light %>% filter(class_id == "RC"),
  #LTR_TEs = repeats_light %>% filter(class_id == "LTR"),
  Gypsy_TEs = repeats_light %>% filter(family_id == "Gypsy"),
  Copia_TEs = repeats_light %>% filter(family_id == "Copia"),
  #LINE_TEs = repeats_light %>% filter(class_id == "LINE"),
  #SINE_TEs = repeats_light %>% filter(class_id == "SINE"),
  #Retroposon_TEs = repeats_light %>% filter(class_id == "Retroposon"),
  #Satellite_TEs = repeats_light %>% filter(class_id == "Satellite"),
  #Unknown_TEs = repeats_light %>% filter(class_id == "Unknown"),
  Other_TEs = repeats_light,
  genes = genes,
  up    = genes %>% flank_upstream(width = 2000),
  down  = genes %>% flank_downstream(width = 2000),
  intergenic = tomato_chrs 
)

annotate_features_list(regions = tomato_chrs, list_of_features = tomato_v4_repeats, name = "Tomato Genome") %>% 
  arrange(-fraction)
```
```{r annotate-dmrs-repeats}
dmrs_list_gr <- read_rds("../data/6_dmrs-2025/DMR-list_TET3_NTS_CHG-combined_GRanges.rds")

annotations_repeats <-  bind_rows(
  annotate_features_list(tomato_chrs,                list_of_features = tomato_v4_repeats, name = "Tomato Genome"),
  annotate_features_list(dmrs_list_gr$TET3$CG$gain,  list_of_features = tomato_v4_repeats, name = "CG TET3 gain"),
  annotate_features_list(dmrs_list_gr$NTS$CG$gain,   list_of_features = tomato_v4_repeats, name = "CG NTS gain"),
  annotate_features_list(dmrs_list_gr$TET3$CG$loss,  list_of_features = tomato_v4_repeats, name = "CG TET3 loss"),
  annotate_features_list(dmrs_list_gr$NTS$CG$loss,   list_of_features = tomato_v4_repeats, name = "CG NTS loss"),
  annotate_features_list(dmrs_list_gr$TET3$CHG$gain, list_of_features = tomato_v4_repeats, name = "CHG TET3 gain"),
  annotate_features_list(dmrs_list_gr$NTS$CHG$gain,  list_of_features = tomato_v4_repeats, name = "CHG NTS gain"),
  annotate_features_list(dmrs_list_gr$TET3$CHG$loss, list_of_features = tomato_v4_repeats, name = "CHG TET3 loss"),
  annotate_features_list(dmrs_list_gr$NTS$CHG$loss,  list_of_features = tomato_v4_repeats, name = "CHG NTS loss"),
  annotate_features_list(dmrs_list_gr$TET3$CHH$gain, list_of_features = tomato_v4_repeats, name = "CHH TET3 gain"),
  annotate_features_list(dmrs_list_gr$NTS$CHH$gain,  list_of_features = tomato_v4_repeats, name = "CHH NTS gain"),
  annotate_features_list(dmrs_list_gr$TET3$CHH$loss, list_of_features = tomato_v4_repeats, name = "CHH TET3 loss"),
  annotate_features_list(dmrs_list_gr$NTS$CHH$loss,  list_of_features = tomato_v4_repeats, name = "CHH NTS loss")
) %>%
  mutate(region = fct_inorder(region))
```

Near 2kb to genes

```{r annotate-dmrs-2kb-repeats}
dmrs_list_gr_near2kb <- dmrs_list_gr %>% 
  map_depth(3, ~ .x %>% 
              add_nearest_distance(genes) %>% 
              subset(distance <= 2000))

annotations_repeats_2kb <- bind_rows(
  annotate_features_list(genes %>% stretch(4000),            list_of_features = tomato_v4_repeats, name = "Genes + 2kb"),
  annotate_features_list(dmrs_list_gr_near2kb$TET3$CG$gain,  list_of_features = tomato_v4_repeats, name = "CG TET3 gain"),
  annotate_features_list(dmrs_list_gr_near2kb$NTS$CG$gain,   list_of_features = tomato_v4_repeats, name = "CG NTS gain"),
  annotate_features_list(dmrs_list_gr_near2kb$TET3$CG$loss,  list_of_features = tomato_v4_repeats, name = "CG TET3 loss"),
  annotate_features_list(dmrs_list_gr_near2kb$NTS$CG$loss,   list_of_features = tomato_v4_repeats, name = "CG NTS loss"),
  annotate_features_list(dmrs_list_gr_near2kb$TET3$CHG$gain, list_of_features = tomato_v4_repeats, name = "CHG TET3 gain"),
  annotate_features_list(dmrs_list_gr_near2kb$NTS$CHG$gain,  list_of_features = tomato_v4_repeats, name = "CHG NTS gain"),
  annotate_features_list(dmrs_list_gr_near2kb$TET3$CHG$loss, list_of_features = tomato_v4_repeats, name = "CHG TET3 loss"),
  annotate_features_list(dmrs_list_gr_near2kb$NTS$CHG$loss,  list_of_features = tomato_v4_repeats, name = "CHG NTS loss"),
  annotate_features_list(dmrs_list_gr_near2kb$TET3$CHH$gain, list_of_features = tomato_v4_repeats, name = "CHH TET3 gain"),
  annotate_features_list(dmrs_list_gr_near2kb$NTS$CHH$gain,  list_of_features = tomato_v4_repeats, name = "CHH NTS gain"),
  annotate_features_list(dmrs_list_gr_near2kb$TET3$CHH$loss, list_of_features = tomato_v4_repeats, name = "CHH TET3 loss"),
  annotate_features_list(dmrs_list_gr_near2kb$NTS$CHH$loss,  list_of_features = tomato_v4_repeats, name = "CHH NTS loss")
) %>%
  mutate(region = fct_inorder(region))
```

```{r fig.width=7, fig.height=4, message=FALSE}
(feature_distribution_plot(annotations_repeats, fraction, palette = "RdYlBu") +
   scale_fill_manual("Feature", values = c(genes = "#225EA8", up = "#41B6C4", down = "#A1DAB4", intergenic = "grey", 
                                           DNA_TEs = "#E31A1C", Gypsy_TEs = "#FD8D3C", Copia_TEs = "#FECC5C", Other_TEs = "#FFFFB2")) +
   ggtitle("All DMRs") |
   
   feature_distribution_plot(annotations_repeats_2kb, fraction, palette = "RdYlBu") +
   scale_fill_manual("Feature", values = c(genes = "#225EA8", up = "#41B6C4", down = "#A1DAB4", intergenic = "grey", 
                                           DNA_TEs = "#E31A1C", Gypsy_TEs = "#FD8D3C", Copia_TEs = "#FECC5C", Other_TEs = "#FFFFB2")) +
   ggtitle("DMRs near genes") +
   theme(axis.text.y = element_blank())) +
  plot_layout(guides = 'collect')

ggsave2("../figures/dmrs-2026-CHG-bins/DMRs-location_Figure-3D.pdf", width = 7, height = 4)
```
