---
title: "Scatter plots and PCAs of counts"
subtitle: "_Tomato TET3 RNA-seq_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

```{r libraries-and-functions, include=FALSE}
library(DESeq2)
library(tidyverse)
library(magrittr)
library(cowplot); theme_set(theme_cowplot())
library(ggforce)
library(patchwork)
library(ggbeeswarm)
library(ggrastr)

tpms_from_dds <- function(dds, gene_lengths) {
  counts(dds, normalized = F) %>%
  as_tibble(rownames = "id") %>%
  inner_join(read_tsv(gene_lengths, col_types = cols()), 
             by="id") %>%
  mutate(across(-c(id, length), ~ (.x / length)),       # rpk = count / length
         across(-c(id, length), ~ 1e+6 * .x / sum(.x)), # tpm = 1M * rpk / sum(rpk)
         across(-c(id, length), ~ log2(1 + .x))) %>%    # log2tpm = log2(1 + tpm) %>%
  dplyr::select(-length)
}
```

## Load counts and run DESeq2

Create sample table
```{r}
(sampleTable <- list(sample = c("WT", "A4", "A17"), replica = c(1, 2)) %>%
  expand_grid(!!!.) %>%  
  pmap_chr(paste, sep = "_") %>%
  tibble(sampleName = .) %>%
   mutate(fileName = paste0(tolower(sampleName), ".counts.unstranded.tsv"),
          condition = str_remove(sampleName, "_[:digit:]") %>% factor(levels = c("WT", "A4", "A17")),
          replica = str_remove(sampleName, ".*_") %>% as_factor()) %>%
   arrange(condition, replica) %>% 
   mutate(description = c("WT", "WT", "A.T1 TET3", "A.T1 TET3", "A.T1 NT", "A.T1 NT")))

write_tsv(sampleTable, "../data/4_counts/sample_table.tsv")
saveRDS(sampleTable, file = "../data/4_counts/sample_table.RDS")
```

Make `dds` object form counts ans run DESeq
```{r}
dds <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable, 
                                  directory = "../data/4_counts", 
                                  design = ~ condition) %>% 
  DESeq()
```
Get normalised counts
```{r}
cts <- dds %>%
  counts(normalized = TRUE) %>% 
  as_tibble(rownames = "id")

```

Get vst (Variance Stabilised Transformation), it will make your valued approach a normal distribution, as compared to a negative binomial distribution of the normalised counts.
```{r}
vsd <- dds %>% 
  vst(blind = FALSE) %>% 
  assay() %>%
  as_tibble(rownames = "id") #%>% 
  #filter(!id %in% c("gene:NeoR/KanR", "gene:TET3c"))
```

Get TPMs
```{r}
tpms <- tpms_from_dds(dds , "/path/to/genome/ITAG4.1/ITAG4.1_gene_lengths.tsv")
```

Save the objects as `.tsv` and as `.RDS` object
```{r save-objects}
saveRDS(dds, file = "../data/4_counts/counts_dds.RDS")

write_tsv(cts,  "../data/4_counts/counts_normalised.tsv")
write_tsv(vsd,  "../data/4_counts/counts_vsd.tsv")
write_tsv(tpms, "../data/4_counts/counts_tpm.tsv")
```

## Scatter plots 


```{r}
# Function to have a matrix of scatter plots
scatter_matrix <- function(cts, condition, log_values = TRUE, title = paste("Scatter plot of log(normalised counts) from", condition), lims = c(NA, NA)) {
  cts %>%
    select(id, starts_with(condition)) %>%
    {if(log_values) mutate(., across(-id, ~ log2(1 + .))) else .} %>%
    add_column(slope = 1) %>% 
    add_column(intercept = 0) %>%
    ggplot(aes(x = .panel_x, y = .panel_y)) +
    geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf), fill = "#440154FF") +
    geom_point(shape = 16, size = 0.3, alpha = 0.5) %>% rasterise(dpi = 300) +
    geom_point(shape = 16, size = 0, alpha = 0) +
    geom_density_2d_filled(contour_var = "ndensity") +
    geom_abline(aes(intercept = intercept, slope = slope), color = "red", alpha = 0.5) +
    geom_autohistogram(color = 1, fill = 3) +
    facet_matrix(vars(-c(id, slope, intercept)), layer.upper = c(2,5), layer.lower = c(1,3,4,5), layer.diag = 6) +
    coord_cartesian(xlim = lims, ylim = lims) +
    theme(axis.line = element_blank(), panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), 
        strip.background = element_blank(), plot.title = element_text(hjust = 0.5), legend.position = "none") +
    ggtitle(title)
}
```

### All conditions compared (vsd)

```{r fig.height=10, fig.width=10}
scatter_matrix(vsd, c("W", "A"), title = "", log_values = FALSE, lims = c(NA, 18)) +
   labs(title = "Scatter plots of vsd (variance stabilised transformed counts)",
        subtitle = "diagonal of the matrix represents distribution of expression values \nbottom half show density plot of expression values",
        caption = "Disclaimer: y axis on diagonal matrix doesn't match the axis labels") 

ggsave(filename = "../figures/scatter_matrix_vsd_all.pdf", height = 13, width = 12)
```

### Each condition separately (cts) {.tabset .tabset-pills}

#### WT

```{r fig.height=5, fig.width=5}
scatter_matrix(cts, "WT")
```

#### A4 

```{r fig.height=5, fig.width=5}
scatter_matrix(cts, "A4")
```

#### A17

```{r fig.height=5, fig.width=5}
scatter_matrix(cts, "A17")
```

### All conditions separated {.tabset .tabset-pills}

#### counts
 
```{r fig.height=5, fig.width=11}
(scatter_matrix_cts <- scatter_matrix(cts, "WT", title = "WT") + 
   scatter_matrix(cts, "A4", title = "A4") +
   scatter_matrix(cts, "A17", title = "A17") +
   plot_annotation(title = "Scatter plots of log(1 + normalised counts)",
                   subtitle = "diagonal of the matrix represents distribution of expression values \nbottom half show density plot of expression values",
                   caption = "Disclaimer: y axis on diagonal matrix doesn't match the axis labels")) 

ggsave(filename = "../figures/scatter_matrix_normalised_counts.pdf", plot = scatter_matrix_cts, height = 5, width = 11)
```

#### vsd {.active}

```{r fig.height=5, fig.width=11}
(scatter_matrix_vsd <- scatter_matrix(vsd, "WT", title = "WT", log_values = FALSE) + 
   scatter_matrix(vsd, "A4", title = "A4", log_values = FALSE) +
   scatter_matrix(vsd, "A17", title = "A17", log_values = FALSE) +
   plot_annotation(title = "Scatter plots of vsd (variance stabilised transformed counts)",
                   subtitle = "diagonal of the matrix represents distribution of expression values \nbottom half show density plot of expression values",
                   caption = "Disclaimer: y axis on diagonal matrix doesn't match the axis labels")) 

ggsave(filename = "../figures/scatter_matrix_vsd.pdf", plot = scatter_matrix_vsd, height = 5, width = 11)
```

## PCAs


```{r}
# General function for all RNA-seq that takes vsd (or other kind of counts) and returns a table of Principal Components 
pca_tidy = function(counts, ntop = 5000, sampletable = NULL) {
  # pick the top n genes by variance
  top_genes <- counts %>%
    rowwise(id) %>% # calculate the variance for each gene
    mutate(var = var(c_across(everything()))) %>%
    ungroup() %>%
    slice_max(var, n = ntop) %>% # pick only the top n
    select(-var)
  
  # perform a PCA on the data
  pca <- top_genes %>%
    select(-id) %>% 
    na.omit() %>% 
    t() %>% 
    prcomp()
  
  # convert it to a table
  pca$x %>%
    as_tibble(rownames = "sample") %>%
    rename_with(~ pca$sdev^2 %>%        # square the sd to get the variance
                  divide_by(sum(.)) %>% # divide it by the sum of variances
                  multiply_by(100) %>%  # *100 to get a percentage of variance
                  round() %>%           # round it
                  str_c(.x, " [", ., "%]"), # paste it to the PCx 
                .cols = -sample) %>% # exclude the sample column 
    {
      if (is.null(sampletable)) .
      else left_join(., sampletable, by = c("sample" = "sampleName"))
    }
}
```

```{r}
pca_tidy(vsd)
```

```{r}
# Aesthetics to plot the PCA and other plots
condition_colors <- c(WT  = "grey40", A4  = "tomato2", A17 = "slateblue2")

replica_colors <- c(WT_1  = "grey60", WT_2  = "grey30", 
                    A4_1  = "tomato1", A4_2  = "tomato3", 
                    A17_1 = "slateblue1", A17_2 = "slateblue3")

replica_shapes <- c("1" = 21, "2" = 24)
```


```{r}
# Function to make a matrix of PCA plots
pca_matrix = function(pctable, pcs = 1:3) {
  ggplot(pctable, aes(x = .panel_x, y = .panel_y)) + 
    theme_cowplot() +
    theme(axis.line=element_blank(), panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), 
          strip.background = element_blank(), plot.title = element_text(hjust = 0.5)) +
    scale_shape_manual(values = replica_shapes) +
    scale_fill_manual(values = condition_colors) +
    guides(fill = guide_legend(override.aes = list(shape = 22)), 
           shape = guide_legend(override.aes = list(fill = "darkseagreen2", size = 3))) +
    #geom_mark_hull(aes(fill = condition), radius = 0.05, concavity = 1, expand = 0.1) +
    geom_mark_ellipse(aes(fill = condition)) +
    geom_autohistogram(aes(fill = condition), color = "black", linewidth = 0.6, alpha = 0.7, show.legend = FALSE) +
    geom_point(aes(fill = condition, shape = replica), size = 2, color = "black") +
    facet_matrix(vars(pcs +1), layer.lower = c(1, 3), layer.upper = 3, layer.diag = 2) +
    labs(title = ("Matrix of PCs for vsd (variance stabilised transformed counts)"),
         subtitle = "Percentage of variance explained by each PC in square brackets")
}
```

> PC1 (75%) separates A17 from the rest, and PC2 WT from TET3 samples. PC3 explains variance between A4 replicas

```{r fig.height=6, fig.width=6}
(pca_plot_vsd <- pca_tidy(vsd, sampletable = sampleTable) %>%
  pca_matrix(pcs = 1:3) +
  labs(title = ("Matrix of PCs for vsd\n(variance stabilised transformed counts)"),
       subtitle = "Percentage of variance explained by each PC in square brackets"))

ggsave(filename = "../figures/PCA_matrix.pdf", plot = pca_plot_vsd, width = 6, height = 6)
```

