---
title: "DEGs"
subtitle: "_Tomato TET3 RNA-seq_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: pygments
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

```{r libraries-and-functions, include=FALSE}
library(DESeq2)
library(tidyverse)
library(magrittr)
library(cowplot); theme_set(theme_cowplot())
library(ggforce)
library(patchwork)
library(ComplexHeatmap)
ht_opt$message = FALSE
library(RColorBrewer)
library(EnhancedVolcano)
library(gprofiler2)
library(kableExtra)
library(shadowtext)
library(ggbeeswarm)
library(scales)

deg_cts = function(res, cts, ...) {
  res %>%
    #as_tibble(rownames = "id") %>%
    inner_join(cts, by = "id") %>%
    filter(...)
}

# Wrapper for enhancedVolcano
enhanced_volcano <- function(res, lab = NA, xlim = c(-5, 5), ylim = c(0, 12.5), pCutoff = 0.01, FCcutoff = 1, 
                             sample = "", reference = "", n_degs = filter(res, padj < pCutoff, abs(log2FoldChange) > FCcutoff) %>% nrow(),
                             comparison = str_c(sample, " vs ", reference),
                             title = str_c(comparison, " (" , n_degs, " DEGs)"), caption = FALSE,
                             label.x = 3.75, label.y = 9, size.title = 20, size.label = size.title / 3) {
  
  n_degs_down <- filter(res, padj < pCutoff, log2FoldChange < -FCcutoff) %>% nrow()
  n_degs_up   <- filter(res, padj < pCutoff, log2FoldChange >  FCcutoff) %>% nrow()
  
  EnhancedVolcano(res, lab = lab, x = 'log2FoldChange', y = 'padj', xlim = xlim, ylim = ylim, pCutoff = pCutoff, FCcutoff = FCcutoff, shape = 16, 
                  title = title, subtitle = NULL, caption = if(caption) paste0("total = ", nrow(res), " genes"), titleLabSize = size.title) +
    annotate("text", x = -label.x, y = label.y, label = paste(n_degs_down, "down"), size = size.label) +
    annotate("text", x =  label.x, y = label.y, label = paste(n_degs_up,     "up"),  size = size.label)
}
```

```{r load-objects}
sample_table <- readRDS("../data/4_counts/sample_table.RDS") %>% 
  separate(description, sep = " ", into = c("line", "genotype"), remove = FALSE, fill = "right") %>% 
  replace_na(list(genotype = "")) %>% 
  mutate(line = fct_inorder(line),
         genotype = fct_inorder(genotype))

dds <- readRDS("../data/4_counts/counts_dds.RDS")
cts  <- read_tsv("../data/4_counts/counts_normalised.tsv", show_col_types = FALSE)
tpms <- read_tsv("../data/4_counts/counts_tpm.tsv", show_col_types = FALSE)

# filter vsd so it has the same length as tpms
vsd  <- read_tsv("../data/4_counts/counts_vsd.tsv", show_col_types = FALSE) %>% 
  filter(id %in% tpms$id)
```

### Transgene expression

Before computing Differentially expressed genes let's have a look at the expression of the transgenes in the T-DNA, in our case the human TET3 catalytic domain and the Kanamycin resistance gene. 

```{r aesthetic-variables}
sample_colors      <- c(WT    = "grey40",          A4  = "tomato2",       A17 = "slateblue2")
description_colors <- c("WT"  = "grey40", "A.T1 TET3"  = "tomato2", "A.T1 NT" = "slateblue2")
```

```{r}
single_gene_expression_plot <- function(cts, gene_id, title = gene_id, sampletable = sample_table,
                                        colors = description_colors) {
  cts %>% 
    filter(id == gene_id) %>% 
    pivot_longer(-id, names_to = "sample", values_to = "counts") %>% 
    left_join(sampletable, by = c("sample" = "sampleName")) %>% 
    ggplot(aes(x = genotype, y = counts, fill = description, shape = )) +
    facet_grid(cols = vars(line), scales = "free_x", space = "free_x", switch = "x") +
    stat_summary(geom = "col", fun = mean, alpha = 0.7,col = 1, position = "dodge2") +
    geom_quasirandom(shape = 21, size = 3, color = 1, dodge.width = 0.8) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_fill_manual("", values = colors) +
    labs(title = title, y = "normalised counts") +
    theme(axis.line.x = element_blank(), #panel.border = element_rect(colour="black", size=1, fill=NA), 
          strip.background = element_rect(fill = "cornsilk"), axis.title.x = element_blank(),
          axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0), strip.placement = "outside",
          axis.ticks.x = element_blank(), panel.grid.major.y = element_line(linewidth = 0.2, color = "grey50"),
          plot.title = element_text(size = 12), legend.position = "none")
}
```

> TET3 is very highly expressed in A.T1 TET3, the highest after CAT1.
The Kanamycin resistance gene is also high expressed around ranking 1000.
A1.T1 NT plant is not expressing the transgenes, confirming it is non transgenic

```{r fig.width=6, fig.height=4}
single_gene_expression_plot(cts, gene_id = "gene:NeoR/KanR", title = "KanR (Kanamycin resistance)") +
  single_gene_expression_plot(cts, gene_id = "gene:TET3c", title = "TET3 Catalytic Domain")  +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Expression of T-DNA transgenes",
                  subtitle = "columns represent average value",
                  caption = "TET3 is very highly expressed in A.T1 TET3, the highest after CAT1.
The Kanamycin resistance gene is also high expressed around ranking 1000.
A1.T1 NT plant is not expressing the transgenes, confirming it is non transgenic.")
  
ggsave2("../figures/transgenes-expression.pdf", width = 6, height = 4)
```

```{r}
cts %>% arrange(-A4_2)
```


### DEGs

Get the statistics for each of the comparisons in a list
```{r}
res = list(
  A4_vs_WT  = results(dds, contrast=c("condition", "A4",  "WT")) %>% as_tibble(rownames = "id"),
  A17_vs_WT = results(dds, contrast=c("condition", "A17", "WT")) %>% as_tibble(rownames = "id"),
  A17_vs_A4 = results(dds, contrast=c("condition", "A17", "A4")) %>% as_tibble(rownames = "id")
)
```

get the counts for the DEGs (also as a list)
```{r}
DEGs_cts <- map(res, ~ deg_cts(.x, cts))
```

get the variance stabilized counts for the DEGs (also as a list)
```{r}
DEGs_vsd <- map(res, ~ deg_cts(.x, vsd))
```

Join all DEGs with its counts in a unique table
```{r}
DEGs_vsd_all <- DEGs_vsd %>%
  map_df(.id = "comparison", ~ .x) %>%
  mutate(comparison = as_factor(comparison))
```

And you can filter it as a tibble

```{r}
DEGs_vsd_all %>%
  filter(id == "gene:Solyc02g071010.1")
```

Alternative you can filter and keep the list structure

```{r}
DEGs_vsd_p01_lfc1 <- map(DEGs_vsd, ~ filter(., padj < 0.01, abs(log2FoldChange) > 1))
```

Write the DEGs to a Excel file with tabs for each comparison

```{r}
dir.create("../data/6_degs/", showWarnings = FALSE)
writexl::write_xlsx(DEGs_vsd_p01_lfc1, path = "../data/6_degs/DEGs_p01_lfc1_with_vsd.xlsx")
```

And save it as a RDS object for easy loading

```{r}
saveRDS(DEGs_vsd_p01_lfc1, file = "../data/6_degs/DEGs.RDS")
```

### DEGs number

```{r fig.height=5, fig.width=5}
DEGs_vsd_all %>%
  filter(padj < 0.01, abs(log2FoldChange) > 1) %>% 
  group_by(comparison,  log2FoldChange > 0) %>%
  dplyr::count(comparison) %>%
  rename(kind = `log2FoldChange > 0`, DEGs = n) %>%
  mutate(kind = case_when(kind == TRUE ~ "up",
                          kind == FALSE ~ "down"),
         group = case_when(str_detect(comparison, "WT") ~ "A.T1\n vs WT",
                           .default = "between\nA.T1") %>% 
           factor(levels = c("A.T1\n vs WT", "between\nA.T1")),
         comparison = str_replace(comparison, "A4", "A.T1 TET3") %>% 
           str_replace("A17", "A.T1 NT") %>% 
           str_replace_all("_", "\n")) %>%
  ggplot(aes(comparison, DEGs, fill = kind, label = DEGs)) +
  facet_grid(cols = vars(group), scales = "free", space = "free") +
  geom_col(position = "stack", color = 1, alpha = 0.7) +
  geom_shadowtext(position = "stack", color = "white", check_overlap = TRUE) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Number of DEGs",
       subtitle = "p.adj < 0.01 & |logFC| > 1",
       caption = "Tomato TET3 RNA-seq") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(axis.line=element_blank(), axis.ticks.x = element_blank(), 
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA),
        legend.position = "top", legend.direction = "horizontal",
        strip.background = element_rect(fill = "pink"))

ggsave("../figures/DEGs_number.pdf", width = 5, height = 5)
```



### Volcano plots {.tabset}

Using enhanced volcano

#### Zoom out

```{r fig.width=10,  fig.height=16}
(volcanos <- enhanced_volcano(res$A4_vs_WT,  sample = "A.T1 TET3 (A4)", reference = "WT", xlim = c(-9, 9), ylim = c(0, 110), label.x = 7, label.y = 60) /
             enhanced_volcano(res$A17_vs_WT, sample = "A.T1 NT (A17)",  reference = "WT", xlim = c(-9, 9), ylim = c(0, 110), label.x = 7, label.y = 60) /
             enhanced_volcano(res$A17_vs_A4, sample = "A.T1 NT", reference = "A.T1 TET3", xlim = c(-9, 9), ylim = c(0, 110), label.x = 7, label.y = 60, caption = TRUE) +
  plot_layout(guides = 'collect') &  theme(legend.position = 'bottom'))

ggsave(filename = "../figures/volcano_plots_all.pdf", plot = volcanos, width = 10, height = 16)
```

#### Zoom in

```{r fig.width=10,  fig.height=16}
(volcanos <- enhanced_volcano(res$A4_vs_WT,  sample = "A.T1 TET3 (A4)", reference = "WT") /
             enhanced_volcano(res$A17_vs_WT, sample = "A.T1 NT (A17)",  reference = "WT") /
             enhanced_volcano(res$A17_vs_A4, sample = "A.T1 NT", reference = "A.T1 TET3", caption = TRUE) +
  plot_layout(guides = 'collect') &  theme(legend.position = 'bottom'))

ggsave(filename = "../figures/volcano_plots_zoom.pdf", plot = volcanos, width = 10, height = 16)
```

#### Zoom in for figure

```{r fig.width=10,  fig.height=16}
(volcanos <- enhanced_volcano(res$A4_vs_WT,  sample = "TET3", reference = "WT", xlim = c(-7, 7), ylim = c(0, 50), label.x = 5, label.y = 40) /
             enhanced_volcano(res$A17_vs_WT, sample = "NTS",  reference = "WT", xlim = c(-7, 7), ylim = c(0, 50), label.x = 5, label.y = 40) /
             enhanced_volcano(res$A17_vs_A4, sample = "NTS", reference = "TET3", xlim = c(-7, 7), ylim = c(0, 50), label.x = 5, label.y = 40, caption = TRUE) +
  plot_layout(guides = 'collect') &  theme(legend.position = 'bottom'))

ggsave(filename = "../figures/volcano_plots_zoom-2.pdf", plot = volcanos, width = 10, height = 16)
```

### Comparison of DEGs with different thresholds

```{r fig.height=7, fig.width=11}
visualise_degs <- function(DEGs_vsd_all, padj_lim, lfc_lim, fill = bin_padj) {
  DEGs_vsd_all %>%
    filter(padj <= padj_lim, abs(log2FoldChange) > lfc_lim) %>%
    mutate(rank = ntile(padj, 10), 
           logpadj = -log10(padj),
           bin_padj = cut(padj, c(1, 0.5, 0.1, 0.05, 0.01, 0.001, 0.0001, 0)),
           bin_padj = fct_reorder(bin_padj, -padj),
           bin_l2fc = cut(log2FoldChange, c(-Inf, -10, -4, -2, -1, -0.5, 0, 0.5, 1, 2, 4, 10, Inf)),
           bin_l2fc = fct_reorder(bin_l2fc, log2FoldChange),
           bin_absl2fc = cut(abs(log2FoldChange), c(0, 0.5, 1, 2, 4, 10, Inf)),
           bin_absl2fc = fct_reorder(bin_absl2fc, abs(log2FoldChange))) %>%
    group_by(comparison, {{ fill }}) %>%
    summarise(count = n(), .groups = 'drop') %>%
    ggplot(aes(x = count, y = comparison, fill = {{ fill }})) +
    geom_col(color = NA) +
    ggtitle(str_c("DEGs with a p-adj <= ", padj_lim, " & abs(log2FC) > ", lfc_lim)) +
    scale_fill_viridis_d(direction = -1, option = if_else(deparse(substitute(fill)) == "bin_padj" , "plasma", "viridis")) +
    #scale_fill_distiller(palette = 2)
    scale_y_discrete(limits = rev)
}

deg_pv <- visualise_degs(DEGs_vsd_all, padj_lim = 0.05, lfc_lim = 1, fill = bin_padj)
deg_fc <- visualise_degs(DEGs_vsd_all, padj_lim = 0.05, lfc_lim = 1, fill = bin_absl2fc)

all_pv <- visualise_degs(DEGs_vsd_all, padj_lim = 1, lfc_lim = 0, fill = bin_padj)
all_fc <- visualise_degs(DEGs_vsd_all, padj_lim = 1, lfc_lim = 0, fill = bin_absl2fc)

(DEGs_padj_l2fc <- (all_pv + all_fc + deg_pv + deg_fc))
ggsave2("../figures/DEGs_padj_l2fc.pdf", width = 11, height = 7)
```

### Heatmaps

```{r}
deg_mat = function(res, cts, ..., scaled = FALSE) {
  deg_cts(res, cts, ...) %>% # ...) %>%
    column_to_rownames("id") %>%
    select(-c(baseMean:padj)) %>%
    rename_with(~ .x %>% toupper() %>% str_replace_all("_", " ")) %>%
    as.matrix() %>%
    {if(scaled) {t(.) %>% scale() %>% t()}
      else .}
}

hm_wrapper = function(mat, scaled = F, cluster_rows = T, cluster_columns = T, title = "") {
  if(scaled) {col = rev(brewer.pal(9, "RdBu"));  name = "scaled\nexpression"}
  else       {col =     brewer.pal(9, "YlOrRd"); name = "expression"}
  Heatmap(mat, cluster_rows = cluster_rows, cluster_columns = cluster_columns, show_row_names = F, col = col, name = name, column_title = title)
}

deg_hm = function(res, cts, ..., scaled = F, file = F, title = NULL) {
  mat = deg_mat(res = res, cts = cts, ... = ..., scaled = scaled)
  if (file != 0) {
    pdf(file, height = 5, width = 5)
      hm <- hm_wrapper(mat, scaled = scaled)
      draw(hm, column_title = paste(nrow(mat), title, "DEGs", deparse(substitute(...))))
    dev.off()
  } else {
    hm_wrapper(mat, scaled = scaled, title = title)
  }
}
```

```{r}
DEGs_vsd_p01_lfc1$A17_vs_WT %>%
  arrange(-log2FoldChange)

DEGs_vsd_p01_lfc1$A17_vs_WT %>%
  arrange(-baseMean)
```

```{r}
mats <- list(
  raw    = map(res, ~ deg_mat(.x, vsd, padj < 0.01, abs(log2FoldChange) > 1)),
  scaled = map(res, ~ deg_mat(.x, vsd, padj < 0.01, abs(log2FoldChange) > 1, scaled = T))
) %>%
  transpose() # to get A4$raw instead of raw$A4
```

```{r}
hm_wrapper(mats$A4$scaled, scaled = T) +
  hm_wrapper(mats$A4$raw)
```

```{r}
deg_hm(res$A4_vs_WT, vsd, padj < 0.01, abs(log2FoldChange) > 1, title = "DEGs A4", scaled = TRUE) +
  deg_hm(res$A4_vs_WT, tpms, padj < 0.01, abs(log2FoldChange) > 1, title = "DEGs A4", scaled = FALSE)
```
A4 
```{r}
deg_hm(res$A4, vsd, 
       padj < 0.01, abs(log2FoldChange) > 1, title = "A4", scaled = TRUE, file="../figures/heatmap_DEGs_A4_lfc1_p0.01.pdf")
deg_hm(res$A4, vsd %>% select(id, WT_1:A4_2),
       padj < 0.01, abs(log2FoldChange) > 1, title = "A4", scaled = TRUE, file="../figures/heatmap_DEGs_A4_only_lfc1_p0.01.pdf")
```

```{r}
hms_A4 <- deg_hm(res$A4_vs_WT,   vsd,                padj < 0.01, abs(log2FoldChange) > 1, title = "A4 DEGs", scaled = TRUE) +
  deg_hm(res$A4_vs_WT, vsd %>% select(id, WT_1:A4_2), padj < 0.01, abs(log2FoldChange) > 1, title = "A4 only", scaled = TRUE) +
  deg_hm(res$A4_vs_WT, tpms,                          padj < 0.01, abs(log2FoldChange) > 1, title = "TPMs", scaled = FALSE) 

hms_A17 <- deg_hm(res$A17_vs_WT,   vsd,                             padj < 0.01, abs(log2FoldChange) > 1, title = "A17 DEGs", scaled = TRUE) +
  deg_hm(res$A17_vs_WT, vsd %>% select(id, WT_1:WT_2, A17_1:A17_2), padj < 0.01, abs(log2FoldChange) > 1, title = "A17 only", scaled = TRUE) +
  deg_hm(res$A17_vs_WT, tpms,                                       padj < 0.01, abs(log2FoldChange) > 1, title = "TPMs", scaled = FALSE)
```

```{r fig.width=6, fig.height=7}
pdf(file = "../figures/heatmap_DEGs_all.pdf", height = 9, width = 7) 
grid.newpage()
pushViewport(viewport(x = 0, y = 1, height = 0.5, just = c("left", "top")))
draw(hms_A4, newpage = FALSE, column_title = "391 A4 DEGs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()

pushViewport(viewport(x = 0, y = 0.5, height = 0.5, just = c("left", "top")))
draw(hms_A17, newpage = FALSE, column_title = "1086 A17 DEGs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()
dev.off()
```

```{r}
hms_A4 <- deg_hm(res$A4_vs_WT, tpms,                          padj < 0.01, abs(log2FoldChange) > 1, title = "TPMs", scaled = FALSE)  +
  deg_hm(res$A4_vs_WT,   vsd,                padj < 0.01, abs(log2FoldChange) > 1, title = "A4 DEGs", scaled = TRUE) +
  deg_hm(res$A4_vs_WT, vsd %>% select(id, WT_1:A4_2), padj < 0.01, abs(log2FoldChange) > 1, title = "A4 only", scaled = TRUE)

hms_A17 <- deg_hm(res$A17_vs_WT, tpms,                                       padj < 0.01, abs(log2FoldChange) > 1, title = "TPMs", scaled = FALSE) +
  deg_hm(res$A17_vs_WT,   vsd,                             padj < 0.01, abs(log2FoldChange) > 1, title = "A17 DEGs", scaled = TRUE) +
  deg_hm(res$A17_vs_WT, vsd %>% select(id, WT_1:WT_2, A17_1:A17_2), padj < 0.01, abs(log2FoldChange) > 1, title = "A17 only", scaled = TRUE)
```

```{r fig.width=6, fig.height=7}
pdf(file = "../figures/heatmap_DEGs_all-unscaled.pdf", height = 9, width = 7) 
grid.newpage()
pushViewport(viewport(x = 0, y = 1, height = 0.5, just = c("left", "top")))
draw(hms_A4, newpage = FALSE, column_title = "391 A4 DEGs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()

pushViewport(viewport(x = 0, y = 0.5, height = 0.5, just = c("left", "top")))
draw(hms_A17, newpage = FALSE, column_title = "1086 A17 DEGs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()
dev.off()
```

### Go term analysis

Done in a different Rmd file

### DEGs across genome

```{r}
library(plyranges)

facet_chr <- function(chr = chr, rows = NULL, scales = "free_x", space = "free_x", breaks = 40e6) {
  list(facet_grid(rows = vars({{ rows }}), cols = vars({{ chr }}), scales = scales, space = space),
       scale_x_continuous(expand = c(0, 0), limits = c(0, NA), breaks = scales::breaks_width(breaks), labels = scales::label_number_si()),
       xlab("Position across the genome (Mbp)"))
}
```

```{r}

genes_df <- read_gff("/path/to/genome/version_4/ITAG4.1_gene_models.gff") %>%
  filter(type == "gene",
         seqnames != "SL4.0ch00") %>%
  select(ID, Name) %>%
  as_tibble() %>%
  mutate(seqnames = str_remove(seqnames, "SL4.0"))

genes <- genes_df %>%
  as_granges()

load(file = "/path/to/project/rnaseq/DEGs.Rdata")
degs_df <- DEGs_cts_p01_lfc1 %>%
  map_df(.id = "comparison", ~ .x %>% 
           mutate(direction = case_when(log2FoldChange > 0 ~ "up",
                                        log2FoldChange < 0 ~ "down",
                                        TRUE ~ "something is wrong"))) %>%
  left_join(genes_df, by = c("id" = "ID")) %>%
  select(-id) %>%
  relocate(seqnames:Name, comparison, direction, log2FoldChange, padj)
```


##### Density

```{r fig.height=4, fig.width=10}

genes %>%
  as_tibble() %>%
  ggplot(aes(x = start)) +
  facet_chr(chr = seqnames) +
  geom_density(adjust = 1/1, alpha = 0.2, color = 1, fill = 1) +
  scale_color_brewer(palette = "Set1", direction = -1) + 
  scale_fill_brewer(palette  = "Set1", direction = -1) +
  guides(color = "none", fill = "none") 

ggplot(degs_df, aes(x = start, color = direction, fill = direction)) +
  facet_chr(chr = seqnames, rows = comparison) +
  geom_density(adjust = 1/3, alpha = 0.2) +
  scale_color_brewer(palette = "Set1", direction = -1) + 
  scale_fill_brewer(palette  = "Set1", direction = -1)
```

##### Count

```{r fig.height=4, fig.width=10}
genes %>%
  as_tibble() %>%
  ggplot(aes(x = start)) +
  facet_chr(chr = seqnames, scales = "free", space = "free") +
  geom_area(aes(y = ..count..), stat = "bin", alpha = 0.2, binwidth = 3e6, color = 2, fill = 2) +
  ylab("genes per 3Mb")
ggsave("/path/to/project/rnaseq/figures/genes_across_genome.pdf", width = 8, height = 3)

ggplot(degs_df, aes(x = start, color = direction, fill = direction)) +
  facet_chr(chr = seqnames, rows = comparison, scales = "free", space = "free") +
  geom_area(aes(y = ..count..), stat = "bin", alpha = 0.2, binwidth = 3e6) +
  scale_color_brewer(palette = "Set1", direction = -1) + 
  scale_fill_brewer(palette  = "Set1", direction = -1) +
  #guides(color = "none", fill = "none")  +
  ylab("DEGs per 3Mb")
ggsave("/path/to/project/rnaseq/figures/DEGs_across_genome.pdf", width = 8, height = 5)
```

