---
title: "GO analysis DEGs TET3"
author: "Javier Antúnez Sánchez"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(kableExtra)
library(gprofiler2)
```

Read Differentially Expressed Genes list
```{r}
DEGs_vsd_p01_lfc1 <- readRDS(file = "../data/6_degs/DEGs.RDS")

kbl(DEGs_vsd_p01_lfc1$A4_vs_WT) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "200px")
```

## Gene Ontology {.tabset}

```{r}
# Function to get GO for tomato from DEGs lists
gost_tom <- function(query, name = deparse(substitute(query))) {
  lst <- list(query$id %>% str_remove("gene:"))
  names(lst) = name
  gost(query = lst, organism = 'slycopersicum', significant = F, user_threshold = 0.01)
}

#Function to display table of GO terms
go_kable <- function(go) {
  go$result %>%
    dplyr::select(term_name, term_id, source, p_value) %>% 
    mutate(log10_pv = log10(p_value) %>% round()) %>%
    kbl() %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "650px", height = "300px", fixed_thead = TRUE)
}


# Generate list of GO results
go <- list()
go$A4_vs_WT <- list()
go$A4_vs_WT$all  <- gost_tom(query = DEGs_vsd_p01_lfc1$A4_vs_WT, name = "DEGs A4 vs WT")
go$A4_vs_WT$up   <- gost_tom(query = DEGs_vsd_p01_lfc1$A4_vs_WT %>% filter(log2FoldChange > 0), name = "DEGs A4 vs WT Up")
go$A4_vs_WT$down <- gost_tom(query = DEGs_vsd_p01_lfc1$A4_vs_WT %>% filter(log2FoldChange < 0), name = "DEGs A4 vs WT Down")
go$A17_vs_WT <- list()
go$A17_vs_WT$all  <- gost_tom(query = DEGs_vsd_p01_lfc1$A17_vs_WT, name = "DEGs A17 vs WT")
go$A17_vs_WT$up   <- gost_tom(query = DEGs_vsd_p01_lfc1$A17_vs_WT %>% filter(log2FoldChange > 0), name = "DEGs A17 vs WT Up")
go$A17_vs_WT$down <- gost_tom(query = DEGs_vsd_p01_lfc1$A17_vs_WT %>% filter(log2FoldChange < 0), name = "DEGs A17 vs WT Down")
go$A17_vs_A4 <- list()
go$A17_vs_A4$all  <- gost_tom(query = DEGs_vsd_p01_lfc1$A17_vs_A4, name = "DEGs A17 vs A4")
go$A17_vs_A4$up   <- gost_tom(query = DEGs_vsd_p01_lfc1$A17_vs_A4 %>% filter(log2FoldChange > 0), name = "DEGs A17 vs A4 Up")
go$A17_vs_A4$down <- gost_tom(query = DEGs_vsd_p01_lfc1$A17_vs_A4 %>% filter(log2FoldChange < 0), name = "DEGs A17 vs A4 Down")
```


### A4 vs WT {.tabset}

#### all {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A4_vs_WT$all, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A4_vs_WT$all, capped = FALSE, interactive = TRUE)
```

##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler))
```{r}
go_kable(go$A4_vs_WT$all)
```

#### up {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A4_vs_WT$up, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A4_vs_WT$up, capped = FALSE, interactive = TRUE)
```

##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler))
```{r}
go_kable(go$A4_vs_WT$up)
```

#### down {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A4_vs_WT$down, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A4_vs_WT$down, capped = FALSE, interactive = TRUE)
```

```{r include=FALSE}
publish_gostplot(p = gostplot(go$A4_vs_WT$down, capped = FALSE, interactive = FALSE),
                 highlight_terms = go$A4_vs_WT$down$result %>%
                   filter(p_value < 0.05) %>% 
                   pull(term_id),
                 filename = "../figures/GO_AT1-TET3-vs-WT_downregulated.pdf", width = 8, height = 6)
```


##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler/))
```{r}
go_kable(go$A4$down)
```

### A17 vs WT {.tabset}
#### all {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_WT$all, capped = TRUE, interactive = TRUE)
```

##### uncapped
```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_WT$all, capped = FALSE, interactive = TRUE)
```

##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler))
```{r}
go_kable(go$A17_vs_WT$all)
```

#### up {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_WT$up, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_WT$up, capped = FALSE, interactive = TRUE)
```

```{r, include=FALSE}
publish_gostplot(p = gostplot(go$A17_vs_WT$up, capped = FALSE, interactive = FALSE), 
                 highlight_terms = go$A17_vs_WT$up$result %>%
                   filter(p_value < 1e-10) %>% 
                   pull(term_id),
                 filename = "../figures/GO_AT1-NT-vs-WT_upregulated.pdf", width = 8, height = 11)
```


##### table

Table of GO terms 

```{r}
go_kable(go$A17_vs_WT$up)
```

#### down {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_WT$down, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_WT$down, capped = FALSE, interactive = TRUE)
```

##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler/))
```{r}
go_kable(go$A17_vs_WT$down)
```




### A17 vs A4 {.tabset}
#### all {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_A4$all, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_A4$all, capped = FALSE, interactive = TRUE)
```

##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler))
```{r}
go_kable(go$A17_vs_A4$all)
```

#### up {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_A4$up, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_A4$up, capped = FALSE, interactive = TRUE)
```

```{r, include=FALSE}
publish_gostplot(p = gostplot(go$A17_vs_A4$up, capped = FALSE, interactive = FALSE), 
                 highlight_terms = go$A17_vs_A4$up$result %>%
                   filter(p_value < 1e-7) %>% 
                   pull(term_id),
                 filename = "../figures/GO_AT1-NT-vs-AT1-TET3_upregulated-.pdf", width = 8, height = 12)
```


##### table

Table of GO terms 

```{r}
go_kable(go$A17_vs_A4$up)
```

#### down {.tabset}

##### capped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_A4$down, capped = TRUE, interactive = TRUE)
```

##### uncapped

```{r fig.height=4, fig.width=8}
gostplot(go$A17_vs_A4$down, capped = FALSE, interactive = TRUE)
```

```{r include=FALSE}
publish_gostplot(p = gostplot(go$A17_vs_A4$down, capped = FALSE, interactive = FALSE),
                 highlight_terms = go$A17_vs_A4$down$result %>%
                   filter(p_value < 0.01) %>% 
                   pull(term_id),
                 filename = "../figures/GO_AT1-NT-vs-AT1-TET3_downregulated.pdf", width = 8, height = 4)
```

##### table

Table of GO terms ([detailed table on gProfiler web](https://biit.cs.ut.ee/gprofiler/))
```{r}
go_kable(go$A17_vs_A4$down)
```

# Intersections {.tabset}

```{r}
DEGs_TET3_down_NT_up <- inner_join(DEGs_vsd_p01_lfc1$A17_vs_WT %>% filter(log2FoldChange > 0),
                                   DEGs_vsd_p01_lfc1$A4_vs_WT %>% filter(log2FoldChange < 0),
                                   by = join_by(id, WT_1, WT_2, A4_1, A4_2, A17_1, A17_2), suffix = c("_TET3", "_NT"))

ids_DEGs_TET3_down_NT_up <- DEGs_TET3_down_NT_up %>% pull(id)

write_lines(ids_DEGs_TET3_down_NT_up, file = "../data/6_degs/Gene-IDs_intersection-TET3-down-NT-up.txt")

go$TET3_down_NT_up <- gost_tom(query = DEGs_TET3_down_NT_up, name = "Intersection DEGs TET3 down with NT up")
```

```{r fig.height=4, fig.width=8}
gostplot(go$TET3_down_NT_up, capped = FALSE, interactive = TRUE)
```

```{r include=FALSE}
publish_gostplot(p = gostplot(go$TET3_down_NT_up, capped = FALSE, interactive = FALSE),
                 highlight_terms = go$TET3_down_NT_up$result %>%
                   filter(p_value < 0.05) %>% 
                   pull(term_id),
                 filename = "../figures/GO_intersection-AT1-TET3-down-NT-up.pdf", width = 8, height = 4)
```
