---
title: "DEGs"
subtitle: "_Tomato TET3 RNA-seq_"
author: "Javier Antúnez Sánchez"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_notebook: 
    theme: flatly
    highlight: pygments
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```

```{r libraries-and-functions, include=FALSE}
library(DESeq2)
library(tidyverse)
library(magrittr)
library(cowplot); theme_set(theme_cowplot())
library(ggforce)
library(patchwork)
library(ComplexHeatmap)
ht_opt$message = FALSE
library(RColorBrewer)
library(EnhancedVolcano)
library(gprofiler2)
library(kableExtra)
library(shadowtext)
library(ggbeeswarm)
library(scales)
library(tidyHeatmap)
library(ggpubr)     # gives stat_compare_means() & stat_pvalue_manual()
library(rstatix)    # tidy anova_test() & tukey_hsd()

deg_cts = function(res, cts, ...) {
  res %>%
    #as_tibble(rownames = "id") %>%
    inner_join(cts, by = "id") %>%
    filter(...)
}

# Wrapper for enhancedVolcano
enhanced_volcano <- function(res, lab = NA, xlim = c(-5, 5), ylim = c(0, 12.5), pCutoff = 0.01, FCcutoff = 1, 
                             sample = "", reference = "", n_degs = filter(res, padj < pCutoff, abs(log2FoldChange) > FCcutoff) %>% nrow(),
                             comparison = str_c(sample, " vs ", reference),
                             title = str_c(comparison, " (" , n_degs, " DEXs)"), caption = FALSE,
                             label.x = 3.75, label.y = 9, size.title = 20, size.label = size.title / 3, ...) {
  
  n_degs_down <- filter(res, padj < pCutoff, log2FoldChange < -FCcutoff) %>% nrow()
  n_degs_up   <- filter(res, padj < pCutoff, log2FoldChange >  FCcutoff) %>% nrow()
  
  EnhancedVolcano(res, lab = lab, x = 'log2FoldChange', y = 'padj', xlim = xlim, ylim = ylim, pCutoff = pCutoff, FCcutoff = FCcutoff, shape = 16, 
                  title = title, subtitle = NULL, caption = if(caption) paste0("total = ", nrow(res), " elements"), titleLabSize = size.title, ... = ...) +
    annotate("text", x = -label.x, y = label.y, label = paste(n_degs_down, "down"), size = size.label) +
    annotate("text", x =  label.x, y = label.y, label = paste(n_degs_up,     "up"),  size = size.label)
}
```

```{r load-objects}
sample_table <- readRDS("../data/4_counts/TEcounts/sample_table_TEcounts.RDS") %>% 
  separate(description, sep = " ", into = c("line", "genotype"), remove = FALSE, fill = "right") %>% 
  replace_na(list(genotype = "")) %>% 
  mutate(line = fct_inorder(line),
         genotype = fct_inorder(genotype))

dds <- readRDS("../data/4_counts/TEcounts/TEcounts_dds.rds")
cts  <- read_tsv("../data/4_counts/TEcounts/TEcounts_normalised.tsv", show_col_types = FALSE)
vsd  <- read_tsv("../data/4_counts/TEcounts/TEcounts_vsd.tsv", show_col_types = FALSE)
```

```{r aesthetic-variables}
sample_colors      <- c(WT    = "grey40",          TET3  = "tomato2",       NTS = "slateblue2")
description_colors <- c("WT"  = "grey40", "A.T1 TET3"  = "tomato2", "A.T1 NT" = "slateblue2")
```

### DEGs

Get the statistics for each of the comparisons in a list
```{r}
res = list(
  TET3_vs_WT  = results(dds, contrast=c("condition", "TET3",  "WT")) %>% as_tibble(rownames = "id"),
  NTS_vs_WT = results(dds, contrast=c("condition", "NTS", "WT")) %>% as_tibble(rownames = "id"),
  NTS_vs_TET3 = results(dds, contrast=c("condition", "NTS", "TET3")) %>% as_tibble(rownames = "id")
)

res_genes <- map(res, ~ filter(.x,  str_starts(id, "gene")))
res_TEs   <- map(res, ~ filter(.x, !str_starts(id, "gene")))
```

get the counts for the DEGs (also as a list)
```{r}
DEGs_cts <- map(res, ~ deg_cts(.x, cts))
```

get the variance stabilized counts for the DEGs (also as a list)
```{r}
DEGs_vsd <- map(res, ~ deg_cts(.x, vsd))
```

Join all DEGs with its counts in a unique table
```{r}
DEGs_vsd_df <- DEGs_vsd %>%
  map_df(.id = "comparison", ~ .x) %>%
  mutate(comparison = as_factor(comparison), 
         element = case_when(str_starts(id, "gene") ~ "gene",
                             !str_starts(id, "gene") ~ "TE")) %>% 
  relocate(element, .after = id)
```

Filter and keep the list structure

```{r}
DEGs_vsd_p01_lfc1 <- map(DEGs_vsd, ~ filter(., padj < 0.01, abs(log2FoldChange) > 1) %>% 
                           mutate(element = case_when(str_starts(id, "gene") ~ "gene",
                                                      !str_starts(id, "gene") ~ "TE")) %>% 
                           relocate(element, .after = id))

DEGs_vsd_p05_lfc1 <- map(DEGs_vsd, ~ filter(., padj < 0.05, abs(log2FoldChange) > 1) %>% 
                           mutate(element = case_when(str_starts(id, "gene") ~ "gene",
                                                      !str_starts(id, "gene") ~ "TE")) %>% 
                           relocate(element, .after = id))
```

Write the DEGs to a Excel file with tabs for each comparison

```{r}
dir.create("../data/6_degs/TEcounts/", showWarnings = FALSE)
writexl::write_xlsx(DEGs_vsd_p01_lfc1, path = "../data/6_degs/TEcounts/DEGs_p01_lfc1_with_vsd.xlsx")
writexl::write_xlsx(DEGs_vsd_p05_lfc1, path = "../data/6_degs/TEcounts/DEGs_p05_lfc1_with_vsd.xlsx")
```

And save it as a RDS object for easy loading

```{r}
write_rds(DEGs_vsd_p01_lfc1, file = "../data/6_degs/TEcounts/DEGs-p01.rds")
write_rds(DEGs_vsd_p01_lfc1, file = "../data/6_degs/TEcounts/DEGs-p05.rds")
```

### DEGs number

```{r fig.height=5, fig.width=5}
DEGs_vsd_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1) %>% 
  group_by(comparison, element,  log2FoldChange > 0) %>%
  dplyr::count(comparison) %>%
  rename(kind = `log2FoldChange > 0`, DEGs = n) %>%
  mutate(kind = case_when(kind == TRUE ~ "up",
                          kind == FALSE ~ "down"),
         group = case_when(str_detect(comparison, "WT") ~ "A.T1\n vs WT",
                           .default = "between\nA.T1") %>% 
           factor(levels = c("A.T1\n vs WT", "between\nA.T1")),
         comparison = str_replace(comparison, "TET3", "A.T1 TET3") %>% 
           str_replace("NTS", "A.T1 NT") %>% 
           str_replace_all("_", "\n")) %>%
  ggplot(aes(comparison, DEGs, fill = kind, label = DEGs)) +
  facet_grid(cols = vars(group), rows = vars(element), scales = "free", space = "free_x") +
  geom_col(position = "stack", color = 1, alpha = 0.7) +
  geom_shadowtext(position = "stack", color = "white", check_overlap = TRUE) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Number of DEGs",
       subtitle = "p.adj < 0.01 & |logFC| > 1",
       caption = "Tomato TET3 RNA-seq") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(axis.line=element_blank(), axis.ticks.x = element_blank(), 
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA),
        legend.position = "top", legend.direction = "horizontal",
        strip.background = element_rect(fill = "pink"))

ggsave("../figures/TEcounts/DEGs_number.pdf", width = 5, height = 5)
```

```{r}
comparison_colors  <- c("NTS\nvs\nWT"         = "slateblue2", 
                        "TET3\nvs\nWT"       = "tomato2", 
                        "NTS\nvs\nTET3"  = "goldenrod2")

DEG_numbers <- DEGs_vsd_p01_lfc1 %>% 
  map_df(.id = "comparison", ~ .x) %>%
  mutate(comparison = as_factor(comparison)) %>% 
  group_by(comparison, element, log2FoldChange > 0) %>%
  dplyr::count(comparison) %>%
  rename(kind = `log2FoldChange > 0`, DEGs = n) %>%
  mutate(kind = case_when(kind == TRUE ~ "up",
                          kind == FALSE ~ "down"),
         comparison = str_replace(comparison, "A4", "A.T1 TET3") %>% 
           str_replace("A17", "A.T1 NT") %>% 
           str_replace_all("_", "\n"))
```

```{r}
ggplot(DEG_numbers %>% filter(kind == "down", element == "gene"), 
       aes(x = DEGs, y = comparison, fill = comparison, label = DEGs)) +
  facet_grid(cols = vars(kind), switch = "y") +
  geom_col() +
  #geom_shadowtext(position = position_stack(vjust = 0.5), color = "white") +
  geom_text(position = position_stack(vjust = 0.5)) +
  scale_y_discrete(position = "right") +
  scale_x_reverse(expand = expansion(mult = c(0.05, 0))) +
  scale_fill_manual(values = comparison_colors) +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),legend.position= "none",
        axis.ticks.y = element_blank(), axis.title.y = element_blank(), axis.line.y = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA)) +
  
  ggplot(DEG_numbers %>% filter(kind == "up", element == "gene"), 
       aes(x = DEGs, y = comparison, fill = comparison, label = DEGs)) +
  facet_grid(cols = vars(kind), switch = "y") +
  geom_col() +
  #geom_shadowtext(position = position_stack(vjust = 0.5), color = "white") +
  geom_text(position = position_stack(vjust = 0.5)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = comparison_colors) +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),legend.position= "none",
        axis.ticks.y = element_blank(), axis.title.y = element_blank(), axis.line.y = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA)) +
  
  plot_annotation(title = "Differentially Expressed Genes (DEGs)",
                  subtitle = "p.adj < 0.01 & |logFC| > 1") +
  plot_layout(widths = c(1.2, 4))

ggsave2("../figures/TEcounts/colplot-DEGs-Figure-4B.pdf", width = 190, height = 100, units = "mm")
```

```{r}
ggplot(DEG_numbers %>% filter(kind == "down", element == "TE"), 
       aes(x = DEGs, y = comparison, fill = comparison, label = DEGs)) +
  facet_grid(cols = vars(kind), switch = "y")+ #, scale = "free", space = "free") +
  geom_col() +
  #geom_shadowtext(position = position_stack(vjust = 0.5), color = "white") +
  geom_text(position = position_stack(vjust = 0.5)) +
  scale_y_discrete(position = "right") +
  scale_x_reverse(expand = expansion(mult = c(0.05, 0))) +
  scale_fill_manual(values = comparison_colors) +
   labs(x = "DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),legend.position= "none",
        axis.ticks.y = element_blank(), axis.title.y = element_blank(), axis.line.y = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA)) +
  
  ggplot(DEG_numbers %>% filter(kind == "up", element == "TE"), 
       aes(x = DEGs, y = comparison, fill = comparison, label = DEGs)) +
  facet_grid(cols = vars(kind), switch = "y") +
  geom_col() +
  #geom_shadowtext(position = position_stack(vjust = 0.5), color = "white") +
  geom_text(position = position_stack(vjust = 0.5)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = comparison_colors) +
   labs(x = "DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"),legend.position= "none",
        axis.ticks.y = element_blank(), axis.title.y = element_blank(), axis.line.y = element_blank(),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA)) +
  
  plot_annotation(title = "Differentially Expressed TEs (DETEs)",
                  subtitle = "p.adj < 0.01 & |logFC| > 1") +
  plot_layout(widths = c(6,4))

ggsave2("../figures/TEcounts/colplot-DETEs-Figure-4B.pdf", width = 190, height = 100, units = "mm")
```

### Volcano plots {.tabset}

Using enhanced volcano

#### Zoom out

```{r fig.width=10,  fig.height=16}
(volcanos <- enhanced_volcano(res_genes$TET3_vs_WT,  sample = "TET3", reference = "WT: Genes", xlim = c(-9, 9), ylim = c(0, 110), label.x = 5, label.y = 60) +
             enhanced_volcano(res_TEs$TET3_vs_WT,    sample = "TET3", reference = "WT", xlim = c(-9, 9), ylim = c(0, 110), label.x = 5, label.y = 60) +
             enhanced_volcano(res_genes$NTS_vs_WT, sample = "NTS",  reference = "WT", xlim = c(-9, 9), ylim = c(0, 110), label.x = 5, label.y = 60) +
             enhanced_volcano(res_TEs$NTS_vs_WT, sample = "NTS",  reference = "WT", xlim = c(-9, 9), ylim = c(0, 110), label.x = 5, label.y = 60) +
             enhanced_volcano(res_genes$NTS_vs_TET3, sample = "NTS", reference = "NTS", xlim = c(-9, 9), ylim = c(0, 110), label.x = 5, label.y = 60, caption = TRUE) +
             enhanced_volcano(res_TEs$NTS_vs_TET3, sample = "NTS", reference = "NTS", xlim = c(-9, 9), ylim = c(0, 110), label.x = 5, label.y = 60, caption = TRUE) +
  plot_layout(nrow = 3, guides = 'collect') &  theme(legend.position = 'top'))

ggsave(filename = "../figures/TEcounts/volcano_plots_all.pdf", plot = volcanos, width = 10, height = 16)
```

#### Zoom in

```{r fig.width=10,  fig.height=16}
(volcanos <- enhanced_volcano(res_genes$TET3_vs_WT,  sample = "TET3", reference = "WT") +
             enhanced_volcano(res_TEs$TET3_vs_WT,    sample = "TET3", reference = "WT") +
             enhanced_volcano(res_genes$NTS_vs_WT, sample = "NTS",  reference = "WT") +
             enhanced_volcano(res_TEs$NTS_vs_WT, sample = "NTS",  reference = "WT") +
             enhanced_volcano(res_genes$NTS_vs_TET3, sample = "NTS", reference = "NTS", caption = TRUE) +
             enhanced_volcano(res_TEs$NTS_vs_TET3, sample = "NTS", reference = "NTS", caption = TRUE) +
  plot_layout(nrow = 3, guides = 'collect') &  theme(legend.position = 'top'))

ggsave(filename = "../figures/TEcounts/volcano_plots_zoom.pdf", plot = volcanos, width = 10, height = 16)
```

#### For figures

```{r fig.width=8,  fig.height=4}
volcano_figure <- function(res, sample, reference) {
  enhanced_volcano(res, sample = sample, reference = reference,  xlim = c(-10, 10), ylim = c(0, 20), label.x = 5, label.y = 20, 
                   title = str_c(sample, " vs ", reference), caption = FALSE, size.label = 6,
                   lab = res$id %>%  str_remove_all(".*[[:digit:]\\)]:") %>% str_remove_all(":LTR"), 
                 #col = c("grey30", "grey30", "grey30", "red2"),
                 pointSize = 2,
                 labSize = 4,
                 labCol = "black",
                 labFace = "plain",
                 boxedLabels = FALSE,
                 parseLabels = FALSE,
                 drawConnectors = TRUE,
                 #drawConnectors = FALSE,
                 widthConnectors = 0.5,
                 typeConnectors = "open",
                 endsConnectors = "first",
                 lengthConnectors = unit(0.01, "npc"),
                 colConnectors = "grey10",
                 max.overlaps = 17,
                 maxoverlapsConnectors = NULL,
                 min.segment.length = 0,
                 directionConnectors = "both",
                 arrowheads = TRUE,
                 ) +
  theme_cowplot() +
  theme(legend.position = "none")
}

volcano_figure(res_TEs$TET3_vs_WT, sample = "TET3", reference = "WT") +
  volcano_figure(res_TEs$NTS_vs_WT, sample = "NTS", reference = "WT")
  
ggsave2("../figures/TEcounts/volcano_plots_figure-4A-.pdf", width = 200, height = 100, unit = "mm")
```


```{r fig.width=8,  fig.height=4}
volcano_figure(res_TEs$NTS_vs_TET3, sample = "NTS", reference = "TET3")
  
ggsave2("../figures/TEcounts/volcano_plots_figure-supp.pdf", width = 100, height = 100, unit = "mm")
```

#### Coloring by TE type

```{r fig.width=12,  fig.height=4}
volcano_grouping <- function(res, sample, reference) {
  
  keyvals <- case_when(str_detect(res$id, "DNA") ~ 17,
                       str_detect(res$id, "LTR") ~ 18,
                       .default = 16)
  keyvals[is.na(keyvals)] <- 16
  names(keyvals)[keyvals == 17] <- 'dna'
  names(keyvals)[keyvals == 18] <- 'ltr'
  names(keyvals)[keyvals == 16] <- 'other'
  
  keyvals <- case_when(str_detect(res$id, "DNA") ~ "red",
                       str_detect(res$id, "LTR") ~ "royalblue",
                       .default = "green3")
  keyvals[is.na(keyvals)] <- 16
  names(keyvals)[keyvals == "red"] <- 'DNA'
  names(keyvals)[keyvals == "royalblue"] <- 'LTR'
  names(keyvals)[keyvals == "green3"] <- 'other'
  
  enhanced_volcano(res, sample = sample, reference = reference,  xlim = c(-10, 10), ylim = c(0, 20), label.x = 5, label.y = 20, 
                   title = str_c(sample, " vs ", reference), caption = FALSE, size.label = 6,
                   lab = res$id %>%  str_remove_all(".*[[:digit:]\\)]:") %>% str_remove_all(":LTR"), 
                 #col = c("grey30", "grey30", "grey30", "red2"),
                 pointSize = 2,
                 #shapeCustom = keyvals,
                 colCustom = keyvals,
                 col = c("grey30", "forestgreen", "royalblue", "red2"),labSize = 4,
                 labCol = "black",
                 labFace = "plain",
                 boxedLabels = FALSE,
                 parseLabels = FALSE,
                 drawConnectors = TRUE,
                 #drawConnectors = FALSE,
                 widthConnectors = 0.5,
                 typeConnectors = "open",
                 endsConnectors = "first",
                 lengthConnectors = unit(0.01, "npc"),
                 colConnectors = "grey10",
                 max.overlaps = 15,
                 maxoverlapsConnectors = NULL,
                 min.segment.length = 0.4,
                 directionConnectors = "both",
                 arrowheads = TRUE,
                 ) +
  theme_cowplot() +
  theme(legend.position = "right")
}



volcano_grouping(res_TEs$TET3_vs_WT, sample = "TET3", reference = "WT") +
  volcano_grouping(res_TEs$NTS_vs_WT, sample = "NTS", reference = "WT") +
  plot_layout(guides = "collect")
  
ggsave2("../figures/TEcounts/volcano_plots_figure-4A-colors.pdf", width = 250, height = 100, unit = "mm")
```

```{r fig.width=8,  fig.height=4}
volcano_grouping(res_TEs$NTS_vs_TET3, sample = "NTS", reference = "TET3")
  
ggsave2("../figures/TEcounts/volcano_plots_figure-supp-colors.pdf", width = 160, height = 100, unit = "mm")
```


### Comparison of DEGs with different thresholds

```{r fig.height=7, fig.width=11}
visualise_degs <- function(DEGs_vsd_df, padj_lim, lfc_lim, fill = bin_padj) {
  DEGs_vsd_df %>%
    filter(padj <= padj_lim, abs(log2FoldChange) > lfc_lim, element == "TE") %>%
    mutate(rank = ntile(padj, 10), 
           logpadj = -log10(padj),
           bin_padj = cut(padj, c(1, 0.5, 0.1, 0.05, 0.01, 0.001, 0.0001, 0)),
           bin_padj = fct_reorder(bin_padj, -padj),
           bin_l2fc = cut(log2FoldChange, c(-Inf, -10, -4, -2, -1, -0.5, 0, 0.5, 1, 2, 4, 10, Inf)),
           bin_l2fc = fct_reorder(bin_l2fc, log2FoldChange),
           bin_absl2fc = cut(abs(log2FoldChange), c(0, 0.5, 1, 2, 4, 10, Inf)),
           bin_absl2fc = fct_reorder(bin_absl2fc, abs(log2FoldChange))) %>%
    group_by(comparison, {{ fill }}) %>%
    summarise(count = n(), .groups = 'drop') %>%
    ggplot(aes(x = count, y = comparison, fill = {{ fill }})) +
    geom_col(color = NA) +
    ggtitle(str_c("DETEs with a p-adj <= ", padj_lim, " & abs(log2FC) > ", lfc_lim)) +
    scale_fill_viridis_d(direction = -1, option = if_else(deparse(substitute(fill)) == "bin_padj" , "plasma", "viridis")) +
    #scale_fill_distiller(palette = 2)
    scale_y_discrete(limits = rev)
}

deg_pv <- visualise_degs(DEGs_vsd_df, padj_lim = 0.05, lfc_lim = 1, fill = bin_padj)
deg_fc <- visualise_degs(DEGs_vsd_df, padj_lim = 0.05, lfc_lim = 1, fill = bin_absl2fc)

all_pv <- visualise_degs(DEGs_vsd_df, padj_lim = 1, lfc_lim = 0, fill = bin_padj)
all_fc <- visualise_degs(DEGs_vsd_df, padj_lim = 1, lfc_lim = 0, fill = bin_absl2fc)

(DEGs_padj_l2fc <- (all_pv + all_fc + deg_pv + deg_fc))
ggsave2("../figures/TEcounts/DETEs_padj_l2fc.pdf", width = 11, height = 7)
```

### Heatmaps

```{r}
deg_mat = function(res, cts, ..., scaled = FALSE) {
  deg_cts(res, cts, ...) %>% # ...) %>%
    column_to_rownames("id") %>%
    select(-c(baseMean:padj)) %>%
    rename_with(~ .x %>% toupper() %>% str_replace_all("_", " ")) %>%
    as.matrix() %>%
    {if(scaled) {t(.) %>% scale() %>% t()}
      else .}
}

hm_wrapper = function(mat, scaled = F, cluster_rows = T, cluster_columns = T, title = "") {
  if(scaled) {col = rev(brewer.pal(9, "RdBu"));  name = "scaled\nexpression"}
  else       {col =     brewer.pal(9, "YlOrRd"); name = "expression"}
  Heatmap(mat, cluster_rows = cluster_rows, cluster_columns = cluster_columns, show_row_names = F, col = col, name = name, column_title = title)
}

deg_hm = function(res, cts, ..., scaled = F, file = F, title = NULL) {
  mat = deg_mat(res = res, cts = cts, ... = ..., scaled = scaled)
  if (file != 0) {
    pdf(file, height = 5, width = 5)
      hm <- hm_wrapper(mat, scaled = scaled)
      draw(hm, column_title = paste(nrow(mat), title, "DEGs", deparse(substitute(...))))
    dev.off()
  } else {
    hm_wrapper(mat, scaled = scaled, title = title)
  }
}

cts_log <- cts %>% mutate(across(-id, ~ log10(.x + 1)))
```

#### Scaled first

```{r}
hms_TET3 <- deg_hm(res_TEs$TET3_vs_WT,   vsd,                padj < 0.01, abs(log2FoldChange) > 1, title = "TET3 DEGs", scaled = TRUE) +
  deg_hm(res_TEs$TET3_vs_WT, vsd %>% select(id, WT_1:TET3_2), padj < 0.01, abs(log2FoldChange) > 1, title = "TET3 only", scaled = TRUE) +
  deg_hm(res_TEs$TET3_vs_WT, cts_log,                         padj < 0.01, abs(log2FoldChange) > 1, title = "log10(counts + 1)", scaled = FALSE) 

hms_NTS <- deg_hm(res_TEs$NTS_vs_WT,   vsd,                             padj < 0.01, abs(log2FoldChange) > 1, title = "NTS DEGs", scaled = TRUE) +
  deg_hm(res_TEs$NTS_vs_WT, vsd %>% select(id, WT_1:WT_2, NTS_1:NTS_2), padj < 0.01, abs(log2FoldChange) > 1, title = "NTS only", scaled = TRUE) +
  deg_hm(res_TEs$NTS_vs_WT, cts_log,                                    padj < 0.01, abs(log2FoldChange) > 1, title = "log10(counts + 1)", scaled = FALSE)

hms_sibs <- deg_hm(res_TEs$NTS_vs_TET3,   vsd,                             padj < 0.01, abs(log2FoldChange) > 1, title = "sibs DEGs", scaled = TRUE) +
  deg_hm(res_TEs$NTS_vs_TET3, vsd %>% select(id, WT_1:WT_2, NTS_1:NTS_2), padj < 0.01, abs(log2FoldChange) > 1, title = "sibs only", scaled = TRUE) +
  deg_hm(res_TEs$NTS_vs_TET3, cts_log,                                    padj < 0.01, abs(log2FoldChange) > 1, title = "log10(counts + 1)", scaled = FALSE)
```

```{r}
pdf(file = "../figures/TEcounts/heatmap_DETEs_all.pdf", height = 12, width = 7) 
grid.newpage()
pushViewport(viewport(x = 0, y = 1, height = 0.3333, just = c("left", "top")))
draw(hms_TET3, newpage = FALSE, column_title = "33 TET3 DETEs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()

pushViewport(viewport(x = 0, y = 0.6666, height = 0.3333, just = c("left", "top")))
draw(hms_NTS, newpage = FALSE, column_title = "40 NTS DETEs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()

pushViewport(viewport(x = 0, y = 0.3333, height = 0.3333, just = c("left", "top")))
draw(hms_sibs, newpage = FALSE, column_title = "46 siblings DETEs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()
dev.off()
```

#### Raw first

```{r}
hms_TET3 <- deg_hm(res_TEs$TET3_vs_WT, cts_log,                          padj < 0.01, abs(log2FoldChange) > 1, title = "log10(counts + 1)", scaled = FALSE)  +
  deg_hm(res_TEs$TET3_vs_WT,   vsd,                padj < 0.01, abs(log2FoldChange) > 1, title = "TET3 DEGs", scaled = TRUE) +
  deg_hm(res_TEs$TET3_vs_WT, vsd %>% select(id, WT_1:TET3_2), padj < 0.01, abs(log2FoldChange) > 1, title = "TET3 only", scaled = TRUE)

hms_NTS <- deg_hm(res_TEs$NTS_vs_WT, cts_log,                                       padj < 0.01, abs(log2FoldChange) > 1, title = "log10(counts + 1)", scaled = FALSE) +
  deg_hm(res_TEs$NTS_vs_WT,   vsd,                             padj < 0.01, abs(log2FoldChange) > 1, title = "NTS DEGs", scaled = TRUE) +
  deg_hm(res_TEs$NTS_vs_WT, vsd %>% select(id, WT_1:WT_2, NTS_1:NTS_2), padj < 0.01, abs(log2FoldChange) > 1, title = "NTS only", scaled = TRUE)

hms_sibs <- deg_hm(res_TEs$NTS_vs_TET3, cts_log,                                    padj < 0.01, abs(log2FoldChange) > 1, title = "log10(counts + 1)", scaled = FALSE) +
  deg_hm(res_TEs$NTS_vs_TET3,   vsd,                             padj < 0.01, abs(log2FoldChange) > 1, title = "sibs DEGs", scaled = TRUE) +
  deg_hm(res_TEs$NTS_vs_TET3, vsd %>% select(id, WT_1:WT_2, NTS_1:NTS_2), padj < 0.01, abs(log2FoldChange) > 1, title = "sibs only", scaled = TRUE)
```

```{r}
pdf(file = "../figures/TEcounts/heatmap_DETEs_all-unscaled.pdf", height = 12, width = 7) 
grid.newpage()
pushViewport(viewport(x = 0, y = 1, height = 0.3333, just = c("left", "top")))
draw(hms_TET3, newpage = FALSE, column_title = "33 TET3 DETEs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()

pushViewport(viewport(x = 0, y = 0.6666, height = 0.3333, just = c("left", "top")))
draw(hms_NTS, newpage = FALSE, column_title = "40 NTS DETEs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()

pushViewport(viewport(x = 0, y = 0.3333, height = 0.3333, just = c("left", "top")))
draw(hms_sibs, newpage = FALSE, column_title = "46 siblings DETEs padj < 0.01 & |log2FC| > 1", 
     column_title_gp = gpar(fontsize = 16, fontface = "bold"))
popViewport()
dev.off()
```

## Tidy Heatmaps 

```{r}
# color ramp from a Brewer palette
brewer.ramp <- function(start, end, pal) {
  palette = RColorBrewer::brewer.pal(n = RColorBrewer::brewer.pal.info[pal, "maxcolors"],  name = pal)
  sequence = seq(start, end, (end - start)/(RColorBrewer::brewer.pal.info[pal, "maxcolors"] - 1))
  ramp = circlize::colorRamp2(sequence, palette)
  return(ramp)
}

# Function that takes a tibble of combined DMR comparisons and returns a long table
filter_degs_and_pivot <- function(tb_degs, comp) {
  tb_degs %>% 
    filter(comparison == comp)  %>% 
    pivot_longer(cols = -c(comparison:padj), values_to = "counts", names_to = "sample") %>% 
    mutate(condition = str_remove(sample, "_.$") %>% fct_inorder()) %>% 
    # scale expression
    group_by(id) %>% 
    mutate(scaled = scale(counts) %>% as.numeric()) %>% 
    ungroup()
}

#Function that takes a ComplexHeatmap list and makes it work with Patchwork syntax
wrap_heatmap_list <- function(heatmap_list, draw = FALSE) {
  heatmap_list %>% 
    {if (draw) ComplexHeatmap::draw(.) else .} %>%
    grid::grid.grabExpr() %>% 
    patchwork::wrap_elements()
}

# function to plot heatmaps
heatmap_degs <- function(tb_degs_long, show_legends = TRUE, title = character(0), max_n = Inf, 
                         top_tile_variable = condition, top_tile_colors = condition_colors_2,
                         use_raster = TRUE, show_row_names = FALSE) {
  
  n_degs = unique(tb_degs_long$id) %>% length()
  
  tb_degs_long_sample <- tb_degs_long %>% 
    group_by(kind) %>% 
    mutate(genotype = condition) 
  #mutate(genotype = {{ top_tile_variable }})
  
  heatmap(tb_degs_long_sample, .row = id, .column = sample, .value = scaled, 
          cluster_columns = TRUE, show_row_names = FALSE, use_raster = use_raster, 
          palette_value = color_ramp_scaled, name = "scaled", 
          palette_grouping = list(c("firebrick3", "dodgerblue3")), use_raster = TRUE,
          row_title = NULL, column_title = NULL, width = unit(4, "cm"),
          show_row_names = show_row_names) %>% 
    annotation_tile(genotype, palette = top_tile_colors, show_annotation_name = FALSE, size = unit(0.3, "cm")) %>% 
    annotation_tile(log2FC, palette = color_ramp_l2fc, annotation_name_side = "top", size = unit(0.7, "cm")) %>% 
    as_ComplexHeatmap() %>% 
    draw(row_title = str_c(title, " (n=" , n_degs, ")"), merge_legend = TRUE, 
         show_heatmap_legend = show_legends, show_annotation_legend = show_legends) %>% 
    wrap_heatmap_list(draw = FALSE)
}
```

```{r}
tb_degs <- DEGs_vsd_p01_lfc1 %>% 
  map_df(.id = "comparison", ~ .x) %>%
  mutate(comparison = as_factor(comparison)) %>% 
  rename(log2FC = log2FoldChange) %>% 
  mutate(kind = case_when(log2FC > 0 ~ "up",
                          log2FC < 0 ~ "down")) %>% 
  relocate(kind, .after = log2FC)

color_ramp_l2fc  <- brewer.ramp(5, -5, "Spectral")
color_ramp_scaled   <- brewer.ramp(2, -2, "RdBu")

condition_colors_2 <- c("WT"  = "grey40", "TET3"  = "tomato2", "NT" = "slateblue2") %>% 
  fct_inorder()
```

```{r fig.width=10, fig.height=12}
tb_degs %>% 
  filter(element == "gene") %>% 
  filter_degs_and_pivot(comp = "TET3_vs_WT") %>% 
  heatmap_degs(title = "TET3 vs WT: Genes", show_legends = FALSE) +
  
  tb_degs %>% 
  filter(element == "gene") %>% 
  filter_degs_and_pivot(comp = "NTS_vs_WT") %>% 
  heatmap_degs(title = "NTS vs WT: Genes", show_legends = FALSE) +
  
  tb_degs %>% 
  filter(element == "gene") %>% 
  filter_degs_and_pivot(comp = "NTS_vs_TET3") %>% 
  heatmap_degs(title = "NTS vs TET3: Genes") +
  
  tb_degs %>% 
  filter(element == "TE") %>% 
  filter_degs_and_pivot(comp = "TET3_vs_WT") %>% 
  heatmap_degs(title = "TET3 vs WT: TEs", show_legends = FALSE) +
  
  tb_degs %>% 
  filter(element == "TE") %>% 
  filter_degs_and_pivot(comp = "NTS_vs_WT") %>% 
  heatmap_degs(title = "NTS vs WT: TEs", show_legends = FALSE) +
  
  tb_degs %>% 
  filter(element == "TE") %>% 
  filter_degs_and_pivot(comp = "NTS_vs_TET3") %>% 
  heatmap_degs(title = "NTS vs TET3: TEs") +
  
  plot_layout(nrow = 2, widths = c(8.5, 8.5, 11))
  
ggsave2("../figures/TEcounts/heatmaps-DEGs-Figure-4C.pdf", width = 250, height = 300, units = "mm")
```

TEs with names

```{r fig.width=14, fig.height=9}
tb_degs %>% 
  filter(element == "TE") %>% 
  mutate(id = str_remove(id, "rnd-._family-") %>% str_replace_all(":", "  ")) %>% 
  filter_degs_and_pivot(comp = "TET3_vs_WT") %>% 
  heatmap_degs(title = "TET3 vs WT: TEs", show_legends = FALSE, show_row_names = TRUE) +
  
  tb_degs %>% 
  filter(element == "TE") %>% 
  mutate(id = str_remove(id, "rnd-._family-") %>% str_replace_all(":", "  ")) %>% 
  filter_degs_and_pivot(comp = "NTS_vs_WT") %>% 
  heatmap_degs(title = "NTS vs WT: TEs", show_legends = FALSE, show_row_names = TRUE) +
  
  tb_degs %>% 
  filter(element == "TE") %>% 
  mutate(id = str_remove(id, "rnd-._family-") %>% str_replace_all(":", "  ")) %>% 
  filter_degs_and_pivot(comp = "NTS_vs_TET3") %>% 
  heatmap_degs(title = "NTS vs TET3: TEs", show_row_names = TRUE) +
  
  plot_layout(nrow = 1, widths = c(8.5, 8.5, 11))
  
ggsave2("../figures/TEcounts/heatmaps-DETEs-with-names.pdf", width = 14, height = 9)
```

## Boxplots of expression by direction

### vsd 

```{r}
degs_tes_vsd_long <- tb_degs %>% 
  filter(element == "TE") %>% 
  pivot_longer(WT_1:NTS_2, names_to = "sample", values_to = "expression") %>% 
  mutate(sample = str_replace(sample, "_", " ") %>%  fct_inorder(),
         condition = str_remove(sample, " ."))
```

```{r, fig.width=4, fig.height=3}
degs_tes_vsd_long %>% 
  filter(comparison == "TET3_vs_WT") %>% 
  ggplot(aes(x = sample, y = expression, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  labs(subtitle = "TET3 vs WT") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank())
```

```{r, fig.width=4, fig.height=3}
degs_tes_vsd_long %>% 
  filter(comparison == "NTS_vs_WT") %>% 
  ggplot(aes(x = sample, y = expression, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  labs(subtitle = "NTS vs WT") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank())
```

```{r, fig.width=4, fig.height=3}
degs_tes_vsd_long %>% 
  filter(comparison == "NTS_vs_TET3") %>% 
  ggplot(aes(x = sample, y = expression, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  labs(subtitle = "NTS vs TET3") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank())
```

### cts (log scale)

```{r}
degs_tes_cts_long <- map(DEGs_cts, ~ filter(., padj < 0.01, abs(log2FoldChange) > 1) %>% 
      mutate(element = case_when(str_starts(id, "gene") ~ "gene",
                                 !str_starts(id, "gene") ~ "TE")) %>% 
      relocate(element, .after = id)) %>% 
  map_df(.id = "comparison", ~ .x) %>%
  mutate(comparison = as_factor(comparison)) %>% 
  rename(log2FC = log2FoldChange) %>% 
  mutate(kind = case_when(log2FC > 0 ~ "up",
                          log2FC < 0 ~ "down")) %>% 
  relocate(kind, .after = log2FC) %>% 
  filter(element == "TE") %>% 
  pivot_longer(WT_1:NTS_2, names_to = "sample", values_to = "cts") %>% 
  mutate(sample = str_replace(sample, "_", " ") %>%  fct_inorder(),
         condition = str_remove(sample, " ."),
         log_cts = log10(1 + cts))
```

```{r, fig.width=4, fig.height=3}
degs_tes_cts_long %>% 
  filter(comparison == "TET3_vs_WT") %>% 
  ggplot(aes(x = sample, y = cts, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous("counts", transform = "log1p", breaks = c(0,1,3,10,30,100,300,1000,3000)) +
  labs(subtitle = "TET3 vs WT DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank())

ggsave2("../figures/TEcounts/boxplots-counts-DEGs-TET3_Figure4B.pdf",width = 100, height = 100, unit = "mm")
```

```{r, fig.width=8, fig.height=3}
degs_tes_cts_long %>% 
  filter(comparison == "NTS_vs_WT") %>% 
  ggplot(aes(x = sample, y = cts, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous("counts", transform = "log1p", breaks = c(0,1,3,10,30,100,300,1000,3000)) +
  labs(subtitle = "NTS vs WT DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank()) +
degs_tes_cts_long %>% 
  filter(comparison == "NTS_vs_TET3") %>% 
  ggplot(aes(x = sample, y = cts, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous("counts", transform = "log1p", breaks = c(0,1,3,10,30,100,300,1000,3000)) +
  labs(subtitle = "NTS vs TET3 DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank())

ggsave2("../figures/TEcounts/boxplots-counts-DEGs-NTS_Fig-Supp.pdf",width = 200, height = 100, unit = "mm")
```

### cts (log scale) with stats

Run one-way ANOVA separately for each comparison in up & down

```{r}
degs_tes_cts_long %>%
  group_by(comparison, kind) %>%                       
  anova_test(log_cts ~ sample) %>%
  get_anova_table()
```

Function to get letters (CLD) from a Tukey test

```{r}
# get_letters.R ------------------------------------------------------------
# Pipe‑friendly helper to create compact‑letter displays (CLDs)
# from a tidy Tukey HSD result (e.g. rstatix::tukey_hsd() output).
#
# Author : ChatGPT (OpenAI)
# Date   : 2025‑07‑21
# -----------------------------------------------------------------------------
#' get_letters
#'
#' Quickly derive compact‑letter displays (CLDs) from a tidy Tukey‑HSD table.
#'
#' This is a %>%‑friendly wrapper around **multcompView::multcompLetters()**
#' that plays nicely with the tidy output produced by
#' `rstatix::tukey_hsd()` or any tibble that contains at least three
#' columns: the first group, the second group, and the adjusted p‑value.
#'
#' If your Tukey table contains extra grouping/facetting columns
#' (e.g. `kind == "up"/"down"`), pass their unquoted names via `by`.
#'
#' @param data    A data.frame/tibble with pairwise comparisons.
#' @param p_col   Name of the column that stores adjusted p‑values.
#' @param group1  Name of the first group column.
#' @param group2  Name of the second group column.
#' @param by      Optional tidy‑select of columns to treat as grouping
#'                variables (facets).  Default = NULL.
#' @param threshold Significance threshold for grouping letters (default 0.05).
#' @param sample_levels Optional character vector giving the order of levels
#'                in the output; falls back to sorted union of group names.
#'
#' @return A tibble with columns `sample` and `letter`, plus any columns
#'         listed in `by`, suitable to join back to your original data
#'         or feed straight into `geom_text()`.
#'
#' @examples
#' # Example with the built‑in PlantGrowth data ---------------------------
#' if (requireNamespace("rstatix", quietly = TRUE)) {
#'   library(dplyr)
#'   library(rstatix)
#'
#'   tk_tbl <- PlantGrowth %>%
#'     rstatix::aov_test(weight ~ group) %>%            # ANOVA
#'     rstatix::tukey_hsd()                             # pairwise Tukey
#'
#'   get_letters(tk_tbl)
#' }
#'
#' @export
get_letters <- function(data,
                        p_col = "p.adj",
                        group1 = "group1",
                        group2 = "group2",
                        by = NULL,
                        threshold = 0.05,
                        sample_levels = NULL) {
  # required pkgs ---------------------------------------------------------
  if (!requireNamespace("multcompView", quietly = TRUE)) {
    stop("Package 'multcompView' is required but not installed.")
  }
  if (!requireNamespace("dplyr", quietly = TRUE)) {
    stop("Package 'dplyr' is required but not installed.")
  }
  if (!requireNamespace("tidyr", quietly = TRUE)) {
    stop("Package 'tidyr' is required but not installed.")
  }
  
  dplyr  <- asNamespace("dplyr")  # avoid R CMD check notes
  tidyr  <- asNamespace("tidyr")
  mcView <- asNamespace("multcompView")
  
  # convert `by` to quosures ---------------------------------------------
  by_quos <- rlang::enquos(by, .ignore_empty = "all")
  if (length(by_quos) == 0) by_quos <- NULL
  
  # helper to process a single split (facet) ------------------------------
  cld_one <- function(df) {
    g1   <- df[[group1]]
    g2   <- df[[group2]]
    pval <- df[[p_col]]
    
    all_lvls <- sort(unique(c(as.character(g1), as.character(g2))))
    if (!is.null(sample_levels)) {
      # respect user‑supplied order but keep only observed levels
      all_lvls <- sample_levels[sample_levels %in% all_lvls]
    }
    # build symmetric matrix ---------------------------------------------
    mat <- matrix(1, nrow = length(all_lvls), ncol = length(all_lvls),
                  dimnames = list(all_lvls, all_lvls))
    # fill lower & upper triangles
    for (i in seq_len(nrow(df))) {
      mat[ as.character(g1[i]), as.character(g2[i]) ] <- pval[i]
      mat[ as.character(g2[i]), as.character(g1[i]) ] <- pval[i]
    }
    # multcompView letters ------------------------------------------------
    letters_vec <- mcView$multcompLetters(mat, threshold = threshold)$Letters
    tibble::tibble(sample = names(letters_vec), letter = letters_vec)
  }
  
  # split‑apply depending on `by` ----------------------------------------
  if (is.null(by_quos)) {
    out <- cld_one(data)
  } else {
    out <- dplyr::group_split(dplyr::group_by(data, !!!by_quos)) %>%
      purrr::map_dfr(~ {
        key_vals <- .x %>% dplyr::distinct(!!!by_quos) %>% dplyr::slice(1)
        dplyr::bind_cols(key_vals, cld_one(.x))
      })
  }
  # ensure factor order ---------------------------------------------------
  if (!is.null(sample_levels)) {
    out$sample <- factor(out$sample, levels = sample_levels)
  }
  out
}

# -------------------------------------------------------------------------
# Test‑that examples -------------------------------------------------------
# These tests run only if **testthat** is installed and we're *not* in
# production knitting; they do nothing otherwise.
if (requireNamespace("testthat", quietly = TRUE)) {
  testthat::test_that("get_letters reproduces PlantGrowth CLDs", {
    testthat::skip_if_not_installed("rstatix")
    library(rstatix)
    tk_tbl <- PlantGrowth %>%
      aov(weight ~ group) %>%
      tukey_hsd()
    cld <- get_letters(tk_tbl)
    testthat::expect_true(all(c("sample", "letter") %in% names(cld)))
    # ensure at least two distinct letters (there are sig diffs)
    testthat::expect_gt(length(unique(cld$letter)), 1)
  })
}

```

TET3_vs_WT

```{r}
# get the maximum value per sample, for plotting letter above it
y_tbl <- degs_tes_cts_long %>%
  filter(comparison == "TET3_vs_WT") %>% 
  group_by(kind, sample) %>% 
  summarise(y = max(cts) * 1.3, .groups = "drop")

# perform Tukey test
tukey_tbl <- degs_tes_cts_long %>%
  filter(comparison == "TET3_vs_WT") %>% 
  group_by(kind) %>%         
  tukey_hsd(log_cts ~ sample)

# obtain significant letters
letters_tbl <- tukey_tbl %>%              
  get_letters(by = kind,                  # <- facet column(s)
              threshold = 0.05,           # default; change if you like
              sample_levels = levels(df$sample))  %>% 
  left_join(y_tbl, by = c("kind","sample"))
```

```{r fig.width=4, fig.height=3}
degs_tes_cts_long %>% 
  filter(comparison == "TET3_vs_WT") %>% 
  ggplot(aes(x = sample, y = cts, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous("counts", transform = "log1p", breaks = c(0,1,3,10,30,100,300,1000,3000)) +
  labs(subtitle = "TET3 vs WT DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank()) +
  
  geom_text(data = letters_tbl, aes(sample, y, label = letter), inherit.aes = FALSE, size = 4, vjust = 0)

ggsave2("../figures/TEcounts/boxplots-counts-DEGs-TET3_Figure4B-stats.pdf",width = 100, height = 100, unit = "mm")
```

The other two comparisons

```{r}
# get the maximum value per sample, for plotting letter above it
y_tbl <- degs_tes_cts_long %>%
  filter(comparison == "NTS_vs_WT") %>% 
  group_by(kind, sample) %>% 
  summarise(y = max(cts) * 1.3, .groups = "drop")

# perform Tukey test
tukey_tbl <- degs_tes_cts_long %>%
  filter(comparison == "NTS_vs_WT") %>% 
  group_by(kind) %>%         
  tukey_hsd(log_cts ~ sample)

# obtain significant letters
letters_tbl_NTS_WT <- tukey_tbl %>%              
  get_letters(by = kind,                  # <- facet column(s)
              threshold = 0.05,           # default; change if you like
              sample_levels = levels(df$sample))  %>% 
  left_join(y_tbl, by = c("kind","sample"))
```

```{r}
# get the maximum value per sample, for plotting letter above it
y_tbl <- degs_tes_cts_long %>%
  filter(comparison == "NTS_vs_TET3") %>% 
  group_by(kind, sample) %>% 
  summarise(y = max(cts) * 1.3, .groups = "drop")

# perform Tukey test
tukey_tbl <- degs_tes_cts_long %>%
  filter(comparison == "NTS_vs_TET3") %>% 
  group_by(kind) %>%         
  tukey_hsd(log_cts ~ sample)

# obtain significant letters
letters_tbl_NTS_TET3 <- tukey_tbl %>%              
  get_letters(by = kind,                  # <- facet column(s)
              threshold = 0.05,           # default; change if you like
              sample_levels = levels(df$sample))  %>% 
  left_join(y_tbl, by = c("kind","sample"))
```

```{r fig.width=8, fig.height=3}
degs_tes_cts_long %>% 
  filter(comparison == "NTS_vs_WT") %>% 
  ggplot(aes(x = sample, y = cts, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous("counts", transform = "log1p", breaks = c(0,1,3,10,30,100,300,1000,3000)) +
  labs(subtitle = "NTS vs WT DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank()) +
  
  geom_text(data = letters_tbl_NTS_WT, aes(sample, y, label = letter), inherit.aes = FALSE, size = 4, vjust = 0) +
  
  degs_tes_cts_long %>% 
  filter(comparison == "NTS_vs_TET3") %>% 
  ggplot(aes(x = sample, y = cts, fill = condition)) +
  facet_wrap(facets = vars(kind), scales = "free") +
  geom_boxplot(color = "black", outlier.colour = NA, alpha = 0.7) +
  geom_sina(shape = 16, size = 1, alpha = 0.5) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous("counts", transform = "log1p", breaks = c(0,1,3,10,30,100,300,1000,3000)) +
  labs(subtitle = "NTS vs TET3 DETEs") +
  theme(strip.background = element_rect(colour = NA, fill = "papayawhip"), legend.position= "none",
        axis.ticks.x = element_blank(), axis.title.x = element_blank(),  axis.text.x = element_text(angle = 90, hjust = 1),
        panel.border = element_rect(colour = "black", linewidth = 1, fill = NA), axis.line = element_blank()) +
  
  geom_text(data = letters_tbl_NTS_TET3, aes(sample, y, label = letter), inherit.aes = FALSE, size = 4, vjust = 0)

ggsave2("../figures/TEcounts/boxplots-counts-DEGs-NTS_Fig-Supp-stats.pdf",width = 200, height = 100, unit = "mm")
```
